---
title: "Explore Treehouse PolyA Compendium"
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

We are interested in adding more datasets to the Treehouse Splice Compendium prior to submitting the manuscript to a journal.
Currently, the priorities are to include samples in the Treehouse polyA compendia,
with healthy comparators from dGTEx.

This notebook reads in the clinical metadata from the v25 Treehouse PolyA Compendium and explores how many samples in each disease type are represented.
Additionally, it reads in sample metadata from the first release of the developmental GTEx (dGTEx) project to explore how many RNAseq samples of what normal tissue types are represented.

## Setup
Libraries
```{r}
library(ggplot2)
```

Files and directories
```{r files_and_directories}
## Directories ##
repo_root <- rprojroot::find_root(rprojroot::is_git_root)
metadata_dir <- file.path(repo_root, "metadata")

## Files ##
# Treehouse polyA compendium with disease type of samples
polya_clinical_file <- file.path(metadata_dir, "GSE294351_clinical_Treehouse-Tumor-Compendium-25.01-PolyA_20250131v1.tsv")
# dgtex metadata with library prep and tissue type info
dgtex_metadata_file <- file.path(metadata_dir, "dGTEx_Analysis_2026-01-30_v1_Sample_Attributes_Dataset_DS_open.xlsx")

## Read in files ##
polya_metadata_df <- readr::read_tsv(polya_clinical_file)
```
Summarize number of pedaya samples in polyA compendium by disease type
```{r}
polya_metadata_df |>
  dplyr::summarise(.by = source_name,
                   n = dplyr::n())
```
So we want to exclude samples from TARGET and TCGA 

```{r}
polya_summary_min_sample <- polya_metadata_df |>
  # filter for human pedaya samples not in TARGET and TCGA
  dplyr::filter(pedaya == "yes",
                organism == "Homo sapiens",
                ! source_name %in% c("TARGET", "TCGA")) |>
  dplyr::summarise(.by = c(disease),
                   n = dplyr::n()) |>
  # filter for diseases with at least 15 samples to make plot readable
  dplyr::filter( n >= 15)
```

Plot polyA compendium samples with n >= 15 

```{r}
#| label: fig-treehouse_polya_samples
#| fig-width: 9
#| fig-height: 7

# plot histogram
ggplot(polya_summary_min_sample, 
       # sort x axis by descending y axis value
       aes(x = reorder(disease, -n),
           y = n)) +
  geom_col() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
        ) +
  labs(
    x = "Treehouse polyA compendium disease"
  ) +
  # add N observations to bars
  geom_text(
    aes(label = n), 
    hjust = 0.5,
    vjust = -1
    ) 
```
Can Treehouse polyA compendium samples of rarer diseases be grouped into larger categories?
Filter for compendium samples with under 15 samples
```{r}
polya_summary_less_than_min_sample <- polya_metadata_df |>
  # filter for human pedaya samples not in TARGET and TCGA
  dplyr::filter(pedaya == "yes",
                organism == "Homo sapiens",
                ! source_name %in% c("TARGET", "TCGA")) |>
  dplyr::summarise(.by = c(disease),
                   n = dplyr::n()) |>
  # filter for diseases with under 15 samples
  dplyr::filter( n < 15)

# count how many uniquely labeled disease types here have less than 15 samples
nrow(polya_summary_less_than_min_sample)
```

There are altogether many samples in the compendium that are in a category with less than 15 samples.
Most of these are weird sarcomas, which will likely be hard to find a comparator for.

Below is a list of diseases that I think are easy to find comparators for/handle, from me eyeballing:

fibrolamellar hepatocellular carcinoma ( n = 4 ); undifferentiated hepatic sarcoma ( n = 2); kidney clear cell carcinoma (1) ; hepatocellular carcinoma (4); clear cell sarcoma of the kidney (1): kidney
colon adenocarcinoma ( n = 3 ) : colon
wilms tumor ( n = 13 ) : kidney
gastrointestinal stromal tumor ( n = 3) : stomach?
alveolar soft part sarcoma ( n = 3 ); lung adenocarcinoma (1) : lung?
melanoma (n = 9) : skin
retinoblastoma ( n = 2 ): eye