---
title: "Prep psi matrix for analysis"
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

This notebook reads in the PSI tables generated from the Treehouse Splice Compendium workflow, merges them into one matrix for ease of downstream analysis, and adds in a gene_name column to associate ENSG IDs with gene names.

## Setup

Load libraries

```{r load_libraries}
library(rtracklayer)
```

Define directories and file paths

```{r directories_and_files}
# find the root-level repo directory so results files can be accessed
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## directories ##
# parent directories
exploration_dir <- file.path(repo_root, "exploration")
results_dir <- file.path(repo_root, "results")
references_dir <- file.path(repo_root, "references")

# psi table dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined", "results", "splicing")

# output dir of psi table with gene name and splice event annotation status
v1_compendium_resuts_dir <- file.path(results_dir, "v1_compendium")

## files ##
# gtf file for converting ensg id to gene names
gtf_file <- file.path(references_dir, "gencode.v47.primary_assembly.annotation.gtf")

# psi matrix to make nice
psi_matrix_file <- file.path(combined_results_dir, "PSI_matrix_sample.txt")
# output nice psi matrix file
out_file <- file.path(v1_compendium_resuts_dir, "v1_compendium_psi_matrix.txt")

# define list of PSI results files produced by Shiba
psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of combined run shiba results
shiba_psi_paths <-file.path(combined_results_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

# Check output dir exists; if not, create it
if (!dir.exists(v1_compendium_resuts_dir)) {
  dir.create(v1_compendium_resuts_dir)
}
```

Read in files

```{r read_in_files}
## gene annotation file ##

# import gtf
gtf <- import(gtf_file)

# make a named vector of gene names to IDs
gene_names = setNames(gtf$gene_name, gtf$gene_id)

# remove the gtf when you are done with it
rm(gtf)

## splice result files ##
# read merged splice table from v1 workflow on the TARGET pilot samples
psi_results <- shiba_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric
      dplyr::mutate(across(contains("_PSI"), \(x) as.numeric(x)))
  })

# combine PSI values of samples run together into one dataframe to compare against other PSI tables
psi_table <- purrr::list_rbind(psi_results, names_to = "event_type")
```

## Add gene names to psi matrix made from combining all psi tables

Merge PSI matrix with gene info and annotation columns from merged psi table Each row of this matrix corresponds to a unique splice event. columns are event_type, pos_id, gene_id, gene_name, label (annotation status), and the sample-level PSI values (columns start with "SRR")

```{r add_gene_annotation_to_matrix}
nice_psi_df <-psi_table |> dplyr::mutate(
  # make a gene name column based off the gene name/gene id vector from gtf
 gene_name = gene_names[gene_id]
)

# print column names of non-PSI value columns
colnames(nice_psi_df[,sapply(nice_psi_df,is.character)])
```

## Write output

```{r write_output}
readr::write_tsv(nice_psi_df, out_file)
```

## Print session info

```{r session_info}
sessionInfo()
```
