---
title: Check Merge vs. Combined TARGET Pilot GTF and Junction files
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Read in directories and files

Define directories and file paths

```{r directories_and_files}
### Directories ###
## combined shiba run results on target pilot samples ##
# merged table eval dir
exploration_eval_dir <- file.path("exploration", "merged-method-target-pilot-eval")

# shiba results dir
# target pilot explroation dir
target_pilot_dir <- file.path("exploration", "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# junctions dir
combined_junctions_dir <- file.path(combined_results_dir, "junctions")
# gtf dir
combined_gtf_dir <- file.path(combined_results_dir, "annotation")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path("results", "merged_shiba", "target_pilot")

### Files ###
## combined shiba target pilot files
target_pilot_sample_sheet <- file.path(target_pilot_dir, "experiment.tsv")
# combined target pilot outputs
combined_gtf_file <- file.path(combined_gtf_dir, "assembled_annotation.gtf")
combined_junctions_file <- file.path(combined_junctions_dir, "junctions.bed")
# deduplicated combined junctions file - made by running bash scripts/deduplicated_shiba_junctions.sh
deduplicated_combined_junctions_file <- file.path(combined_junctions_dir, "deduplicated_junctions.bed")

## merged shiba target pilot files
merged_gtf_file <- file.path(merged_results_dir, "merged_gtf.gtf")
merged_junctions_file <- file.path(merged_results_dir, "merged_junctions.bed")

## output files
only_in_comb_junc <- file.path(exploration_eval_dir, "jcn_only_in_combined.bed")
only_in_merged_junc <- file.path(exploration_eval_dir, "jcn_only_in_merged.bed")

```

Create output dir if it does not exist

```{r create_dirs}
# Check if it exists; if not, create it
if (!dir.exists(exploration_eval_dir)) {
  dir.create(exploration_eval_dir, recursive = TRUE)
}
```

Read in files

```{r read_in_files}
# list of target pilot sample IDs
target_pilot_samples <- readr::read_tsv(target_pilot_sample_sheet) |>
  dplyr::pull(sample)

# minimum number of samples to filter minimum number of junctions by
min_samples <- 5

# read in files generated from combined shiba run
combined_junctions <- readr::read_tsv(
  combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in combined junctions table that have been deduplicated
deduplicated_combined_junctions <- readr::read_tsv(
  deduplicated_combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in files generated from merged shiba run
merged_junctions <- readr::read_tsv(
  merged_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)


```

## Check tables for parsing errors
### Print parsing errors in combined junctions table

```{r check_parsing_errors}
combined_junctions_dat <- vroom::vroom(combined_junctions_file)
combined_junctions_problems <- vroom::problems(combined_junctions_dat)
combined_junctions_problems
```
We have `r nrow(unique(combined_junctions_problems$row))` junctions in the combined junctions file that are impacted by the concatenation bug in Shiba.
PSI files for events involving these junctions are expected to be problematic in the combined PSI file.

### Check for parsing errors in the merged and deduplicated junctions tables

```{r}
merged_junctions_dat <- vroom::vroom(merged_junctions_file)
```

```{r}
deduplicated_comb_junctions_dat <- vroom::vroom(deduplicated_combined_junctions_file)
deduplicated_comb_junctions_problems <- vroom::problems(deduplicated_combined_junctions)
```

No problems are present in merged_junctions file or deduplicated junctions file

## Check if files are identical

```{r}
# ensure column order is the same before checking if data frames are identical
deduplicated_combined_junctions <- deduplicated_combined_junctions |>
  dplyr::relocate(dplyr::all_of(target_pilot_samples), .after = dplyr::last_col())

merged_junctions <- merged_junctions |>
  dplyr::relocate(dplyr::all_of(target_pilot_samples), .after = dplyr::last_col())

identical(deduplicated_combined_junctions, merged_junctions)
```

Deduplicated and merged junctions are still not identical, so need to drill down on what junctions are different

## Check differences in junctions bedfiles

Dimensions of each bedfile

```{r}
dim(deduplicated_combined_junctions)
```

```{r}
dim(merged_junctions)
```

There are `r nrow(merged_junctions) - nrow(deduplicated_combined_junctions)` more junctions in the merged junction file compared to the combined junction file.

## Subsample junctions to those in chromosome 1
To allow for running on my laptop, subsample junctions files for those in chromosome 1 prior to checking for differences

```{r subsample_junctions}
merged_junctions_chr1 <- merged_junctions |> dplyr::filter(chr == "chr1")
deduplicated_combined_junctions_chr1 <- deduplicated_combined_junctions |> dplyr::filter(chr == "chr1")
```

Check head of merged chromosome 1 junctions

```{r head_merged_junctions}
merged_junctions_chr1 |>
  dplyr::arrange(start) |>
  head()
```

Check head of deduplicated combined chromosome 1 junctions

```{r head_dedup_combined_junctions}
deduplicated_combined_junctions_chr1 |>
  dplyr::arrange(start) |>
  head()
```

Many junctions also have counts below 10 across all samples.
These junctions are not expected to make it into PSI calculation based off default Shiba filters (\> 10 junctions).

Check number of junctions in each subsetted dataframe
```{r check_num_junctions_chr1}
nrow(merged_junctions_chr1)
nrow(deduplicated_combined_junctions_chr1)
```

There are `r nrow(merged_junctions_chr1) - nrow(deduplicated_combined_junctions_chr1)` more junctions in the merged bedfile, compared to the combined bedfile.
The chromosome 1-subsetted junction bedfiles are not identical.

Remove junctions in chromosome 1 whose counts are \< 10 in all samples and check for differences

```{r filter_junction_counts}
merged_junctions_chr1_10 <- merged_junctions_chr1 |> dplyr::filter(
  dplyr::if_all(
    dplyr::all_of(target_pilot_samples), ~ . > 10
    )
  )

merged_junctions_chr1_10 |> 
  dplyr::arrange(start) |> 
  head()
```

```{r}
deduplicated_combined_junctions_chr1_10 <- deduplicated_combined_junctions_chr1 |> dplyr::filter(
  dplyr::if_all(
    dplyr::all_of(target_pilot_samples), ~ . > 10
    )
  )

deduplicated_combined_junctions_chr1_10 |> 
  dplyr::arrange(start) |> 
  head()
```

The two tables look identical so far - check with setdiff

```{r}
identical(deduplicated_combined_junctions_chr1, merged_junctions_chr1)

identical(deduplicated_combined_junctions_chr1_10, merged_junctions_chr1_10)
```

Check what's different in chromosome 1 junctions \> 10 counts

```{r}
combined_only_chr1 <- dplyr::setdiff(deduplicated_combined_junctions_chr1, merged_junctions_chr1)
nrow(combined_only_chr1)

# check fraction of combined_only_chr1 junctions that are only in chr1 of deduplicated combined junctions 
nrow(combined_only_chr1) / nrow(deduplicated_combined_junctions_chr1)

merged_only_chr1 <- dplyr::setdiff(merged_junctions_chr1, deduplicated_combined_junctions_chr1)
nrow(merged_only_chr1)

# check fraction of merged_only_chr1 junctions that are only in chr1 of merged combined junctions
nrow(merged_only_chr1) / nrow(merged_junctions_chr1)

combined_only_chr1_10 <- dplyr::setdiff(deduplicated_combined_junctions_chr1_10, merged_junctions_chr1_10)
nrow(combined_only_chr1_10)

# check fraction of combined_only_chr1 junctions > 10 counts that are only in chr1 of deduplicated combined junctions 
nrow(combined_only_chr1_10) / nrow(deduplicated_combined_junctions_chr1)

merged_only_chr1_10 <- dplyr::setdiff(merged_junctions_chr1_10, deduplicated_combined_junctions_chr1_10)
nrow(merged_only_chr1_10)
# check fraction of combined_only_chr1 junctions > 10 counts that are only in chr1 of deduplicated combined junctions 
nrow(merged_only_chr1_10) / nrow(merged_junctions_chr1)
```

Overall, only 1% of all chromosome 1 junctions are only found in either the combined or merged junctions files.
If we filter for junctions with more than 10 counts, we miss a small (0.1%) fraction of junctions in the combined file using the merged method, but we do not have any junctions not found in the combined method, which is great.

Potential sources of differences in the merged method that may result in junctions or junction values not present in the combined junctions file:

- After merging separate junction.bed files, I manually convert all NAs to 0.
  I don't know how Shiba handles junctions not found in all samples in the combined version

- I don't know how shiba "decides" what junctions get the duplication bug and how the bug impacts how junction coordinates are defined in each run

`r nrow(combined_only_chr1_10)`

Another note - because the concatenated junctions bug impacted one junction with reads > 10 in the combined junctions bedfile, I expect this will result in differenes in the PSI table

-m filte for things with 10 counts ina t least 5 samples to keep consistent with psi filtering
-explain waqrnings seen in eac file more and print parsing errors
show examples of junctions that dont match - any glaring things that explain why they are not in the merged method?
