---
title: Check Merge vs. Combined TARGET Pilot GTF and Junction files
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Read in directories and files

Define directories and file paths

```{r directories_and_files}
### Directories ###
## combined shiba run results on target pilot samples ##
# merged table eval dir
exploration_eval_dir <- file.path("exploration", "merged-method-target-pilot-eval")

# shiba results dir
combined_results_dir <- file.path("exploration", "merging-psi-tables", "target_pilot", "shiba_combined")
# junctions dir
combined_junctions_dir <- file.path(combined_results_dir, "junctions")
# gtf dir
combined_gtf_dir <- file.path(combined_results_dir, "annotation")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path("results", "merged_shiba", "target_pilot")

### Files ###
## combined shiba target pilot files
combined_gtf_file <- file.path(combined_gtf_dir, "assembled_annotation.gtf")
combined_junctions_file <- file.path(combined_junctions_dir, "junctions.bed")
# deduplicated combined junctions file - made by running bash scripts/deduplicated_shiba_junctions.sh
deduplicated_combined_junctions_file <- file.path(combined_junctions_dir, "deduplicated_junctions.bed")

## merged shiba target pilot files
merged_gtf_file <- file.path(merged_results_dir, "merged_gtf.gtf")
merged_junctions_file <- file.path(merged_results_dir, "merged_junctions.bed")

## output files
only_in_comb_junc <- file.path(exploration_eval_dir, "jcn_only_in_combined.bed")
only_in_merged_junc <- file.path(exploration_eval_dir, "jcn_only_in_merged.bed")

```

Create output dir if it does not exist
```{r create_dirs}
# Check if it exists; if not, create it
if (!dir.exists(exploration_eval_dir)) {
  dir.create(exploration_eval_dir, recursive = TRUE)
}
```


Read in files

```{r read_in_files}

# read in files generated from combined shiba run
combined_junctions <- readr::read_tsv(
  combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in combined junctions table that have been deduplicated
deduplicated_combined_junctions <- readr::read_tsv(
  deduplicated_combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in files generated from merged shiba run
merged_junctions <- readr::read_tsv(
  merged_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)


```

Check tables for parsing errors
```{r check_parsing_errors}
combined_junctions_dat <- vroom::vroom(combined_junctions_file)
combined_junctions_problems <- vroom::problems(combined_junctions_dat)
```

```{r}
merged_junctions_dat <- vroom::vroom(merged_junctions_file)
```

```{r}
deduplicated_comb_junctions_dat <- vroom::vroom(deduplicated_combined_junctions_file)
deduplicated_comb_junctions_problems <- vroom::problems(deduplicated_combined_junctions)
```

No problems are present in merged_junctions file

## Check if files are identical

### Check if combined and merged junctions.bed files are identical

```{r check_identical}
identical(combined_junctions, merged_junctions)
```

### Check if deduplicated combined and merged junctions are identical

```{r}
identical(deduplicated_combined_junctions, merged_junctions)
```

## Check differences in junctions bedfiles
Dimensions of each bedfile
```{r check_dimensions}
dim(combined_junctions)
```

```{r}
dim(deduplicated_combined_junctions)
```

```{r}
dim(merged_junctions)
```

Elements unique to each junction bedfile

```{r print_unique}
# elements in deduplicated combined junctions df but not in merged
junctions_unique_to_combined <- setdiff(deduplicated_combined_junctions, merged_junctions)

# sort so that we get IDs printed in the same order
junctions_unique_to_combined <- junctions_unique_to_combined[order(junctions_unique_to_combined$ID), ]

# print number of junctions that are unique to combined bedfile
nrow(junctions_unique_to_combined)

```

```{r}
# elements in combined junctions df but not in merged
junctions_unique_to_merged <- setdiff(merged_junctions, deduplicated_combined_junctions)

# sort so that we get IDs printed in the same order
junctions_unique_to_merged <-junctions_unique_to_merged[order(junctions_unique_to_merged$ID)]

# print number of junctions that are unique to merged bedfile
nrow(junctions_unique_to_merged)

```

```{r save-output}
# save junctions only in merged or combined tables as output
readr::write_tsv(junctions_unique_to_merged, file = jcn_only_in_merged)

readr::write_tsv(junctions_unique_to_combined, file = jcn_only_in_combined)

```
