---
title: "Debug merge junctions warning"
format: html
editor: visual
---

## Files and directories

```{r}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## define the data directories ##
exploration_dir <- file.path(repo_root, "exploration")
results_dir <- file.path(repo_root, "results", "merged_shiba", "test") # manually merged junction and gtf files and shiba results
merged_test_dir <- file.path(exploration_dir, "merged_shiba_test")
shiba_results_dir <- file.path(merged_test_dir, "shiba_results")

## data files ##
# combined run shiba results, junctions, and gtf ("control" results and files)
combined_junctions_file <- file.path(shiba_results_dir, "combined_run", "junctions", "junctions.bed") 

# junction bedfiles generated from separate shiba runs
SRR4376025_bedfile <- file.path(repo_root, "results/target/shiba/SRR4376025/junctions/junctions.bed")
SRR1559031_bedfile <- file.path(repo_root, "results/target/shiba/SRR1559031/junctions/junctions.bed")

# Read in junctions bed paths into a vector
junction_paths <- c(SRR4376025_bedfile, 
                   SRR1559031_bedfile)

```

## Print problems in each of the junction files

### Check separate junction bedfiles for problems

Check SRR4376025 bedfile for problems:

```{r}
SRR4376025_problems <- vroom::vroom(SRR4376025_bedfile)

vroom::problems(SRR4376025_problems) |> table()
```

No problems in the SRR4376025 bedfile, though we may want to set the default column type for numeric values as double (d), instead of integer (i)

Check SRR1559031 bedfile for problems:

```{r}
SRR1559031_problems <- vroom::vroom(SRR1559031_bedfile)

vroom::problems(SRR1559031_problems)
```

One source of the merging problem is in row 434393 in SRR1559031, where the position coordinates seem to be duplicated

Check what line looks like in the bedfile:

```{r}
readLines(file.path(repo_root, "results/target/shiba/SRR1559031/junctions/junctions.bed"))[434393]
```

I also noticed at least one entry in this problem sample where the "chr" field has duplicated values
```{r}
SRR1559031_problems |> dplyr::slice(434391, 434392, 434393)
```

chr9:35660647-35660648 is our problem junction but has high read counts ( > 10 ) so it seems like an important junction to be able to keep.
Maybe a workaround would be to manually fix the chr/start/end coordinates of these cases by extracting them from the ID field, 
assuming these cases are always duplicates...

## Check if the combined junctions.bed has this weird junction coordinate

Check for problems in combined junctions.bed

```{r}
combined_jcn_problems <- vroom::vroom(combined_junctions_file) 

vroom::problems(combined_jcn_problems) |> table()
```
The problem is not in the combined junction bedfile.

Check for the specific junction in combined junctions.bed
```{r}
combined_jcn_problems |>
  dplyr::filter(ID == "chr9:35660647-35660648")
```
The event looks normal in the combined file.

## Check problem junction before fix
```{r}
# read in junctions.bed files created from separate Shiba runs
merged_junctions <- purrr::map(junction_paths, \(file) {
    readr::read_tsv(file,
               # make sure columns are types that we expect
               col_types = readr::cols(.default = "d", ID = "c", chr = "c", start = "c", end = "c"))
  })|>
    # merge junctions tables from multiple samples
    # the resulting table separates junction counts from each sample by columns with the sample ID
    purrr::reduce(\(x, y) dplyr::full_join(x, y, by = c("chr", "start", "end", "ID")))

# check problem junction
merged_junctions |>
  dplyr::filter(ID == "chr9:35660647-35660648")

```

## Proposed new junction merging method:

Combine start/end coords, treating them as

```{r}
# read in junctions.bed files created from separate Shiba runs
merged_junctions <- purrr::map(junction_paths, \(file) {
    readr::read_tsv(file,
               # make sure columns are types that we expect
               col_types = readr::cols(.default = "d", ID = "c", chr = "c", start = "c", end = "c")) |>
    # fix duplicated entries separated by semicolons
    dplyr::mutate(
      dplyr::across(c(chr, start, end),
      \(x) {
        sapply(x, \(val) {
          if (!stringr::str_detect(val, ";")) {
            return(val)
          }
          
          parts <- stringr::str_split(val, ";")[[1]]
          
          if (length(unique(parts)) == 1) {
            parts[1]
          } else {
            break
          }
        }
      )
  
})
)
}
) |>
    # merge junctions tables from multiple samples
    # the resulting table separates junction counts from each sample by columns with the sample ID
    purrr::reduce(\(x, y) dplyr::full_join(x, y, by = c("chr", "start", "end", "ID")))

# Check problem junction
merged_junctions |>
  dplyr::filter(ID == "chr9:35660647-35660648")
```
