---
title: Filtering GTEX and TARGET samples from SRA
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

The goals of this notebook are to filter the following SRA tables for files to download:

-   TARGET SRA table, for pediatric cancer bulk RNA-seq to download. These sequences will be used as a query group for performing splicing

-   GTEX SRA table and metadata matching ages to subject IDs, for samples from the youngest donors

Additionally, this notebook obtains an estimate of how much space files will take up. There is 2.64Tb of free space on the huge OpenStack instance.

Metadata used for this analysis are:

-   `SraRunTable-TARGET.csv`, metadata file of accession IDs for dbGaP samples with the TARGET dataset. Created by downloading the SRA run table on dbGaP, with the "phs000218" filter (parent phs ID for TARGET dataset)
-   `GTEX_SraRunTable.csv`, metadata file of dbGaP accession IDs containing GTEX samples.
-   `GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt`, public annotation file consisting of GTEX sample IDs and their ages in 10-year brackets, obtained from <https://storage.googleapis.com/adult-gtex/annotations/v10/metadata-files/GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt> , which is accessible by clicking the download icon next to "GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt" on the [GTEX v10 metadata page](https://gtexportal.org/home/downloads/adult-gtex/metadata). The download command for this is in `data/scripts/00-reference_download.sh`.

## Setup

```{r, setup, include=FALSE}
# set seed for random subsampling of TARGET dataset
set.seed(1)

# load libraries
library(ggplot2)
```

## Read in input and output paths

Read in directory and file paths

```{r}
# find the project directory
base_dir <- here::here()

# find the root-level repo directory
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# define metadata directory
metadata_dir <- file.path(repo_root, "data/metadata")
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")

# define paths to input files
target_sra_file <- file.path(gtex_target_metadata_dir, "SraRunTable-TARGET.csv")
gtex_sra_file <- file.path(gtex_target_metadata_dir, "GTEX_SraRunTable.csv")
gtex_ages_file <- file.path(gtex_target_metadata_dir, "GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt")

# define path to output files
target_accession_path <- file.path(gtex_target_metadata_dir, "target_accessions.tsv")
gtex_accession_path <- file.path(gtex_target_metadata_dir, "gtex_accessions.tsv")

# read in files as variables
target_sra <- readr::read_csv(target_sra_file, col_types = readr::cols(Bytes = "d", .default = "c"))

gtex_sra <- readr::read_csv(gtex_sra_file, col_types = readr::cols(Bytes = "d", .default = "c")) |>
  dplyr::rename(SUBJID = submitted_subject_id)

gtex_ages <- readr::read_tsv(gtex_ages_file, col_types = readr::cols(.default = "c"))
```

# Filtering GTEX samples for download

Combine GTEX age table with GTEX accessions to filter by age and check how many samples of each age bracket are represented by each tumor type

```{r}
# Merge GTEX SRA table with ages metadata to associate IDs with ages
gtex_sra_ages <- gtex_sra |>
  dplyr::left_join(gtex_ages, by = "SUBJID") |>
  dplyr::filter(
    analyte_type == "RNA:Total RNA",
    # Exclude cell line samples
    !grepl("Cells", body_site)
  ) |>
  dplyr::select(AGE, Run, body_site, Bytes, SUBJID, BioProject, BioSample, `SRA Study`, LibraryLayout, version, create_date, ReleaseDate)

# Summarize number of samples per tissue type in each age bracket
gtex_ages_summary <- gtex_sra_ages |>
  dplyr::group_by(body_site, AGE) |>
  dplyr::summarise(age_count = dplyr::n())
```

check distribution of accession IDs in each version number before filtering out cell lines. Only versions up to 4 are recorded, all others are NA

```{r}
gtex_sra |>
  dplyr::group_by(version) |>
  dplyr::summarise(version_count = dplyr::n())
```

Check distribution of GTEX create dates for accessions without version number, after filtering out cell line data. Unfortunately, no creation date is recorded for versions above 4.

```{r}
gtex_sra_ages |>
  dplyr::filter(is.na(version)) |>
  dplyr::group_by(create_date) |>
  dplyr::summarise(date_count = dplyr::n())
```

Check distribution of GTEX release dates of accession IDs without version numbers. The release date for version 6p1 is 2015 ([source](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000424.v6.p1)), so these values represent samples from later versions that are only accessible on AnVIL.

```{r}
gtex_sra_ages |>
  dplyr::mutate(ReleaseDate = stringr::str_remove(ReleaseDate, "-.*")) |>
  dplyr::group_by(ReleaseDate) |>
  dplyr::summarise(date_count = dplyr::n())
```

Plot distribution of ages across tissue types in GTEX to visually check age bins in each tissue type.

There appear to be samples in the youngest age bracket (20-29 years) for every tissue type, except for "Cervix - Endocervix". Additionally, the next youngest age bracket for "Cervix - Endocervix" samples is 40-49 years, instead of 30-39 years.

```{r}
#| label: fig-gtex_age_distribution

ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_color_brewer(palette = "Set1") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 11),
    axis.text = element_text(size = 6),
    axis.text.x = element_text(
      hjust = 1,
      margin = margin(t = 0, r = 80, b = 0, l = 0),
      size = 5
    )
  ) +
  labs(
    title = "Age distribution of GTEX noncancerous tissue types",
    x = "Tissue type of GTEX sample",
    y = "# samples in each age category"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  coord_cartesian(
    xlim = NULL,
    ylim = c(0, 25),
    expand = TRUE,
    default = FALSE,
    clip = "on"
  )
```

Plot distribution of ages as a percent stacked bar plot.

```{r}
#| label: fig-gtex_age_dist_stacked_bar

ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_color_brewer(palette = "Set1") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 11),
    axis.text = element_text(size = 6),
    axis.text.x = element_text(
      hjust = 1,
      margin = margin(t = 0, r = 80, b = 0, l = 0),
      size = 5
    )
  ) +
  labs(
    title = "Age distribution of GTEX noncancerous tissue types",
    x = "Tissue type of GTEX sample",
    y = "Fraction of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45))
```

Estimate amount of space GTEX samples will take up if cell lines and version numbers above 4 are excluded

```{r}
gtex_sra_version_filtered <- gtex_sra_ages |>
  dplyr::filter(version <= 6)

# estimate amount of space files will take up
gtex_sra_version_filtered_space <- (sum(gtex_sra_version_filtered$Bytes) / 1e12)
paste0("About ", gtex_sra_version_filtered_space, " terabytes of space will be taken up by file downloads.")
```

Print number of samples in all tissue types after excluding cell lines and version numbers above 4

```{r n_in_each_tissue_type_after_version_filtering}
gtex_sra_ages |>
  dplyr::filter(version <= 6) |>
  dplyr::group_by(body_site) |>
  dplyr::summarize(n_after_excluding_versions = dplyr::n())
```

Estimate amount of space GTEX samples will take up if only cell lines are excluded

```{r}
# estimate amount of space files will take up
gtex_sra_space <- (sum(gtex_sra_ages$Bytes) / 1e12)
paste0("About ", gtex_sra_space, " terabytes of space will be taken up by file downloads.")
```

Print number of samples in each tissue type of full GTEX dataset, excluding cell lines only

```{r n-in-each-tissuetype}
gtex_sra_ages |>
  dplyr::group_by(body_site) |>
  dplyr::summarize(n = dplyr::n())
```

Filter for GTEX samples in the youngest age brackets for each tissue type. Subset samples by sorting data by age and taking lowest 7 values per tissue type by minimum age.

```{r}
gtex_age_subsampled <- gtex_sra_ages |>
  # filter for samples from GTEX version 6 or below (versions above this value are stored in AnVIL and cannot be accessed through sra-toolkit)
  # Only versions up to 4 are recorded in this metadata, all other samples have an NA value for this column
  dplyr::filter(version <= 6) |>
  # take smallest age in each bracket and convert to numeric for slicing
  dplyr::mutate(
    age_min = as.numeric(stringr::str_split_i(AGE, "-", 1))
  ) |>
  # select rows with smallest age_min while ignoring ties
  dplyr::slice_min(age_min, n = 7, with_ties = FALSE, by = body_site)

# count how many rows there are
paste0("There are ", nrow(gtex_age_subsampled), " accessions matching the filters.")

# estimate amount of space files will take up
subsampled_gtex_space <- (sum(gtex_age_subsampled$Bytes) / 1e12)
paste0("About ", subsampled_gtex_space, " terabytes of space will be taken up by file downloads.")
```

Print breakdown of number of samples per tissue type after subsampling GTEX

The majority of samples have 7 replicates, but one has 6 (ectocervix) and one has 5 (endocervix)

```{r}
gtex_age_subsampled |>
  dplyr::count(body_site) |>
  tidyr::pivot_wider(
    names_from = n,
    values_from = body_site
  )
```

Summarize size of samples in GTEX subset

```{r gtex_size_table}
gtex_age_subsampled |>
  dplyr::summarise(mean(Bytes), max(Bytes))
```

Summarize library type (paired or unpaired) to inform alignment approach

```{r}
table(gtex_age_subsampled$LibraryLayout)
```

Export filtered and subsampled GTEX accession file

```{r}
# Move run ID to first column, like in TARGET accessions file
gtex_age_subsampled <- gtex_age_subsampled |>
  dplyr::relocate(Run)

readr::write_tsv(gtex_age_subsampled, file = file.path(gtex_accession_path))
```

# Filter TARGET dataset for download

Estimate amount of space all of TARGET will take up, only excluding cell line samples

```{r}
target_no_cell_line_sra <- target_sra |>
  # filter only for accession IDs with sequence files associated with them
  dplyr::filter(
    # Filter for TARGET samples
    biospecimen_repository == "NCI_TARGET",
    # exclude cell line RNA-seq
    study_name != "TARGET: Cancer Model Systems (MDLS): Cell Lines and Xenografts (including PPTP)",
    # filter for samples with fastqs
    grepl("fastq", `DATASTORE filetype`)
  ) |>
  # transform Bytes column to numeric for estimating space
  dplyr::mutate_at(dplyr::vars(Bytes), as.numeric) |>
  # select for only relevant columns
  dplyr::select(
    Run, biospecimen_repository, `DATASTORE filetype`, body_site, Bytes, study_name, `DATASTORE provider`, analyte_type, `Assay Type`, `SRA Study`, BioProject, BioSample, LibraryLayout
  )

# count how many rows there are
paste0("There are ", nrow(target_no_cell_line_sra), " accessions matching the filters.")

# estimate amount of space files will take up
file_space <- sum(target_no_cell_line_sra$Bytes) / 1e12
paste0("About ", file_space, " terabytes of space will be taken up by file downloads.")
```

Filter accession metadata for TARGET RNA-seq datasets

```{r}
# filter accession table for accession IDs of TARGET samples with fastqs
target_filtered_sra <- target_sra |>
  # filter only for accession IDs with sequence files associated with them
  dplyr::filter(
    # Filter for TARGET samples
    biospecimen_repository == "NCI_TARGET",
    # exclude osteosarcomas due to their complex transcriptome
    study_name != "TARGET: Osteosarcoma (OS)",
    # also exclude the pilot phase 1 ALL samples because there are far more ALL samples in pilot phase 2
    study_name != "TARGET: Acute Lymphoblastic Leukemia (ALL) Pilot Phase 1",
    # exclude cell line RNA-seq
    study_name != "TARGET: Cancer Model Systems (MDLS): Cell Lines and Xenografts (including PPTP)",
    # filter for samples with fastqs
    grepl("fastq", `DATASTORE filetype`)
  ) |>
  # transform Bytes column to numeric for estimating space
  dplyr::mutate_at(dplyr::vars(Bytes), as.numeric) |>
  # select for only relevant columns
  dplyr::select(
    Run, biospecimen_repository, `DATASTORE filetype`, body_site, Bytes, study_name, `DATASTORE provider`, analyte_type, `Assay Type`, `SRA Study`, BioProject, BioSample, LibraryLayout
  )

# count how many rows there are
paste0("There are ", nrow(target_filtered_sra), " accessions matching the filters.")

# estimate amount of space files will take up
file_space <- sum(target_filtered_sra$Bytes) / 1e12
paste0("About ", file_space, " terabytes of space will be taken up by file downloads.")
```

Summarize distribution of analyte type for TARGET samples (RNA, DNA, or both)

```{r}
target_filtered_sra |> dplyr::count(analyte_type)
```

Summarize distribution of assay type for TARGET samples

```{r}
target_filtered_sra |> dplyr::count(`Assay Type`)
```

Conclusions: All TARGET samples appear to be RNA-seq. Another dataset with matched DNA & RNA sequencing will be needed to examine SNV status of pediatric cancers.

Summarize distribution of tumor types across TARGET samples:

```{r}
target_filtered_sra |> dplyr::count(study_name)
```

Conclusions: The predominant cancer type in the TARGET dataset consist of leukemias (ALL and AML).

Subsample TARGET dataset to obtain a max of 20 samples per tumor type (aiming to take up \~ half the space of the huge OpenStack instance, \~1.3TB

```{r}
subsampled_target <- target_filtered_sra |>
  dplyr::slice_sample(n = 15, by = study_name)

# count how many rows there are
paste0("There are ", nrow(subsampled_target), " accessions matching the filters.")

# estimate amount of space files will take up
subsampled_file_space <- (sum(subsampled_target$Bytes) / 1e12)
paste0("About ", subsampled_file_space, " terabytes of space will be taken up by file downloads.")
```

Summarize distribution of tumor types across TARGET samples, to confirm subsampling:

```{r}
subsampled_target |> dplyr::count(study_name)
```

Note: Clear Cell Sarcoma samples are the least represented in this subsampled dataset, since it has the smallest N.

Check max Bytes of dataset to estimate max filesize to enable for download

```{r}
subsampled_target |>
  dplyr::summarise(mean(Bytes), max(Bytes))
```

Print Byte info of accession ID that was filling the cache:

```{r target_size_table}
subsampled_target |>
  dplyr::filter(Run == "SRR1559070") |>
  dplyr::select(Run, Bytes, study_name)
```

SRR1559070 is the sample that produced the prefetch and fasterq-dump errors due to its large file size. Since it is the largest file in the subset I am downloading, set the max prefetch file size to a value well above this sample's estimated bytes (50000000 Bytes, or 50 GB).

Check whether libraries are paired or unpaired for alignment

```{r}
table(subsampled_target$LibraryLayout)
```

Export filtered and subsampled TARGET accession file

```{r}
target_accessions <- subsampled_target |>
  dplyr::select(Run, `SRA Study`, BioProject, BioSample, study_name)

readr::write_tsv(target_accessions, file = file.path(target_accession_path))
```

Print session info

```{r}
sessionInfo()
```
