---
title: Summary of samples on splice compendium v1
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

Splice compendium v1 contains data from TARGET and GTEx datasets. This notebook summarizes

Metadata used for this analysis are:

- [sample_sheet.tsv](https://github.com/UCSC-Treehouse/splicing-compendium/blob/main/config/sample_sheet.tsv): Accessions list of samples processed in splice compendium v1. Columns include accession ID ("sample") and dataset ("group").
- [gtex_accessions.tsv](https://github.com/UCSC-Treehouse/splicing-compendium/blob/main/metadata/filter_target_gtex/gtex_accessions.tsv): metadata file of GTEX sample accession IDs, along with age of donor and tissue type of the sample.
- [target_accessions_clinical.tsv](https://github.com/UCSC-Treehouse/splicing-compendium/blob/celiang/new-clinical-metadata/metadata/filter_target_gtex/target_accessions_clinical.tsv): metadata file of TARGET sample accession IDs, along with age of donor, cancer type of the sample, and sequencing center of origin of the sample.

## Setup

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)
library(patchwork)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)
```

## Read in input and output paths

Read in directory and file paths

```{r}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# find the project directory (splicing-compendium)
# working directory is in scripts
base_dir <- here::here()

## Directories ##
# define metadata directory
metadata_dir <- file.path(base_dir, "metadata")
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")

# config directory of compendium
config_dir <- file.path(repo_root, "config")

### Files ##
# gtex metadata with age info
gtex_metadata_file <- file.path(gtex_target_metadata_dir, "gtex_accessions.tsv")
# target metadata with age info
target_metadata_file <- file.path(gtex_target_metadata_dir, "target_accessions_clinical.tsv")
# sample sheet of compendium files 
compendium_sample_file <- file.path(config_dir, "sample_sheet.tsv")

## Read in files ##
gtex_metadata <- readr::read_tsv(
  gtex_metadata_file, 
  col_types = readr::cols(Bytes = "d", 
                          .default = "c")
)

target_metadata <- readr::read_tsv(
  target_metadata_file,
  col_types = readr::cols(age_at_earliest_diagnosis_in_years.diagnoses.xena_derived = "d",
  .default = "c")
)

sample_df <- readr::read_tsv(
  compendium_sample_file,
  col_types = readr::cols(.default = "c")
)
```

## Filter TARGET and GTEx samples for what's in the compendium

Not all samples filtered for downloading made it to the compendium due to `fasterq-dump` download issues or the file size being too large to align in a timely manner.

```{r filter_for_compendium_samples}

# obtain list of accessions in compendium for each group

target_accessions_list <- compendium_samples |>
  dplyr::filter(group == "target") |>
  dplyr::pull(sample)

gtex_accessions_list <- compendium_samples |>
  dplyr::filter(group == "gtex") |>
  dplyr::pull(sample)

# filter gtex and target metadata for those corresponding to samples in compendium

target_compendium_df <- target_metadata |>
  dplyr::filter(Run %in% target_accessions_list)

gtex_compendium_df <- gtex_metadata |>
  dplyr::filter(Run %in% gtex_accessions_list)

```

## Summarize sample composition of GTEx accessions

### Table of number of samples in each tissue type
Print a table summarizing number of samples in each tissue type in compendium

```{r gtex_num_tissue_types_summary_table}
# summarize number of samples in each tissue type
gtex_compendium_df |>
  dplyr::group_by(body_site) |>
  dplyr::summarise(
    n = dplyr::n()
    )
```
Our curated GTEx compendium dataset is primarily composed of whole blood (for leukemia comparisons) and skeletal muscle (for soft tissue sarcoma comparison) samples.

### Table of age distribution for the GTEx samples in the compendium

Print a table of the age distribution breakdown of GTEx compendium samples 

```{r gtex_ages_summary_table}

# Summarize number of samples per tissue type in each age bracket
gtex_ages_summary <- gtex_compendium_df |>
  dplyr::group_by(body_site, AGE) |>
  dplyr::summarise(age_count = dplyr::n())

gtex_ages_summary
```

As expected, there is a small fraction of filtered GTEx samples under the PEDAYA (\< 30 years) category.
A future addition of value for the compendium will be to include tissue samples from the developmental GTEx project.

### Plot of age distribution of GTEx samples in compendium

```{r}
#| label: fig-compendium_gtex_age_stacked_bar
#| fig-width: 11
#| fig-height: 8

frac_age_plot <- ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    x = "Tissue type of GTEX sample",
    y = "Fraction of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45),
                   labels = function(x){stringr::str_wrap(x, 30)}) 

number_age_plot <- ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    x = "Tissue type of GTEX sample",
    y = "Number of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45), 
                   labels = function(x){stringr::str_wrap(x, 30)}) 

# Plot age distribution together
combined_age_plot <- number_age_plot +
frac_age_plot +
  plot_annotation(
    title = "Age distribution of GTEX tissue types in compendium",
    theme = theme(plot.title = element_text(size = 22),
                  plot.margin = margin(t = 10, r = 10, b = 70, l = 50)
  )
  )
  
combined_age_plot + plot_layout(guides = "collect") + plot_layout(axes = "collect")
```

### Table of sequencing center of origin for compendium GTEx samples

Print a table of the breakdown of what sequencing center the tissue types are sequenced from.
This information is for checking what sorts of batch effects may be expected within tissue types.

```{r gtex_sequencing_center}
# summarize number of samples in each tissue type come from what sequencing center
gtex_compendium_df |>
  dplyr::group_by(body_site, `Center Name`) |>
  dplyr::summarise(
    n = dplyr::n()
    )

```

All tissue types in the GTEx datasets come from the same sequencing center.

## Summarize sample composition of TARGET accessions in compendium

### Table of number of samples in each cancer type

Print a table summarizing number of samples in each tissue type in compendium
```{r target_ages_summary_table}
# summarize number of samples in each tissue type
target_compendium_df |>
  dplyr::group_by(study_name) |>
  # the xena browser metadata's "name.project" has the cancer type in a prettier format 
  # (no "TARGET:" prefix) 
  # but since some ALL phase 2 samples were in ALL phase 3 xena metadata I am afraid to use the xena browser project name
  dplyr::summarise(
    n = dplyr::n()
    )
```

### Table of age distribution for the TARGET samples in the compendium

```{r target_ages_summary_table}
# Summarize number of samples per tissue type in each age bracket
target_ages_summary <- target_compendium_df |>
  dplyr::mutate(
    # bin TARGET sample age in years into 10 year blocks to make data closer to GTEx representation
    age_bin = cut(
    age_at_earliest_diagnosis_in_years.diagnoses.xena_derived,
    breaks = seq(0, 100, by = 10)),
    # remove TARGET prefix from study_name values
    study_name = stringr::str_remove(study_name, "TARGET: ")) |>
  dplyr::group_by(study_name, age_bin) |>
  dplyr::summarise(age_count = dplyr::n())

target_ages_summary
```
Check why NA values are present in 4 ALL Phase 2 samples
```{r check_target_age_nas}
target_compendium_df |>
  dplyr::filter(is.na(age_at_earliest_diagnosis_in_years.diagnoses.xena_derived))
```
The NA values correspond to 4 TARGET ALL Phase 2 samples (mislabeled as phase 3 in Xena browser metadata) without age at diagnosis information.

### Plot of age distribution of TARGET samples in compendium

```{r}
#| label: fig-compendium_target_age_stacked_bar
#| fig-width: 20
#| fig-height: 8

target_frac_age_plot <- ggplot(target_ages_summary, aes(fill = age_bin, x = study_name, y = age_count)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    x = "Tissue type of TARGET sample",
    y = "Fraction of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 1),
                   labels = function(x){stringr::str_wrap(x, 40)}) 

target_number_age_plot <- ggplot(target_ages_summary, aes(fill = age_bin, x = study_name, y = age_count)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    x = "Tissue type of TARGET sample",
    y = "Number of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 1), 
                   labels = function(x){stringr::str_wrap(x, 40)}
                   ) 

# Plot age distribution together
target_combined_age_plot <- target_number_age_plot +
target_frac_age_plot +
  plot_annotation(
    title = "Age distribution of TARGET cancer types in compendium",
    theme = theme(plot.title = element_text(size = 22),
                  plot.margin = margin(t = 0, r = 10, b = 70, l = 120)
  )
  )
  
target_combined_age_plot + plot_layout(guides = "collect") + plot_layout(axes = "collect")
```
### Table of sex distribution for the TARGET samples in the compendium

```{r target_sex_summary_table}
# Summarize number of samples per tissue type in each age bracket
target_sex_summary <- target_compendium_df |>
  dplyr::mutate(
    # remove TARGET prefix from study_name values
    study_name = stringr::str_remove(study_name, "TARGET: ")) |>
  dplyr::group_by(study_name, gender.demographic) |>
  dplyr::summarise(sex_count = dplyr::n())

target_sex_summary
```
### Plots of sex distribution of TARGET samples in compendium
```{r}
#| label: fig-compendium_target_sex_stacked_bar
#| fig-width: 20
#| fig-height: 8

target_frac_sex_plot <- ggplot(target_sex_summary, aes(fill = gender.demographic, x = study_name, y = sex_count)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    x = "Tissue type of TARGET sample",
    y = "Fraction of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 1),
                   labels = function(x){stringr::str_wrap(x, 40)}) 

target_number_sex_plot <- ggplot(target_sex_summary, aes(fill = gender.demographic, x = study_name, y = sex_count)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    x = "Tissue type of TARGET sample",
    y = "Number of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45, n.dodge = 1), 
                   labels = function(x){stringr::str_wrap(x, 40)}
                   ) 

# Plot age distribution together
target_combined_sex_plot <- target_number_sex_plot +
target_frac_sex_plot +
  plot_annotation(
    title = "Sex distribution of TARGET cancer types in compendium",
    theme = theme(plot.title = element_text(size = 22),
                  plot.margin = margin(t = 0, r = 10, b = 70, l = 120)
  )
  )
  
target_combined_sex_plot + plot_layout(guides = "collect") + plot_layout(axes = "collect")
```
CCSK samples in the compendiumae the most sex-imbalanced, being composed completely of male samples.

### Table of sequencing center of origin for compendium TARGET samples
```{r target_sequencing_center}
# summarize number of samples in each tissue type come from what sequencing center
target_compendium_df |>
  dplyr::group_by(study_name, `Center Name`) |>
  dplyr::summarise(
    n = dplyr::n()
    )
```
TARGET AML samples come from two different sequencing centers, so may have batch effects associated with their sequences.

```{r}
sessionInfo()
```
