---
title: Summary of samples on splice compendium v1
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

Splice compendium v1 contains data from TARGET and GTEx datasets. This notebook summarizes

Metadata used for this analysis are:

- [sample_sheet.tsv](https://github.com/UCSC-Treehouse/splicing-compendium/blob/main/config/sample_sheet.tsv): Accessions list of samples processed in splice compendium v1. Columns include accession ID ("sample") and dataset ("group").
- [gtex_accessions.tsv](https://github.com/UCSC-Treehouse/splicing-compendium/blob/main/metadata/filter_target_gtex/gtex_accessions.tsv): metadata file of GTEX sample accession IDs, along with age of donor and tissue type of the sample.
- [target_accessions_clinical.tsv](https://github.com/UCSC-Treehouse/splicing-compendium/blob/celiang/new-clinical-metadata/metadata/filter_target_gtex/target_accessions_clinical.tsv): metadata file of TARGET sample accession IDs, along with age of donor, cancer type of the sample, and sequencing center of origin of the sample.

## Setup

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)
```

## Read in input and output paths

Read in directory and file paths

```{r}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# find the project directory (splicing-compendium)
# working directory is in scripts
base_dir <- here::here()

## Directories ##
# define metadata directory
metadata_dir <- file.path(base_dir, "metadata")
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")

# config directory of compendium
config_dir <- file.path(repo_root, "config")

### Files ##
# gtex metadata with age info
gtex_metadata_file <- file.path(gtex_target_metadata_dir, "gtex_accessions.tsv")
# target metadata with age info
target_metadata_file <- file.path(gtex_target_metadata_dir, "target_accessions_clinical.tsv")
# sample sheet of compendium files 
compendium_sample_file <- file.path(config_dir, "sample_sheet.tsv")

## Read in files ##
gtex_metadata <- readr::read_tsv(
  gtex_metadata_file, 
  col_types = readr::cols(Bytes = "d", 
                          .default = "c")
)

target_metadata <- readr::read_tsv(
  target_metadata_file,
  col_types = readr::cols(age_at_earliest_diagnosis_in_years.diagnoses.xena_derived = "d",
  .default = "c")
)

sample_df <- readr::read_tsv(
  compendium_sample_file,
  col_types = readr::cols(.default = "c")
)
```

## Filter TARGET and GTEx samples for what's in the compendium

Not all samples filtered for downloading made it to the compendium due to `fasterq-dump` download issues or the file size being too large to align in a timely manner.

```{r filter_for_compendium_samples}

# obtain list of accessions in compendium for each group

target_accessions_list <- compendium_samples |>
  dplyr::filter(group == "target") |>
  dplyr::pull(sample)

gtex_accessions_list <- compendium_samples |>
  dplyr::filter(group == "gtex") |>
  dplyr::pull(sample)

# filter gtex and target metadata for those corresponding to samples in compendium

target_compendium_df <- target_metadata |>
  dplyr::filter(Run %in% target_accessions_list)

gtex_compendium_df <- gtex_metadata |>
  dplyr::filter(Run %in% gtex_accessions_list)

```

## Summarize compendium GTEx samples

Check library selection method of GTEx samples that are paired:

```{r}
unique(gtex_select$LibrarySelection)
```

The metadata says cDNA, but the [GTEx methods](https://www.gtexportal.org/home/methods) site says all their versions are polyA selected.

### Plot sample composition of filtered GTEx accessions

Plot distribution of ages in the filtered GTEx samples as a percent stacked bar plot.

```{r}
#| label: fig-gtex_age_dist_stacked_bar
#| fig-width: 11
#| fig-height: 8

# Summarize number of samples per tissue type in each age bracket
gtex_ages_summary <- gtex_select |>
  dplyr::group_by(body_site, AGE) |>
  dplyr::summarise(age_count = dplyr::n())

ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    title = "Age distribution of GTEX noncancerous tissue types",
    x = "Tissue type of GTEX sample",
    y = "Fraction of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  plot_theme
```

As expected, there is a very small fraction of filtered GTEx samples under the PEDAYA (\< 30 years) category.

Check breakdown of what sequencing center the tissue types are sequenced from

```{r}
# summarize number of samples in each tissue type come from what sequencing center
gtex_select |>
  dplyr::group_by(body_site, `Center Name`) |>
  dplyr::summarise(
    n = dplyr::n()
    )

```

All samples appear to come from the Broad Institute.

Print table of the number of samples in all tissue types in the filtered GTEx set

```{r n_in_each_tissue_type_after_version_filtering}
gtex_select |>
  dplyr::group_by(body_site) |>
  dplyr::summarize(n_samples = dplyr::n())
```

Summarize number of accessions in each batch

```{r}
gtex_select |>
  dplyr::summarise(
    .by = batch_id,
    n = dplyr::n()
  )
```

### Check GTEx versions in filtered samples

GTEx v8 and above are also [only accessible on anvil](https://www.gtexportal.org/home/protectedDataAccess). We must filter out samples that come from GTEx v8 and above.

Check number of samples in each GTEx version after filtering.

```{r}
gtex_select |>
  dplyr::group_by(version) |>
  dplyr::summarise(version_count = dplyr::n())
```

Only versions up to 3 are recorded in these samples.

The release date for version 8p2 is 2019-07-18 ([source](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000424.v8.p2)), so these values represent samples from later versions that are only accessible on AnVIL. We should filter out samples with release dates greater than or equal to 2018.

```{r}
gtex_select |>
  dplyr::mutate(ReleaseDate = stringr::str_remove(ReleaseDate, "-.*")) |>
  dplyr::group_by(ReleaseDate) |>
  dplyr::summarise(date_count = dplyr::n())
```

All 1098 filtered GTEx samples should be downloadable outside of AnVIL because their release dates are below 2018.

Estimate amount of space filtered GTEx samples will take up

```{r}
# estimate amount of space files will take up
gtex_select_space <- (sum(gtex_select$Bytes) / 1e12)
paste0("About ", gtex_select_space, " terabytes of space will be taken up by file downloads.")
```

## Export GTEx filtered accessions to file

```{r}
readr::write_tsv(gtex_select, file = gtex_accession_path)
```

Print session info

```{r}
sessionInfo()
```
