---
title: Filtering GTEX and TARGET samples from SRA
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

The goals of this notebook are to filter the following SRA tables for files to download:

-   TARGET SRA table, for pediatric cancer bulk RNA-seq to download. These sequences will be used as a query group for performing splicing

-   GTEX SRA table and metadata matching ages to subject IDs, for samples from the youngest donors

Additionally, this notebook obtains an estimate of how much space files will take up. There is 2.64Tb of free space on the huge OpenStack instance.

Metadata used for this analysis are:

-   `SraRunTable-TARGET.csv`, metadata file of accession IDs for dbGaP samples with the TARGET dataset. Created by downloading the SRA run table on dbGaP, with the "phs000218" filter (parent phs ID for TARGET dataset)
-   `GTEX_SraRunTable.csv`, metadata file of dbGaP accession IDs containing GTEX samples.
-   `GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt`, public annotation file consisting of GTEX sample IDs and their ages in 10-year brackets, obtained from <https://storage.googleapis.com/adult-gtex/annotations/v10/metadata-files/GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt> , which is accessible by clicking the download icon next to "GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt" on the [GTEX v10 metadata page](https://gtexportal.org/home/downloads/adult-gtex/metadata). The download command for this is in `data/scripts/00-reference_download.sh`.

## Setup

```{r, setup, include=FALSE}
# set seed for random subsampling of TARGET dataset
set.seed(1)

# load libraries
library(ggplot2)
```

## Read in input and output paths

Read in directory and file paths

```{r}
# find the project directory
base_dir <- here::here()

# find the root-level repo directory
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# define metadata directory
metadata_dir <- file.path(repo_root, "metadata")
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")

# define paths to input files
target_sra_file <- file.path(gtex_target_metadata_dir, "SraRunTable-TARGET.csv")
gtex_sra_file <- file.path(gtex_target_metadata_dir, "GTEX_SraRunTable.csv")
gtex_ages_file <- file.path(gtex_target_metadata_dir, "GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt")

# define path to output files
target_accession_path <- file.path(gtex_target_metadata_dir, "target_accessions.tsv")
gtex_accession_path <- file.path(gtex_target_metadata_dir, "gtex_accessions.tsv")

# read in files as variables
target_sra <- readr::read_csv(target_sra_file, col_types = readr::cols(Bytes = "d", .default = "c"))

gtex_sra <- readr::read_csv(gtex_sra_file, col_types = readr::cols(Bytes = "d", .default = "c")) |>
  dplyr::rename(SUBJID = submitted_subject_id)

gtex_ages <- readr::read_tsv(gtex_ages_file, col_types = readr::cols(.default = "c"))
```

# Filtering GTEX samples for download

Combine GTEX age table with GTEX accessions to filter by age and check how many samples of each age bracket are represented by each tumor type

```{r}

# tissue types of interest
tissue_types <- c("Brain - Nucleus accumbens (basal ganglia)",
                  "Brain - Caudate (basal ganglia)",
                  "Brain - Spinal cord (cervical c-1)",
                  "Brain - Hypothalamus",
                  "Kidney - Cortex",
                  "Brain - Frontal Cortex (BA9)",
                  "Muscle - Skeletal",
                  "Brain - Cerebellum",
                  "Whole Blood",
                  "Brain - Cerebellar Hemisphere",
                  "Brain - Hippocampus",
                  "Brain - Anterior cingulate cortex (BA24)",
                  "Brain - Cortex",
                  "Brain - Putamen (basal ganglia)",
                  "Brain - Substantia nigra",
                  "Brain - Amygdala"
                  )

# Merge GTEX SRA table with ages metadata to associate IDs with ages
gtex_sra_ages <- gtex_sra |>
  dplyr::left_join(gtex_ages, by = "SUBJID") |>
  dplyr::filter(
    analyte_type == "RNA:Total RNA",
    # Exclude cell line samples
    !grepl("Cells", body_site),
    # filter for polyA samples
    LibraryLayout == "PAIRED",
    # filter for tissue types in select list
    body_site %in% tissue_types
  ) |>
  dplyr::select(AGE, Run, body_site, Bytes, SUBJID, BioProject, BioSample, `SRA Study`, LibraryLayout, version, create_date, ReleaseDate, LibrarySelection)

# Summarize number of samples per tissue type in each age bracket
gtex_ages_summary <- gtex_sra_ages |>
  dplyr::group_by(body_site, AGE) |>
  dplyr::summarise(age_count = dplyr::n())
```

We end up with 5405 samples after this filtering.

Check library selection method of GTEx samples that are paired:

```{r}
unique(gtex_sra_ages$LibrarySelection)
```

The metadata says cDNA, but the GTex website says all their versions are polyA selected: <https://www.gtexportal.org/home/methods>

check distribution of accession IDs in each version number after filtering. Only versions up to 4 are recorded, all others are NA

```{r}
gtex_sra_ages |>
  dplyr::group_by(version) |>
  dplyr::summarise(version_count = dplyr::n())
```

Check distribution of GTEX create dates for accessions without version number. No creation date is recorded for samples from unrecorded versions

```{r}
gtex_sra_ages |>
  dplyr::filter(is.na(version)) |>
  dplyr::group_by(create_date) |>
  dplyr::summarise(date_count = dplyr::n())
```

Check distribution of GTEX release dates of accession IDs without version numbers. The release date for version 6p1 is 2015 ([source](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000424.v6.p1)), so these values represent samples from later versions that are only accessible on AnVIL.

Gtex v8 and above are also only accessible on anvil: <https://www.gtexportal.org/home/protectedDataAccess>

```{r}
gtex_sra_ages |>
  dplyr::mutate(ReleaseDate = stringr::str_remove(ReleaseDate, "-.*")) |>
  dplyr::group_by(ReleaseDate) |>
  dplyr::summarise(date_count = dplyr::n())
```

Plot distribution of ages across tissue types in GTEX to visually check age bins in each tissue type.

```{r}
#| label: fig-gtex_age_distribution

ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_color_brewer(palette = "Set1") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 11),
    axis.text = element_text(size = 6),
    axis.text.x = element_text(
      hjust = 1,
      margin = margin(t = 0, r = 80, b = 0, l = 0),
      size = 5
    )
  ) +
  labs(
    title = "Age distribution of GTEX noncancerous tissue types",
    x = "Tissue type of GTEX sample",
    y = "# samples in each age category"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  coord_cartesian(
    xlim = NULL,
    ylim = c(0, 25),
    expand = TRUE,
    default = FALSE,
    clip = "on"
  )
```

Plot distribution of ages as a percent stacked bar plot.

```{r}
#| label: fig-gtex_age_dist_stacked_bar

ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_color_brewer(palette = "Set1") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 11),
    axis.text = element_text(size = 6),
    axis.text.x = element_text(
      hjust = 1,
      margin = margin(t = 0, r = 80, b = 0, l = 0),
      size = 5
    )
  ) +
  labs(
    title = "Age distribution of GTEX noncancerous tissue types",
    x = "Tissue type of GTEX sample",
    y = "Fraction of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45))
```

Estimate amount of space GTEX samples will take up if cell lines and version numbers above 6 are excluded

```{r}
gtex_sra_version_filtered <- gtex_sra_ages |>
  dplyr::filter(ReleaseDate < 2015)

# estimate amount of space files will take up
gtex_sra_version_filtered_space <- (sum(gtex_sra_version_filtered$Bytes) / 1e12)
paste0("About ", gtex_sra_version_filtered_space, " terabytes of space will be taken up by file downloads.")
```

There are 2353 samples fitting this criteria to download.

Print number of samples in all tissue types after excluding cell lines and version numbers above 6

```{r n_in_each_tissue_type_after_version_filtering}
gtex_sra_version_filtered |>
  dplyr::group_by(body_site) |>
  dplyr::summarize(n_after_excluding_versions = dplyr::n())
```

Export filtered and subsampled GTEX accession file

```{r}
# Move run ID to first column, like in TARGET accessions file
gtex_select <- gtex_sra_version_filtered |>
  dplyr::relocate(Run)

readr::write_tsv(gtex_select, file = file.path(gtex_accession_path))
```

# Filter TARGET dataset for download

Estimate amount of space all of TARGET will take up, only excluding cell line samples

```{r}
target_filtered_sra <- target_sra |>
  # filter only for accession IDs with sequence files associated with them
  dplyr::filter(
    # Filter for TARGET samples
    biospecimen_repository == "NCI_TARGET",
    # exclude cell line RNA-seq
    study_name != "TARGET: Cancer Model Systems (MDLS): Cell Lines and Xenografts (including PPTP)",
    # exclude osteosarcomas due to their complex transcriptome and we have no bone gtex
    study_name != "TARGET: Osteosarcoma (OS)",
    # # filter for samples with fastqs
    # grepl("fastq", `DATASTORE filetype`),
    # # filter for paired end reads
    LibraryLayout == "PAIRED",
    # Filter for RNA-seq
    analyte_type == "RNA",
    # filter out ssRNA-seq because I don't know how to process that data
    `Assay Type`!= "ssRNA-seq"
  ) |>
  # transform Bytes column to numeric for estimating space
  dplyr::mutate_at(dplyr::vars(Bytes), as.numeric) |>
  # select for only relevant columns
  dplyr::select(
    Run, biospecimen_repository, `DATASTORE filetype`, body_site, Bytes, study_name, `DATASTORE provider`, analyte_type, `Assay Type`, `SRA Study`, BioProject, BioSample, LibraryLayout, LibrarySelection
  )

# count how many rows there are
paste0("There are ", nrow(target_filtered_sra), " accessions matching the filters.")

# estimate amount of space files will take up
file_space <- sum(target_filtered_sra$Bytes) / 1e12
paste0("About ", file_space, " terabytes of space will be taken up by file downloads.")
```
Summarize distribution of library prep type for TARGET samples (when I filter for "paired", no results show up)

```{r}
target_filtered_sra |> dplyr::count(LibrarySelection)
```

This information unfortunately tells me nothing. 
For now, I will just export the SRA table of the TARGET samples under my existing filters.
I sent a message to ask Maryke how she found out some of her TARGET samples were ribo-D.

Summarize distribution of tumor types across TARGET samples:

```{r}
target_filtered_sra |> dplyr::count(study_name)
```

Export filtered and subsampled TARGET accession file

```{r}
target_accessions <- target_filtered_sra |>
  dplyr::select(Run, `SRA Study`, BioProject, BioSample, study_name)

readr::write_tsv(target_accessions, file = file.path(target_accession_path))
```

Print session info

```{r}
sessionInfo()
```
