---
title: Filtering GTEX samples from SRA
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

The goals of this notebook are to filter the following SRA tables for files to download:

-   GTEx SRA table and metadata matching ages to subject IDs, for samples from the youngest donors

Additionally, this notebook obtains an estimate of how much space files will take up. There is 2.64Tb of free space on the huge OpenStack instance.

Metadata used for this analysis are:

-   `GTEX_SraRunTable.csv`, metadata file of dbGaP accession IDs containing GTEX samples.
-   `GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt`, public annotation file consisting of GTEX sample IDs and their ages in 10-year brackets, obtained from <https://storage.googleapis.com/adult-gtex/annotations/v10/metadata-files/GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt> , which is accessible by clicking the download icon next to "GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt" on the [GTEX v10 metadata page](https://gtexportal.org/home/downloads/adult-gtex/metadata). The download command for this is in `data/scripts/00-reference_download.sh`.

## Setup

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)
```

## Read in input and output paths

Read in directory and file paths

```{r}
# find the project directory (splicing-compendium)
# working directory is in scripts
base_dir <- here::here()

# define metadata directory
metadata_dir <- file.path(base_dir, "metadata")
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")

# define paths to input files
gtex_sra_file <- file.path(gtex_target_metadata_dir, "GTEX_SraRunTable.csv")
gtex_ages_file <- file.path(gtex_target_metadata_dir, "GTEx_Analysis_v10_Annotations_SubjectPhenotypesDS.txt")

# define path to output files
gtex_accession_path <- file.path(gtex_target_metadata_dir, "gtex_accessions.tsv")

# read in files
gtex_sra <- readr::read_csv(gtex_sra_file, col_types = readr::cols(Bytes = "d", .default = "c")) |>
  dplyr::rename(SUBJID = submitted_subject_id)

gtex_ages <- readr::read_tsv(gtex_ages_file, col_types = readr::cols(.default = "c"))
```

## Filtering GTEX samples for download

Combine GTEX age table with GTEX accessions so that we can check how many samples of each age bracket are represented by each tumor type after filtering

```{r}
# Define tissue types of interest
# kidney, muscle, and blood samples
tissue_types <- c(
  "Kidney - Cortex",
  "Muscle - Skeletal",
  "Whole Blood"
)

# Merge GTEX SRA table with ages metadata to associate IDs with ages
# then filter tables for our criteria (RNA, paired-end, select tissue types)
gtex_sra_ages <- gtex_sra |>
  dplyr::left_join(gtex_ages, by = "SUBJID") |>
  dplyr::filter(
    analyte_type == "RNA:Total RNA",
    # Exclude cell line samples
    !grepl("Cells", body_site),
    # filter for polyA samples
    LibraryLayout == "PAIRED",
    # filter for tissue types in select list
    body_site %in% tissue_types,
    # filter for accessions with fastqs
    stringr::str_detect(`DATASTORE filetype`, "sra")
  ) |>
  dplyr::select(AGE, Run, body_site, Bytes, SUBJID, BioProject, BioSample, `SRA Study`, LibraryLayout, version, create_date, ReleaseDate, `DATASTORE filetype`, LibrarySelection, `Center Name`)

dim(gtex_sra_ages)
```

We end up with 973 samples after this filtering.

Check library selection method of GTEx samples that are paired:

```{r}
unique(gtex_sra_ages$LibrarySelection)
```

The metadata says cDNA, but the [GTEx methods](https://www.gtexportal.org/home/methods) site says all their versions are polyA selected.

### Check GTEx versions in filtered samples

GTEx v8 and above are also [only accessible on anvil](https://www.gtexportal.org/home/protectedDataAccess). We must filter out samples that come from GTEx v8 and above.

Check number of samples in each GTEx version after filtering.

```{r}
gtex_sra_ages |>
  dplyr::group_by(version) |>
  dplyr::summarise(version_count = dplyr::n())
```

Only versions up to 3 are recorded in these samples.

The release date for version 8p2 is 2019-07-18 ([source](https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000424.v8.p2)), so these values represent samples from later versions that are only accessible on AnVIL. We should filter out samples with release dates greater than or equal to 2018.

```{r}
gtex_sra_ages |>
  dplyr::mutate(ReleaseDate = stringr::str_remove(ReleaseDate, "-.*")) |>
  dplyr::group_by(ReleaseDate) |>
  dplyr::summarise(date_count = dplyr::n())
```

All 973 filtered GTEx samples should be downloadable outside of AnVIL because their release dates are below 2018.

Estimate amount of space filtered GTEx samples will take up

```{r}
# estimate amount of space files will take up
gtex_sra_ages_space <- (sum(gtex_sra_ages$Bytes) / 1e12)
paste0("About ", gtex_sra_ages_space, " terabytes of space will be taken up by file downloads.")
```

### Plot sample composition of filtered GTEx accessions

Plot distribution of ages in the filtered GTEx samples as a percent stacked bar plot.

```{r}
#| label: fig-gtex_age_dist_stacked_bar
#| fig-width: 11
#| fig-height: 8

# Summarize number of samples per tissue type in each age bracket
gtex_ages_summary <- gtex_sra_ages |>
  dplyr::group_by(body_site, AGE) |>
  dplyr::summarise(age_count = dplyr::n())

ggplot(gtex_ages_summary, aes(fill = AGE, x = body_site, y = age_count)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  plot_theme +
  labs(
    title = "Age distribution of GTEX noncancerous tissue types",
    x = "Tissue type of GTEX sample",
    y = "Fraction of samples"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  plot_theme
```

As expected, there is a very small fraction of filtered GTEx samples under the PEDAYA (\< 30 years) category.

Check breakdown of what sequencing center the tissue types are sequenced from

```{r}
# summarize number of samples in each tissue type come from what sequencing center
gtex_sra_ages |>
  dplyr::group_by(body_site, `Center Name`) |>
  dplyr::summarise(
    n = dplyr::n()
    )

```

All samples appear to come from the Broad Institute

Print table of the number of samples in all tissue types in the filtered GTEx set

```{r n_in_each_tissue_type_after_version_filtering}
gtex_sra_ages |>
  dplyr::group_by(body_site) |>
  dplyr::summarize(n_samples = dplyr::n())
```

### Export filtered GTEX accession file

```{r}
# Move run ID to first column, like in TARGET accessions file
gtex_select <- gtex_sra_ages |>
  dplyr::relocate(Run)

readr::write_tsv(gtex_select, file = gtex_accession_path)
```

Print session info

```{r}
sessionInfo()
```
