---
title: Filtering GTEX and TARGET samples from SRA
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

The goals of this notebook are to filter the following SRA tables for files to download:

-   TARGET SRA table, for pediatric cancer bulk RNA-seq to download. These sequences will be used as a query group for performing splicing

Additionally, this notebook obtains an estimate of how much space files will take up. There is 2.64Tb of free space on the huge OpenStack instance.

Metadata used for this analysis are:

-   `SraRunTable-TARGET.csv`, metadata file of accession IDs for dbGaP samples with the TARGET dataset. Created by downloading the SRA run table on dbGaP, with the "phs000218" filter (parent phs ID for TARGET dataset)

We also use this resource: <https://www.cancer.gov/ccg/research/genome-sequencing/target/using-target-data/technology#all-phase-1>

In order to filter for what samples have gone through polyA selection, check this documentation linked above

we should also include what center did the sequencing (BCC / British Columbia Cancer Agency) as it is a source of batch

-   Prioritize downloading all of TARGET data regardless (just prioritize excluding samples we know for sure are ribo deplete from the ALL paper)

-   Look for info on TARGET pages to filter for what samples are defininitely polyA, try to filter down to samples that we know for sure are polyA and start downloading those

-   NO!!!: For now just start downloading GTEx (since I am reasonably sure they are all polyA)

-   Try to email st jude ?\
    prioritize samples with larger N

## Setup

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)
```

## Read in input and output paths

Read in directory and file paths

```{r}
# find the project directory (splicing-compendium)
# working directory is in scripts
base_dir <- here::here()

# define metadata directory
metadata_dir <- file.path(base_dir, "metadata")
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")

## define paths to input files
# TARGET accessions metadata from dbGaP
target_sra_file <- file.path(gtex_target_metadata_dir, "SraRunTable-TARGET.csv")

## ALL expansion phase 2 metadata
# sample telemetry report that includes sample IDs of ALL expansion phase 2 samples
all_phase_two_telemetry <- file.path(gtex_target_metadata_dir, "phs000464.v23.txt")
# excel sheet that includes sample IDs of 266 ribo-D RNA-seq samples from the TARGET ALL phase 2 expansion
# sample IDs are under "RNA_barcode_D" column in the first tab
ribod_all_phase_two_samples <- file.path(gtex_target_metadata_dir, "41588_2017_BFng3909_MOESM2_ESM.xlsx")

# define path to output files
target_accession_path <- file.path(gtex_target_metadata_dir, "target_accessions.tsv")

# read in files
target_sra <- readr::read_csv(target_sra_file, col_types = readr::cols(Bytes = "d", .default = "c"))
all_phase_two_telemetry <- readr::read_tsv(all_phase_two_telemetry, col_names = TRUE)
all_phase_two_ribod <- readxl::read_excel(ribod_all_phase_two_samples, sheet = "Table S1 cohort")
```

## Filter TARGET dataset for download

Estimate amount of space filtered TARGET samples will take up

```{r}
target_filtered_sra <- target_sra |>
  # filter for TARGET accessions
  dplyr::filter(
    # Filter for TARGET samples
    biospecimen_repository == "NCI_TARGET",
    # exclude cell line RNA-seq
    study_name != "TARGET: Cancer Model Systems (MDLS): Cell Lines and Xenografts (including PPTP)",
    # exclude osteosarcomas because they have a complex transcriptome and we have no bone GTEx comparator
    study_name != "TARGET: Osteosarcoma (OS)",
    # filter for samples with fastqs
    grepl("fastq", `DATASTORE filetype`),
    # filter for paired end reads
    LibraryLayout == "PAIRED",
    # Filter for RNA-seq
    analyte_type == "RNA"
  ) |>
  # transform Bytes column to numeric for estimating space
  dplyr::mutate_at(dplyr::vars(Bytes), as.numeric)

  # don't select for relevant columns for now in case they become useful
  # select for only relevant columns
  # dplyr::select(
  #   Run, biospecimen_repository, `DATASTORE filetype`, body_site, Bytes, study_name, `DATASTORE provider`, analyte_type, `Assay Type`, `SRA Study`, BioProject, BioSample, LibraryLayout, LibrarySelection
  # )

# count how many rows there are
paste0("There are ", nrow(target_filtered_sra), " accessions matching the filters.")

# estimate amount of space files will take up
file_space <- sum(target_filtered_sra$Bytes) / 1e12
paste0("About ", file_space, " terabytes of space will be taken up by file downloads.")
```

## Make list of samples we know are ribo-D

### TARGET ALL Expansion Phase 2

This excel sheet has 266 RNAseq samples that are riboD
"DNA barcode" column corresponds to sample_)_id column in the sample telemetry report for the study: https://www.ncbi.nlm.nih.gov/gap/sstr/report/phs000464.v23.p8


Summarize distribution of tumor types across TARGET samples:

```{r}
target_filtered_sra |> dplyr::count(study_name)
```

Export filtered and subsampled TARGET accession file

```{r}
target_accessions <- target_filtered_sra |>
  dplyr::select(Run, `SRA Study`, BioProject, BioSample, study_name)

readr::write_tsv(target_accessions, file = file.path(target_accession_path))
```

Print session info

```{r}
sessionInfo()
```
