---
title: "compare merged sample psi matrix target pilot"
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

The Treehouse Splice Compendium workflow has been modified to allow for all 2581 samples of the compendium to be run given the Pheonix cluster's memory limits. These changes consist of taking the merged junctions bedfile containing counts and junction positions of all samples, splitting it into one file per sample so that each per-sample bedfile contains junction positions from all junctions, but only quantifications for one sample, and calculating per-sample PSI values using this per-sample junction count with event coordinate positions defined from the GTF made from all samples.

Next, to create the PSI tables for compendium users, these per-sample PSI tables are merged into one table per Shiba output. The resulting merged tables are:

```         

se = "PSI_SE.txt",
afe = "PSI_AFE.txt",
ale = "PSI_ALE.txt",
five = "PSI_FIVE.txt", 
three = "PSI_THREE.txt",
mse = "PSI_MSE.txt",
mxe = "PSI_MXE.txt",
ri = "PSI_RI.txt",
matrix = "PSI_matrix_sample.txt"
```

Where "PSI\_\[EVENT\]" refers to PSI quantifications for each event type, with each row being a unique event's position ID and each column being event information such as gene ID assigned to the event, position coordinates of features such as introns and exons used to define that event, annotation status of that event, junction counts for each of these features for each sample, and PSI values corresponding to that event per sample.

"PSI_matrix_sample.txt" corresponds to the sample-level event matrix. Splice events for all event types are present in this matrix. Each row is a unique splice event. In this martix, columns include the position ID (coordinates) of the splice event, a Shiba-assigned "event_ID" consisting of the event type in acronyms, followed by a number, and the PSI values for each sample.

The goal of this notebook is to check if the merged tables generated from the target pilot pilot samples are the same as the canonical PSI tables.

## Setup

```{r read_files}
## Directories ##
# find project root to access separate results dir
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# result directories
results_dir <- file.path(repo_root, "results")
compendium_results_dir <- file.path(results_dir, "merged_shiba")
version_dir <- file.path(compendium_results_dir, "v1.1_reuse_table_target_pilot")
psi_dir <- file.path(version_dir, "merged_persample_psi")

# canonical shiba psi matrix dir
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merging-psi-tables")
target_pilot_dir <- file.path(merge_exploration_dir, "target_pilot")
# directory of shiba results produced from the same shiba run
combined_dir <- file.path(target_pilot_dir, "shiba_combined", "results")
# psi table directory
combined_splice_results_dir <- file.path(combined_dir, "splicing")

## Files ##

merged_file_list <- c(
  se = "PSI_SE.txt.gz",
  afe = "PSI_AFE.txt.gz",
  ale = "PSI_ALE.txt.gz",
  five = "PSI_FIVE.tx.gz",
  three = "PSI_THREE.txt.gz",
  mse = "PSI_MSE.txt.gz",
  mxe = "PSI_MXE.txt.gz",
  ri = "PSI_RI.txt.gz",
  matrix = "PSI_matrix_sample.txt.gz"
)

combined_file_list <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.tx",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt",
  matrix = "PSI_matrix_sample.txt"
)

# Construct paths to merged and combined sets of files

merged_files <- file.path(psi_dir, merged_file_list)
names(merged_files) <- names(merged_file_list)
combined_files <- file.path(combined_splice_results_dir, combined_file_list)
names(combined_files) <- names(combined_file_list)

```

Define functions

```{r functions}

# read in and merge splice tables into one table of all event types
# final table only contains pos_id, gene_id, label and sample PSI values
# usage: read_psi_tables(paths_to_psi_event_tables)
read_psi_tables <- function(psi_paths) {
  psi_results <- psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric 
      dplyr::mutate(across(contains("_PSI"), \(x) as.numeric(x)))
  })

# combine PSI values of samples run together into one dataframe to compare against other PSI tables
psi_table <- purrr::list_rbind(psi_results, names_to = "event_type")

# return the psi table
return(psi_table)
}

## Manipulating PSI tables for comparison ##

# merge tables by event_type", "pos_id", and "label" only (consider event the same if position is the same regardless of gene)
join_tables <- function(table1, table2) {
  
  # create the merged table of PSI tables to compare
  joined_table <- dplyr::full_join(
  table1,
  table2,
  by = c("event_type", "pos_id", "label"),
  # label PSI values by table they came from
  suffix = c("_table1", "_table2")) |>
  # categorize events by whether they are in the v1 or rep2 tables
  dplyr::mutate(
    # Count events in one table or other if the gene in the other table is NA
    table1_only = is.na(gene_id_table2),
    table2_only = is.na(gene_id_table1),
    shared_pos_id = !(table1_only | table2_only),
    gene_difference = shared_pos_id & gene_id_table1 != gene_id_table2,
    # Label events that are completely identical if event_type/pos_id/label fields are the same and gene_ids are the same
    identical_pos_id_and_gene_id = shared_pos_id & ! gene_difference
  )
  
  return(joined_table)

}

# Summarize events shared and unshared between tables
summarize_shared_events <- function(joined_table) {
  # calculate fraction of events only present in either table and fraction events that are shared
  shared_events_summary <- joined_table |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    total = dplyr::n(),
    # count event types that have share position IDs (includes events with different gene_ids)
    shared_pos_id_count = sum(shared_pos_id),
    # count events in each share category
    shared_pos_and_gene_id_count = sum(identical_pos_id_and_gene_id),
    shared_pos_diff_gene_count = sum(gene_difference),
    table1_only_count = sum(table1_only),
    table2_only_count = sum(table2_only),
    # calculate percentages of each category
    shared_pos_and_gene_id_percent = shared_pos_and_gene_id_count / total * 100,
    shared_pos_diff_gene_percent = shared_pos_diff_gene_count / total * 100,
    table1_only_percent = table1_only_count / total * 100,
    table2_only_percent = table2_only_count / total * 100,
  )
  
  # return summary table
  return(shared_events_summary)
}

# compare if PSI values are identical for a set of event IDs
# usage: 

# compare shared events between two psi tables
compare_shared_events <- function(
    comparison_table, 
    psi_table1, 
    psi_table2, 
    event_id_filter_expr, 
    event_type_filter_expr
    ) {
  shared_ids <- comparison_table |>
    dplyr::filter(
      {{ event_id_filter_expr }},
      {{ event_type_filter_expr }}
    ) |>
    dplyr::pull(pos_id)

  # extract pos_ids from filtered IDs to compare
  extract_shared <- function(psi_table) {
    psi_table |>
      dplyr::filter(pos_id %in% shared_ids) |>
      # filter only for fields to compare (pos_id, psi values of samples)
      dplyr::select(pos_id, ends_with("_PSI")) |>
      # sort by pos_id so order is the same
      dplyr::arrange(pos_id)
  }

  all.equal(extract_shared(psi_table1), extract_shared(psi_table2))
}

```



Based on previous comparison work done in exploration/ we know that retained intron events are different 
(different PSI values, even for events with the same position ID between splice compendium workflow method results and canonical Shiba results). 
So RI event types are filtered out in all matrices prior to comparison.

## Compare PSI sample matrices

Read in matrices to compare and filter out ri events

```{r}
merged_matrix_df <- readr::read_tsv(merged_files[["matrix"]], col_types = readr::cols(.default = "c")) |>
    dplyr::filter(!stringr::str_detect(event_id,"RI_")) |>
  # sort by pos id
  dplyr::arrange(pos_id)

combined_matrix_df <- readr::read_tsv(combined_files[["matrix"]], col_types = readr::cols(.default = "c")) |>
    dplyr::filter(!stringr::str_detect(event_id,"RI_")) |>
  # sort by pos id
  dplyr::arrange(pos_id)
```

Check if matrices are identical - matrices are not identical on Shiba-assigned event IDs

```{r}
all.equal(merged_matrix_df, combined_matrix_df)
```

Check if, excluding event IDs (SE_1, SE_2, etc), matrices are identical:

```{r}
merged_matrix_df_no_id <- merged_matrix_df |>
  dplyr::select(!event_id) |>
  # sort by pos id
  dplyr::arrange(pos_id)

combined_matrix_df_no_id <- combined_matrix_df |>
  dplyr::select(!event_id) |>
  # sort by pos id
  dplyr::arrange(pos_id)

all.equal(merged_matrix_df_no_id, combined_matrix_df_no_id)
```

Merged PSI matrices, excluding RI events, are the same as the canonical PSI sample matrix, except for Shiba-assigned event IDs

Remove PSI matrices from memory prior to comparing other event tables

```{r}
# removed matrices that won't be compared anymore
rm(merged_matrix_df)
rm(combined_matrix_df)
rm(merged_matrix_df_no_id)
rm(combined_matrix_df_no_id)

# garbage collector
gc()
```
## Compare event PSI tables

Knowing that `event_id` will differ in the results generated from the two methods, this column will not be included in subsequent comparisons. 

### SE
```{r read_event_tables}
merged_events_df <- read_psi_tables(merged_files[1:8])
combined_events_df <- read_psi_tables(combined_files[1:8])
```

