---
title: "Test Wilcoxon ranksums on TARGET subset"
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  use_cache: TRUE
  min_samples: 5
  padj: 0.05
  dPSI: 0.15
---

## Introduction
This notebook performs differential analysis on splice event usage (PSI) values generated from the splice compendium workflow. 
The analyses included in this notebook consist of:

1. Statistical test for differential splicing: A Wilcoxon ranksum test is used to compare the PSI value distributions for splice events used in a reference and query group of samples.
P-values are corrected for multiple testing using the Benjamini-Hochberg method.

2. Calculating the magnitude of differential splicing: The magnitude of differential splicing is calculated for each splice event by taking the difference between the median PSI value between the query and reference group.

3. Visualizing splice events by significance and magnitude of differential splicing: Volcano plots are created to plot the adjusted p-value and dPSI values of each splice event type between the query and reference groups.
One Volcano plot is created for all splice event types, and separate volcano plots are provided to split splice event types by category.

4. Filtering splice event types by significance score and ranking by magnitude of differential splicing: splice event position IDs, their dPSI, and adjusted p-values are finally exported as a TSV based on user-defined adjusted p-value and dPSI filters. 
These splice events are ranked in decreasing dPSI magnitude.

## Set up
## load libraries
```{r libraries_and_theme}
# load libraries
library(ggplot2)

set.seed(1)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)
```

## Directories and files

```{r dirs_and_files}
data_dir <- file.path("data")

# target subset metadata
target_metadata_file <- file.path(data_dir, "SraRunTable-TARGET.csv")
# target subset combined psi results
combined_psi_file <- file.path(data_dir, "combined_psi_results.rds")

# output
wilcox_results_file <- file.path(data_dir, "wilcox_test_results.tsv")
```

## read in files

```{r read_in_files}
# read in TARGET sample metadata
target_metadata <- readr::read_csv(target_metadata_file, col_types = c(.default = "c"))

# combined PSI results from notebook 01-target_subset_psi_value_comparison
psi_combined <- readRDS(combined_psi_file)
```

## Combine PSI table with TARGET metadata to associate samples with cancer type
```{r combine_diff_events_psi_tables}
# Filter for body site and run ID fields in target metadata
filtered_target_metadata <- target_metadata |>
  dplyr::select(study_name, Run) |>
  dplyr::filter(grepl("TARGET", study_name))

# Associate the body site with accession number
# first pivot the table longer 
# each row corresponds to PSI values for a single sample/accession ID 
# columns correspond to PSI quantification for a specific splice event position for that accession ID
cancer_type_psi_table <- psi_combined |>
  tidyr::pivot_longer(
    cols = contains("_PSI"),
    names_pattern = "(.*)_PSI",
    names_to = "Run",
    values_to = "PSI"
    ) |>
  dplyr::inner_join(
    filtered_target_metadata, 
    by = c("Run")) |>
  # assign group type (ref vs query)
    dplyr::mutate(
    group = dplyr::case_when(
      study_name ==  "TARGET: Kidney\\, Wilms Tumor (WT)" ~ "query",
      study_name !=  "TARGET: Kidney\\, Wilms Tumor (WT)" ~ "ref"
    )
  )

# Print column names of combined PSI table
colnames(cancer_type_psi_table)
```

Filter to positions where there are sufficient samples.

```{r filter_min_observations}
# require at least min_samples for each group to have non-NA PSI values
cancer_type_psi_table_filtered <- cancer_type_psi_table |>
  tidyr::drop_na(PSI) |>
  dplyr::group_by(pos_id) |>
  dplyr::mutate(
    ref_count = sum(group == "ref"), 
    query_count = sum(group == "query")
  ) |>
  dplyr::filter(
    ref_count >= params$min_samples, 
    query_count >= params$min_samples
  )|>
  # filter out events where all PSI values are identical (nothing different to compare)
  dplyr::filter(any(PSI != PSI[1])) |>
  dplyr::ungroup()

# obtain list of unique position IDs in the table filtered for min_samples
filtered_events <- unique(cancer_type_psi_table_filtered$pos_id)

# subset PSI table for the first 1000 events to save memory
cancer_type_psi_table_subset <- cancer_type_psi_table_filtered |>
  dplyr::filter(pos_id %in% filtered_events[1:1000])

```

### Calculate significance of differential splicing 

Calculate Wilcoxon ranksum for PSI distributions

```{r used_cached_wilcox}
use_cached_wilcox <- file.exists(wilcox_results_file) && params$use_cache
```

Run wilcoxon ranksum test
```{r calculate_wilcox}
if (use_cached_wilcox) {
  wilcox_test_events <- readr::read_tsv(wilcox_results_file)
} else {
wilcox_test_events <- cancer_type_psi_table_filtered |>
  dplyr::group_by(pos_id) |>
  rstatix::wilcox_test(
    PSI ~ group,
    ref.group = "ref"
  ) |>
  # add multiple testing adjustment
  dplyr::mutate(
    p_adj = p.adjust(p, method = "BH")
  )
  # write test results to output 
  readr::write_tsv(wilcox_test_events, wilcox_results_file)
}
```

### Plot significant wilcox test distributions
Plot PSI distributions of splice events with significantly different distributions between query and reference groups.
```{r pass_wilcox_test_dist}
#| label: fig-hist_sig_wilcox_test_dists
#| fig-width: 11
#| fig-height: 8

pass_wilcox_test_events <- wilcox_test_events |>
  dplyr::filter(p_adj <= params$padj) |>
  # arbitrarily grab a couple to plot distributions
 dplyr::slice_min(p_adj, n = 36)

# pivot longer so distribution can be plotted
pass_wilcox_psi <- cancer_type_psi_table|>
  dplyr::filter(pos_id %in% pass_wilcox_test_events$pos_id) |>
  tidyr::drop_na(PSI)

ggplot(pass_wilcox_psi, aes(x = PSI, y = 0, fill = group)) +
  ggridges::geom_density_ridges(stat = "binline", bins = 20, alpha = 0.5) + 
  facet_wrap(~ pos_id) + 
  labs(title = "PSI distributions of wilcox ranksum smallest p_adjust events",
       x = "PSI value",
       y = "Density") 
```
### Calculate magnitude of differential splicing (dPSI) values
```{r calculated_dPSI}

median_PSI_table <- cancer_type_psi_table_filtered |>
  dplyr::group_by(pos_id, group) |>
  dplyr::mutate(
    median_PSI =
      median(PSI)
      )

dPSI_table <- median_PSI_table |>
  dplyr::group_by(pos_id) |>
  dplyr::mutate(
    dPSI = unique(median_PSI[group == "ref"]) - unique(median_PSI[group == "query"])
  )

```


## Supplemental

### Failed Wilcox test distributions

Filter for distributions that failed the wilcoxon test to see if they look like biologically interesting distributions.
Make a series of histograms faceted by pos_id
```{r failed_wilcox_test_dist}
#| label: fig-hist_fail_wilcox_test_dists
#| fig-width: 11
#| fig-height: 8

fail_wilcox_test_events <- wilcox_test_events |>
  dplyr::filter(p >= 0.05) |>
  # arbitrarily grab a couple to plot distributions
 dplyr::slice_sample(n = 36)

# select data from original data
fail_wilcox_psi <- cancer_type_psi_table |>
  dplyr::filter(pos_id %in% fail_wilcox_test_events$pos_id) |>
  tidyr::drop_na(PSI)

ggplot(fail_wilcox_psi, aes(x = PSI, y = group, fill = group)) +
  ggridges::geom_ridgeline(stat = "binline", bins = 20, scale = 1, alpha = 0.5) + 
  facet_wrap(~ pos_id, ncol = 6) + 
  labs(title = "PSI distributions of wilcox ranksum p >= 0.05 events",
       x = "PSI value",
       y = "Number of samples") 
```

These distributions look like reasonable examples with no differential splicing
