---
title: "Test Wilcoxon ranksums on TARGET subset"
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  use_cache: TRUE
  ref: "all"
  query: "TARGET: Kidney\\, Wilms Tumor (WT)" 
  min_samples: 5
  padj: 0.05
  dPSI: 0.15
---

## Introduction
This notebook performs differential analysis on splice event usage (PSI) values generated from the splice compendium workflow. 
The analyses included in this notebook consist of:

1. Statistical test for differential splicing: A Wilcoxon ranksum test is used to compare the PSI value distributions for splice events used in a reference and query group of samples.
P-values are corrected for multiple testing using the Benjamini-Hochberg method.

2. Calculating the magnitude of differential splicing: The magnitude of differential splicing is calculated for each splice event by taking the difference between the median PSI value between the query and reference group.

3. Filtering splice event types by significance score and ranking by magnitude of differential splicing: splice event position IDs, their dPSI, and adjusted p-values are finally exported as a TSV based on user-defined adjusted p-value and dPSI filters. 
These splice events are ranked in decreasing dPSI magnitude.

## Set up

## load libraries
```{r libraries_and_theme}
# load libraries
library(ggplot2)

set.seed(1)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)
```

## Directories and files

```{r dirs_and_files}
data_dir <- file.path("data")

# target subset metadata
target_metadata_file <- file.path(data_dir, "SraRunTable-TARGET.csv")
# target subset combined psi results
combined_psi_file <- file.path(data_dir, "cleaned_psi_matrix.txt")

# output
wilcox_results_file <- file.path(data_dir, "wilcox_test_results.tsv")
diff_splice_results <- file.path(data_dir, "target_pilot_diffsplice.tsv")
```

## read in files

```{r read_in_files}
# read in TARGET sample metadata
target_metadata <- readr::read_csv(target_metadata_file, col_types = c(.default = "c"))

# combined PSI results from notebook 01-target_subset_psi_value_comparison
psi_combined <- readr::read_tsv(combined_psi_file, col_types = c(.default = "c"))
```

## Combine PSI table with TARGET metadata to associate samples with cancer type
```{r combine_diff_events_psi_tables}
# Filter for body site and run ID fields in target metadata
filtered_target_metadata <- target_metadata |>
  dplyr::select(study_name, Run) |>
  dplyr::filter(grepl("TARGET", study_name))

# Associate the body site with accession number
# first pivot the table longer 
# each row corresponds to PSI values for a single sample/accession ID 
# columns correspond to PSI quantification for a specific splice event position for that accession ID
cancer_type_psi_table <- psi_combined |>
  tidyr::pivot_longer(
    cols = contains("SRR"),
    names_to = "Run",
    values_to = "PSI"
    ) |>
  dplyr::inner_join(
    filtered_target_metadata, 
    by = c("Run")) |>
  # assign group type (ref vs query)
    dplyr::mutate(
    group = dplyr::case_when(
      # set query group as specified in params
      study_name == params$query ~ "query",
      # if ref group is "all", all cancer/tissue types other than query group is set as ref group
      if (params$ref == "all") {
        study_name != params$query ~ "ref"
        # otherwise, use user-supplied ref group name as ref group
      } else {
        study_name == params$ref ~ "ref"
      }
    )
  ) |>
  # remove all columns that are not ref or query to save memory
  dplyr::filter(group == "ref" | group == "query")

# Print column names of combined PSI table
colnames(cancer_type_psi_table)
```

Filter to positions where there are sufficient samples.

```{r filter_min_observations}
# require at least min_samples for each group to have non-NA PSI values for a given splice event
cancer_type_psi_table_filtered <- cancer_type_psi_table |>
  tidyr::drop_na(PSI) |>
  dplyr::group_by(pos_id) |>
  # count number of samples in each group with non-NA PSI values for a given splice event
  dplyr::mutate(
    ref_count = sum(group == "ref"), 
    query_count = sum(group == "query")
  ) |>
  dplyr::filter(
    ref_count >= params$min_samples, 
    query_count >= params$min_samples
  )|>
  # filter out events where all PSI values are identical (nothing different to compare)
  dplyr::filter(any(PSI != PSI[1])) |>
  dplyr::ungroup()

# obtain list of unique position IDs in the table filtered for min_samples
filtered_events <- unique(cancer_type_psi_table_filtered$pos_id)

# subset PSI table for the first 100 events to save memory
cancer_type_psi_table_subset <- cancer_type_psi_table_filtered |>
  # I needed to lower this subset from 1:1000 to 1:100 due to memory limitations on my laptop
  dplyr::filter(pos_id %in% filtered_events[1:100])

```

### Calculate significance of differential splicing 

Calculate Wilcoxon ranksum for PSI distributions

```{r used_cached_wilcox}
use_cached_wilcox <- file.exists(wilcox_results_file) && params$use_cache
```

Run wilcoxon ranksum test
```{r calculate_wilcox}
if (use_cached_wilcox) {
  wilcox_test_events <- readr::read_tsv(wilcox_results_file)
} else {
wilcox_test_events <- cancer_type_psi_table_filtered |>
  dplyr::group_by(pos_id) |>
  rstatix::wilcox_test(
    PSI ~ group,
    ref.group = "ref"
  ) |>
  # add multiple testing adjustment
  dplyr::mutate(
    p_adj = p.adjust(p, method = "BH")
  )
  # write test results to output 
  readr::write_tsv(wilcox_test_events, wilcox_results_file)
}
```

### Calculate magnitude of differential splicing (dPSI) values
```{r calculated_dPSI}

# Calculate median PSIs for each splice event, per group
dPSI_table <- cancer_type_psi_table_filtered |>
  dplyr::group_by(pos_id, group) |>
  dplyr::mutate(
    median_PSI =
      median(PSI)
      ) |>
  # Calculate dPSI values by taking the difference between the ref and query group medians
  dplyr::group_by(pos_id) |>
  dplyr::mutate(
    dPSI = unique(median_PSI[group == "query"]) - unique(median_PSI[group == "ref"])
  )
```

### Merge diff splicing test p values with dPSI table
```{r merge_dPSI_with_padj}
# Merge dPSI values with p values from wilcox differential splicing test
p_dPSI_table <- wilcox_test_events |>
  # select for informative columns from wilcox test
  dplyr::select(pos_id, p, p_adj) |>
  # merge wilcox test p values with dPSI table
  dplyr::full_join(dPSI_table, by = "pos_id")

colnames(p_dPSI_table)
```

## write output
```{r}
readr::write_tsv(p_dPSI_table, diff_splice_results)
```
