---
title: "Test Wilcoxon ranksums on TARGET subset"
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Set up
## load libraries
```{r}
# load libraries
library(ggplot2)

set.seed(1)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)
```


## Directories and files

```{r}
data_dir <- file.path("data")

# target subset metadata
target_metadata_file <- file.path(data_dir, "SraRunTable-TARGET.csv")
# target subset combined psi results
combined_psi_file <- file.path(data_dir, "combined_psi_results.rds")

```

## read in files

```{r}
# read in TARGET sample metadata
target_metadata <- readr::read_csv(target_metadata_file, col_types = c(.default = "c"))

# combined PSI results from notebook 01-target_subset_psi_value_comparison
psi_combined <- readRDS(combined_psi_file)
```

## Combine PSI table with TARGET metadata to associate samples with cancer type
```{r}

# We only care about body site and run fields in target metadata
filtered_target_metadata <- target_metadata |>
  dplyr::select(study_name, Run) |>
  dplyr::filter(grepl("TARGET", study_name))

# we need to associate the body site with accession number, so first pivot the table longer (so that we are back to one column of just accession IDs)
cancer_type_psi_table <- psi_combined |>
  tidyr::pivot_longer(
    cols = contains("_PSI"),
    names_pattern = "(.*)_PSI",
    names_to = "Run",
    values_to = "PSI"
    ) |>
  dplyr::inner_join(
    filtered_target_metadata, 
    by = c("Run")) |>
  # assign group type (ref vs query)
    dplyr::mutate(
    group = dplyr::case_when(
      study_name ==  "TARGET: Kidney\\, Wilms Tumor (WT)" ~ "query",
      study_name !=  "TARGET: Kidney\\, Wilms Tumor (WT)" ~ "ref"
    )
  )

```

Filter to positions where there are sufficient samples.

```{r}
min_samples <- 5

# require at least min_samples for each group to have non-NA PSI values
cancer_type_psi_table_filtered <- cancer_type_psi_table |>
  tidyr::drop_na(PSI) |>
  dplyr::group_by(pos_id) |>
  dplyr::mutate(
    ref_count = sum(group == "ref"), 
    query_count = sum(group == "query")
  ) |>
  dplyr::filter(
    ref_count >= min_samples, 
    query_count >= min_samples
  )|>
  # filter where all PSI values are identical
  dplyr::filter(any(PSI != PSI[1])) |>
  dplyr::ungroup()


filtered_events <- unique(cancer_type_psi_table_filtered$pos_id)

cancer_type_psi_table_subset <- cancer_type_psi_table_filtered |>
  dplyr::filter(pos_id %in% filtered_events[1:1000])

```



Calculate Wilcoxon ranksum for PSI distributions

Run wilcoxon ranksum test
```{r}
wilcox_test_events <- cancer_type_psi_table_subset |>
  dplyr::group_by(pos_id) |>
  rstatix::wilcox_test(
    PSI ~ group,
    ref.group = "ref"
  ) |>
  # add multiple testing adjustment
  dplyr::mutate(
    p_adj = p.adjust(p, method = "BH")
  )

```

Filter for distributions that failed the wilcoxon test to see if they look like biologically interesting distributions
Make series of histograms faceted by pos_id
```{r}
#| label: fig-hist_fail_wilcox_test_dists
#| fig-width: 11
#| fig-height: 8

fail_wilcox_test_events <- wilcox_test_events |>
  dplyr::filter(wilcox_p >= 0.05) |>
  # arbitrarily grab a couple to plot distributions
 dplyr::slice_sample(n = 50)

# pivot longer so distribution can be plotted
long_fail_wilcox <- fail_wilcox_test_events |>
  tidyr::pivot_longer(
    contains("_PSI"),
    names_to = "sample", 
    values_to = "PSI"
  ) |>
  # extract group info
  dplyr::mutate(
    group = stringr::str_split_i(sample, "_", i = 3))

ggplot(long_fail_wilcox, aes(x = PSI, y = group, fill = group, alpha = 0.5)) +
  ggridges::geom_ridgeline(stat = "binline", bins = 20, scale = 1) + 
  facet_wrap(~ pos_id) + 
  labs(title = "PSI distributions of wilcox ranksum p >= 0.05 events",
       x = "PSI value",
       y = "Number of samples") 
```
These actually look fairly reasonable for a wilcox test to fail. 
Check what it looks like when the wilcox test is successful
```{r}
#| label: fig-hist_sig_wilcox_test_dists
#| fig-width: 11
#| fig-height: 8

succeed_wilcox_test_events <- wilcox_test_events |>
  dplyr::filter(wilcox_p < 0.05) |>
  # arbitrarily grab a couple to plot distributions
 dplyr::slice_sample(n = 50)

# pivot longer so distribution can be plotted
long_succeed_wilcox <- succeed_wilcox_test_events |>
  tidyr::pivot_longer(
    contains("_PSI"),
    names_to = "sample", 
    values_to = "PSI"
  ) |>
  # extract group info
  dplyr::mutate(
    group = stringr::str_split_i(sample, "_", i = 3))

ggplot(long_succeed_wilcox, aes(x = PSI, y = group, fill = group, alpha = 0.5)) +
  ggridges::geom_ridgeline(stat = "binline", bins = 20, scale = 1) + 
  facet_wrap(~ pos_id) + 
  labs(title = "PSI distributions of wilcox ranksum p < 0.05 events",
       x = "PSI value",
       y = "Number of samples") 
```
I used ridgeplots because stacked histograms made it hard to figure out where the ref and query distributions were


