---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

**Question:** Does merging separate splice tables cause us to lose out on the trustworthiness of unannotated events to an extent that it impacts many samples?

As part of our pipeline, we plan to merge Shiba tables created for individual samples to obtain a final PSI table of all samples.
Before doing so, we use this notebook to test that merging PSI tables from separate runs will not introduce a large amount of untrustworthy splice events.

As part of tracking differences between the two tables, we have labeled NA values introduced at various steps of the data processing pipeline with different negative values:

-   -1: NA values assigned by Shiba (splice events with low coverage)

-   -2: NA values from merging the separate tables together (samples have a NA value after this step if there are genes dropped by Shiba for those samples because there is only one transcript detected for that gene)

-   NA: NA values from merging the separate and combined tables.
    These NAs represent splice events that are only found in one analysis method (combined vs. separate)

We are particularly interested in the impact merging separate tables has on unannotated event detection, since this is one novelty of using Shiba.

**Targeted questions this notebook is trying to answer:**

-   What splice event types do we miss out on with the separate tables method?
-   Of splice events with high numbers of NA values, how many of these events only have a numeric PSI value in one (or a very low amount of) samples?

## Set up

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)
library(patchwork)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)

# color palette for splice event types
cbPalette <- c(
  afe = "#999999",
  ale = "#E69F00",
  five = "#56B4E9",
  mse = "#009E73",
  mxe = "#F0E442",
  ri = "#0072B2",
  se = "#D55E00",
  three = "#CC79A7"
)

# define minimum number of samples to filter splice table by
min_samples <- 5
```

### Define functions

```{r functions}

# PSI distribution plots of events in each method
psi_distributions <- function(
    df,
    event_type_value) {
  df |>
    dplyr::filter(event_type == event_type_value) |>
    ggplot(aes(PSI, fill = event_status)) +
    geom_histogram(bins = 20) +
    facet_wrap(vars(event_type, label, event_status), scales = "free_y") +
    plot_theme +
    # make facet labels bigger
    theme(strip.text.x = element_text(size = global_size),
          # rotate x axis labels so they don't overlap
          axis.text.x = element_text(angle =45))
}

# plot event summary frequencies as percent and raw counts bar plots
plot_event_summary <- function(
    summary_df) {
  # pivot summary df longer for plotting
  long_summary_df <- summary_df |>
    tidyr::pivot_longer(
      cols = c(
        shared_count,
        shared_percent,
        combined_only_count,
        combined_only_percent,
        separate_only_count,
        separate_only_percent
      ),
      # extract percents and counts into separate columns
      names_to = c("category", ".value"),
      names_pattern = "(.+)_(percent|count)"
    )

  # create percent stacked barplots
  percent_plot <-
    ggplot(long_summary_df, aes(fill = category, x = event_type, y = percent)) +
    geom_bar(position = "stack", stat = "identity") +
    plot_theme +
    labs(
      title = "% of splice events in combined, separate, or shared groups",
      x = "Annotation status of event",
      y = "Percentage of events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    plot_theme

  # create raw counts grouped barplot
  counts_plot <-
    ggplot(long_summary_df, aes(fill = category, x = event_type, y = count)) +
    geom_bar(position = "dodge", stat = "identity") +
    plot_theme +
    labs(
      title = "Number of splice events only in combined, separate, or shared groups",
      x = "Annotation status of event",,
      y = "Number of events"
    ) +
    facet_wrap(vars(label), scales = "free_y") +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    plot_theme

  # arrange plots for printing
  percent_plot /
    counts_plot

}

```

## Directories and files

```{r}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# define the data directories
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merging-psi-tables")
target_pilot_dir <- file.path(merge_exploration_dir, "target_pilot")

# directory of shiba results produced from the same shiba run
combined_dir <- file.path(target_pilot_dir, "shiba_combined", "results")

# psi table directory
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# Directory of PSI table, merged from separate shiba runs
separate_psi_table_dir <- file.path(target_pilot_dir, "merged_results")

# merged PSI file of two samples obtained from merge-shiba-psi-tables.qmd
separate_psi_file <- file.path(separate_psi_table_dir, "merged_psi_table.tsv")

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of shiba results of two samples run together
shiba_psi_paths <-file.path(combined_splice_results_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

```

Read in files

```{r}
# read merged splice table from separate runs of Shiba on the TARGET pilot samples
separate_psi_table <- readr::read_tsv(separate_psi_file, col_types=readr::cols(.default = "c")) |>
  # replace remaining NAs from this table representing dropped genes from one sample with no detected splicing for that gene with -2
  dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -2)))

# read in combined splice results to compare against separate results
combined_splice_results <- shiba_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against separate PSI dataframe
combined_psi_table <- purrr::list_rbind(combined_splice_results, names_to = "event_type")
```

### Obtain dimensions of each PSI table

```{r}
# dimensions of separate PSI table
dim(separate_psi_table)
```

```{r}
# dimensions of combined PSI table
dim(combined_psi_table)
```

There are 71,811 more splice events in the separate PSI table than the combined table.

## What splice event types do we miss out on with the separate tables method?

To help us prioritize what splice event types we may be the most confident in after merging separate Shiba PSI tables, we are interested in seeing a breakdown of what event types are most represented in PSI tables constructed from different methods

```{r}
# create df of splice events that are shared between both combined and separate tables
# This is an event-level dataframe (each row is one unique pos_id/event_id)
# columns are each samples' PSI value (with -1/-2/NA values indicating different NA categories)
# combined_event, separate_event, and shared_event columns label what tables the event is present in
all_events <- dplyr::full_join(
  combined_psi_table,
  separate_psi_table,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_combined", "_separate")) |>
  # categorize events by whether they are in the combined or separate tables
  dplyr::mutate(
    # combined events are counted if the values are not all NA
    combined_event = ! dplyr::if_all(ends_with("_combined"), is.na),
    # separate events are counted if the values for the event are not all NA (indicating the specific position ID is only found in the combined or separate table)
    separate_event = ! dplyr::if_all(ends_with("_separate"), is.na),
    shared_event = combined_event & separate_event
  )

```

Print summary of counts across each method, across all event types

```{r}
# number of total events shared or present only in the combined and separate tables
# this table counts events with -1 and -2 PSI values

all_methods_summary <- all_events |> dplyr::summarise(
    # count number of events in each event type and annotation category
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    separate_only_count = sum(separate_event) - shared_count
)

all_methods_summary
```

Print summary of counts and percentages across each method, broken down by event type

```{r}
# event-level summary of the number of each splice event type in each splice table
all_events_summary <- all_events |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    total = dplyr::n(),
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    separate_only_count = sum(separate_event) - shared_count,
    shared_percent = shared_count / total * 100,
    combined_only_percent = combined_only_count / total * 100,
    separate_only_percent = separate_only_count / total * 100,
)

all_events_summary
```

Plot event-level (one count per unique event) summary of event frequency across methods and event types

```{r}
#| label: fig-splice_events_breakdown
#| fig-width: 11
#| fig-height: 8

plot_event_summary(all_events_summary)
```
The majority of annotated event types are shared between both tables, although there are more splice events unique to the separate table (green) than the combined table (red).
Inflated values in the separate tables may be due to -2 NAs which arise in the Shiba processing of separate samples when when genes without multiple transcripts are dropped.

### What specific event types are impacted (missed) by the separate tables method?
These following plots answer the question for me better than the above summary plots.

```{r}
#| label: fig-combined_only_splice_events
#| fig-width: 11
#| fig-height: 8

combined_only_summary <- all_events_summary |>
  dplyr::select(event_type, label, combined_only_count, total, combined_only_percent)

long_summary_df <- combined_only_summary |>
    tidyr::pivot_longer(
      cols = c(
        combined_only_count,
        combined_only_percent
      ),
      # extract percents and counts into separate columns
      names_to = c("category", ".value"),
      names_pattern = "(.+)_(percent|count)"
    )

# create raw counts grouped barplot
counts_plot <- ggplot(long_summary_df, aes(fill = event_type, x = event_type, y = count)) +
    geom_bar(stat = "identity") +
      scale_fill_manual(values = cbPalette) +
    plot_theme +
    labs(
      title = "# annotated and unannotated splice events only in combined table",
      x = "Event Type",
      y = "Number of events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    theme(strip.text = element_text(size = 15))

# create percents grouped barplot
percents_plot <- ggplot(long_summary_df, aes(fill = event_type, x = event_type, y = percent)) +
    geom_bar(position = "dodge", stat = "identity") +
      scale_fill_manual(values = cbPalette) +
    plot_theme +
    labs(
      title = "% annotated and unannotated splice events only in combined table",
      x = "Annotation status of event",
      y = "Percent of all events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    theme(strip.text = element_text(size = 15))


percents_plot /
  counts_plot + plot_layout(guides = "collect")
```

I think of these as "percent of each event category missed in the separate tables".
In the worst cases, 30% of unannotated events might be only in the combined table (FIVE events).
Losing 30% of unannotated real events seems rough.

### How many events remain if we filter for events with PSI values in a minimum number of samples?

-   Of splice events with high numbers of NA values, how many of these events only have a numeric PSI value in one (or a very low amount of) samples?

Add summary columns to dataframe
```{r}
all_events <- all_events |>
  dplyr::mutate(
    # count the number of PSI values that are not one of our NA types
    n_quantified = rowSums(
      # select columns corresponding to sample PSI values that are not one of our NA types
      dplyr::pick(dplyr::matches("(_PSI_combined|_PSI_separate)$")) >= 0, 
      # ignore NA values (events only in separate or combined tables)
      na.rm = TRUE)
  )

# check what summary looks like
all_events |>
  dplyr::select(pos_id, n_quantified, combined_event, separate_event, shared_event) |>
  head()
```


Filter for minimum of 5 samples with numeric PSI values

```{r}
#| label: fig-splice_events_breakdown_5_samples
#| fig-width: 11
#| fig-height: 8



# event-level summary of the number of each splice event type in each splice table
all_events_summary_min_samples <- all_events |>
  # filter for events where there are numeric PSI values in at least 5 samples
  # there is probably a more secure way to do this but the only numeric values in this table should be the PSI values per sample so i think this works
  dplyr::filter(rowSums(dplyr::across(where(is.numeric), ~ . >= 0 )) >= min_samples) |>
  # this data frame still may include -1 and -2 NA values for other splice events but the sample-level analysis later will tell us more
  dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    total = dplyr::n(),
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    separate_only_count = sum(separate_event) - shared_count,
    shared_percent = shared_count / total * 100,
    combined_only_percent = combined_only_count / total * 100,
    separate_only_percent = separate_only_count / total * 100,
)

plot_event_summary(all_events_summary_min_samples)
```

Print table of splice event counts
```{r}
all_events_summary_min_samples |>
  dplyr::select(event_type, label, shared_count, separate_only_count, combined_only_count)
```

This looks better to me, with the caveat that we lose all the combined_only events in @fig-splice_events_breakdown.
The number of shared AFE events in @fig-splice_events_breakdown is \~40,000 so we seem to get most of those AFE events if we ask for splice events that are shared in 5 or more samples.
The reduction in SE events is small too - we go from around 20,000 in @fig-splice_events_breakdown, to slightly under 20,000 here.

