---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

As part of our pipeline, we plan to merge Shiba tables created for individual samples to obtain a final PSI table of all samples.
Before doing so, we use this notebook to test that merging PSI tables from separate runs will not introduce a large amount of untrustworthy splice events or junction counts.

## Set up

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)

# color palette for splice event types
cbPalette <- c(
afe = "#999999",
ale = "#E69F00",
five = "#56B4E9",
mse = "#009E73",
mxe = "#F0E442",
ri = "#0072B2",
se = "#D55E00",
three = "#CC79A7")
```

## Directories and files

```{r}

# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merging-psi-tables")

# define the data directories
# shiba results dir
shiba_dir <- file.path(merge_exploration_dir, "shiba_results")

# directory of shiba results produced from the same shiba run
combined_dir <- file.path(shiba_dir, "combined_run", "target_pilot", "results")

# psi table directory
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# Directory of PSI table, merged from separate shiba runs
separate_psi_table_dir <- file.path(merge_exploration_dir, "target_pilot", "merged_results")

# merged PSI file of two samples obtained from merge-shiba-psi-tables.qmd
separate_psi_file <- file.path(separate_psi_table_dir, "merged_psi_table.tsv")

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of shiba results of two samples run together
shiba_psi_paths <-file.path(combined_splice_results_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

```

Read in files

```{r}
# read merged splice table from separate runs of Shiba on the TARGET pilot samples
separate_psi_table <- readr::read_tsv(separate_psi_file)

# read in combined splice results to compare against separate results
combined_splice_results_paths <- shiba_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      dplyr::mutate(across(contains("_PSI"), as.numeric))
  })

# combine psi values of samples run together into one dataframe to compare against separate PSI dataframe
combined_psi_table <- purrr::list_rbind(combined_splice_results_paths, names_to = "event_type")
```

### Obtain dimensions of each PSI table

```{r}
# dimensions of separate PSI table
dim(separate_psi_table)
```

```{r}
# dimensions of combined PSI table
dim(combined_psi_table)
```

There are 71,811 more splice events in the separate PSI table than the combined table.

## Examine splice events that are consistently present in both combined and separate splice tables

To help us prioritize what splice event types we may be the most confident in after merging separate Shiba PSI tables, we are interested in seeing a breakdown of what event types are most represented in both PSI tables

```{r}
# obtain df of splice events that are shared between both combined and separate tables
all_events <- dplyr::full_join(
  combined_psi_table,
  separate_psi_table,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_combined", "_separate")) |>
  # categorize events by whether they are in the combined or separate tables
  dplyr::mutate(
    combined_event = ! dplyr::if_all(ends_with("_combined"), is.na),
    # separate events are counted if the values for the event are not all NA or -1 (indicating NA introduced from merging)
    separate_event = ! dplyr::if_all(ends_with("_separate"), ~ sign(.) < 0 | is.na(.)),
    shared_event = combined_event & separate_event
  )
```

Print summary of percentage of each event types found in both methods, or in only one

```{r}
all_events_summary <- all_events |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    count = dplyr::n(),
    shared_percent = sum(shared_event) / count * 100,
    combined_only_percent = ( sum(combined_event) - sum(shared_event) ) / count * 100,
    separate_only_percent = ( sum(separate_event) - sum(shared_event) ) / count * 100,
)

all_events_summary
```

### Visualize breakdown of shared events with stacked bar plots

```{r}
#| label: fig-splice_events_breakdown
#| fig-width: 11
#| fig-height: 8

# pivot summary plot longer so we can make grouped bar plots of the percentages
long_all_events_summary <- all_events_summary |>
  tidyr::pivot_longer(
    cols = c(shared_percent, combined_only_percent, separate_only_percent),
    names_to = "category",
    values_to = "percent"
  ) |>
  # concatenate event type and label so unannotated and annotated events can be plotted together
  tidyr::unite(label_event_type, c("label", "event_type"))

ggplot(long_all_events_summary, aes(fill = category, x = label_event_type, y = percent)) +
  geom_bar(position = "dodge", stat = "identity") +
  plot_theme +
  labs(
    title = "% of splice events only in combined PSI tables",
    x = "Annotation status of event",
    y = "Percentage of events"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  plot_theme


```

The majority of annotated event types are shared between both tables, although there are more splice events unique to the separate table than combined table.
The capturing of unannotated event types suffers more from merging separate PSI tables.
In particular, there are more ALE and MXE events only found in the separate PSI tables.
For other unannotated events, there are 10-30% that are only found in the separate PSI.

### Obtain the number of splice events only found in each table

```{r}
# produce dataframe of elements in combined splice results reference dataframe
combined_only_list <- setdiff(combined_psi_table$pos_id, separate_psi_table$pos_id)

# produce dataframe of elements in merged splice results dataframe but not in the reference df
separate_only_list <- setdiff(separate_psi_table$pos_id, combined_psi_table$pos_id)

# events only present in separate tables
# Number of splice events only in separate PSI table
length(separate_only_list)

# Number of splice events only in combined PSI table
length(combined_only_list)
```


### Check how similar PSI values of shared events are

Pivot longer to create scatter plots

```{r}

long_all_events <- all_events |>
  tidyr::pivot_longer(
    cols = matches("_combined|_separate"),
    names_to = c("sample", "method"),
    values_to = "PSI",
    names_pattern = "(.*)_(combined|separate)"
  )

# pivot wider for scatterplots

all_events_by_method <- long_all_events |> tidyr:: pivot_wider(
    names_from = method,
    values_from = PSI
  )
```

Make scatter plot of unannotated events

```{r}
#| label: fig-unannotated_psi_correlation
#| fig-width: 11
#| fig-height: 8

long_unannotated <- all_events_by_method |>
  dplyr::filter(label == "unannotated")

  ggplot(long_unannotated) +
    aes(
      # read in columns to use for x and y from input
      x = combined,
      y = separate,
    ) +
    geom_point(size = 0.5, alpha = 0.5) +
    labs(
      title = "Spearman correlation of unannotated shared splice events",
      x = "Combined run PSI values",
      y = "Separate run PSI values") +
    facet_wrap(vars(event_type)) +

    ggpubr::stat_cor(method = "spearman", label.x = 0.2, label.y = -0.1, color = "blue")
```

Make scatterplot of annotated events

```{r}
#| label: fig-annotated_psi_correlation
#| fig-width: 11
#| fig-height: 8

long_annotated <- all_events_by_method |>
  dplyr::filter(label == "annotated")


  ggplot(long_annotated) +
    aes(
      # read in columns to use for x and y from input
      x = combined,
      y = separate,
    ) +
    geom_point(size = 0.5, alpha = 0.5) +
    labs(
      title = "Spearman correlation of annotated shared splice events",
      x = "Combined run PSI values",
      y = "Separate run PSI values") +
    facet_wrap(vars(event_type)) +

    ggpubr::stat_cor(method = "spearman", label.x = 0.2, label.y = -0.1, color = "blue")
```

### Takeaways so far

-   In the worst case scenario, about 40% of some splice event types across all unique events found in both the combined and separate (merged) tables are only found in the separate tables.
This is concerning if we assume all events found only in the separate tables are false.

-   Most of the events found only in the separate tables are unannotated events, which are the event types we are hoping to capture most in the splice compendium

-   If we only look at splice events shared between both separate and combined PSI tables, the PSI values appear identical.
We may be able to be more confident in our splice table if we filter out everything with NA values (limit ourselves to PSI values of events found in all samples),
but this may severely shrink down the number of splice events in the compendium.

A next step for this notebook will be to compare the number of NA values in each table (I suspect there will be many more NAs in the separate tables)

Another note here is that we assume the splice events in the combined table are the ground truth, but a separate kind of evaluation would be needed to truly ask how many "real" splice events are in each of the tables.
For instance, we maybe would need to do this experiment on simulated reads where we know going in what all the transcripts are, which may be beyond the scope of this project.

## Compare NA values in each table

Create summary table of % NAs from shiba or from merging
```{r}
na_summary_df <- long_all_events |>
  dplyr::summarise(
    .by = c(event_type, label, method),
    na_count = sum(is.na(dplyr::across(everything()))),
    merge_na_count = sum(PSI == -1, na.rm = TRUE),
    total = dplyr::n(),
    percent_na = na_count / total * 100,
    percent_merge_na = merge_na_count / total * 100
  ) |>
  # concatenate label and event type columns into one again for easy plotting
  tidyr::unite(label_event_type, c("label", "event_type"))

na_summary_df
```

Plot % NAs from shiba
```{r}
#| label: fig-percent_na_comparison
#| fig-width: 11
#| fig-height: 8

ggplot(na_summary_df, aes(fill = method, x = label_event_type, y = percent_na)) +
  geom_bar(position = "dodge", stat = "identity") +
  plot_theme +
  labs(
    title = "% of NA values in separate vs. combined dfs",
    x = "Annotation status of event",
    y = "Percent of events"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  plot_theme
```

Similar to unannotated event types being overrepresented in the separate PSI table in  @fig-unshared_splice_events , we see the NA values becoming more of an issue in unannotated event types in the separate PSI table.


Plot % NAs from merging tables
```{r}
#| label: fig-percent_merge_na_comparison
#| fig-width: 11
#| fig-height: 8

ggplot(na_summary_df, aes(fill = method, x = label_event_type, y = percent_merge_na)) +
  geom_bar(position = "dodge", stat = "identity") +
  plot_theme +
  labs(
    title = "% of NA values from merging tables in separate vs. combined dfs",
    x = "Annotation status of event",
    y = "Percent of events"
  ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  plot_theme
```
A good portion of NA values come from merging in the separate PSI tables