---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

**Question:** Does merging separate splice tables cause us to lose out on the trustworthiness of unannotated events to an extent that it impacts many samples?

As part of our pipeline, we plan to merge Shiba tables created for individual samples to obtain a final PSI table of all samples.
Before doing so, we use this notebook to test that merging PSI tables from separate runs will not introduce a large amount of untrustworthy splice events.

As part of tracking differences between the two tables, we have labeled NA values introduced at various steps of the data processing pipeline with different negative values:

-   -1: NA values assigned by Shiba (splice events with low coverage)

-   -2: NA values from merging the separate tables together (samples have a NA value after this step if there are genes dropped by Shiba for those samples because there is only one transcript detected in the GTF for that gene)

-   NA: NA values from merging the separate and combined tables.
    These NAs represent splice events that are only found in one analysis method (combined vs. separate)

We are particularly interested in the impact merging separate tables has on unannotated event detection, since this is why we chose to use Shiba over other tools.

**Targeted questions this notebook is trying to answer:**

-   What splice event types do we miss out on with the separate tables method?
-   Of splice events with high numbers of NA values, how many of these events only have a numeric PSI value in one (or a very low amount of) samples?
* How many NA values of different NA types are there in each type of splice table?
How are they distributed across events?
I am more concerned with the merge NAs (-2), as there is no good way to replace them in the separate tables (they could either be 0 or 1).
Another concern is events that are only found in the combined table, or only found in the separate table, as there is no easy way to correct for those.
* Can the number of NA values be reduced in the separate tables method if we filter for events with numeric PSI values in a minimum number of samples?
* What is the relationship between PSI values and NAs in each splice table type?
* What is the relationship between PSI values and NAs in each splice table type after filtering for events with numeric PSI values in a minimum number of samples?
Assuming the Shiba NAs (-1) are from low read support, do the separate and combined tables always call the same events as a Shiba NA? Merge NAs (-2) should arise from events dropped by Shiba due to lack of alternative transcripts for a gene, so we expect these NAs to be 0 or 1.
* How well correlated are the number of NA values in each event type for each PSI table type?
* How well correlated are the PSI values of events that are shared?

## Set up

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)
library(patchwork)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)

# color palette for splice event types
cbPalette <- c(
  afe = "#999999",
  ale = "#E69F00",
  five = "#56B4E9",
  mse = "#009E73",
  mxe = "#F0E442",
  ri = "#0072B2",
  se = "#D55E00",
  three = "#CC79A7"
)

# define minimum number of samples to filter splice table by
min_samples <- 5
```

### Define functions

```{r functions}

# PSI distribution plots of events in each method
psi_distributions <- function(
    df,
    event_type_value) {
  df |>
    dplyr::filter(event_type == event_type_value) |>
    ggplot(aes(PSI, fill = event_status)) +
    geom_histogram(bins = 20) +
    facet_wrap(vars(event_type, label, event_status), scales = "free_y") +
    plot_theme +
    # make facet labels bigger
    theme(strip.text.x = element_text(size = global_size),
          # rotate x axis labels so they don't overlap
          axis.text.x = element_text(angle =45))
}

# create summary table of events detected
make_event_summary <- function (events_table) { 
  events_table |>
    dplyr::summarise(
      .by = c(event_type, label),
      # count number of events in each event type and annotation category
      total = dplyr::n(),
      shared_count = sum(shared_event),
      combined_only_count = sum(combined_event) - shared_count,
      separate_only_count = sum(separate_event) - shared_count,
      shared_percent = shared_count / total * 100,
      combined_only_percent = combined_only_count / total * 100,
      separate_only_percent = separate_only_count / total * 100,
  )
}

# plot event summary frequencies as percent and raw counts bar plots
plot_event_summary <- function(
    summary_df) {
  # pivot summary df longer for plotting
  long_summary_df <- summary_df |>
    tidyr::pivot_longer(
      cols = c(
        shared_count,
        shared_percent,
        combined_only_count,
        combined_only_percent,
        separate_only_count,
        separate_only_percent
      ),
      # extract percents and counts into separate columns
      names_to = c("category", ".value"),
      names_pattern = "(.+)_(percent|count)"
    )

  # create percent stacked barplots
  percent_plot <-
    ggplot(long_summary_df, aes(fill = category, x = event_type, y = percent)) +
    geom_bar(position = "stack", stat = "identity") +
    plot_theme +
    labs(
      title = "% of splice events in combined, separate, or shared groups",
      x = "Annotation status of event",
      y = "Percentage of events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    plot_theme

  # create raw counts grouped barplot
  counts_plot <-
    ggplot(long_summary_df, aes(fill = category, x = event_type, y = count)) +
    geom_bar(position = "dodge", stat = "identity") +
    plot_theme +
    labs(
      title = "Number of splice events only in combined, separate, or shared groups",
      x = "Annotation status of event",,
      y = "Number of events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    plot_theme

  # arrange plots for printing
  percent_plot /
    counts_plot

}

# Histogram of quantified events
number_events_dist <- function(
    df,
    psi_type, title) {
    # construct histogram
    ggplot(df,  aes(x = .data[[psi_type]], fill = event_type)) +
    geom_histogram(binwidth = 1) +
    facet_wrap(vars(label)) +
    scale_fill_manual(values = cbPalette) +
    plot_theme +
    # make facet labels bigger
    theme(strip.text.x = element_text(size = global_size),
          # rotate x axis labels so they don't overlap
          axis.text.x = element_text(angle =45)) +
    ggtitle(title)
}

## Sample-level functions

# pivot event-level df long
pivot_long <- function (wide_event_df) {
  wide_event_df |>
    dplyr::select(starts_with("SRR"), event_type, pos_id, label) |>
    tidyr::pivot_longer(
      cols = matches("_PSI_(combined|separate)$"),
      names_to = c("sample", ".value"),
      names_pattern = "(.*)_PSI_(combined|separate)"
  )}

# event summary plots of frequency of events in each category
event_summary <- function(
    long_df,
    min_sample_value) {
  # filter samples by complete PSI value in min number of samples
  all_events_n_filtered <- long_df |>
    dplyr::group_by(pos_id) |>
    # count number of events are present in each shared status
    dplyr::mutate(
      combined_count = sum(method == "combined"),
      separate_count = sum(method == "separate")
    ) |>
    # filter for events with complete PSI values for minimum number of samples
    dplyr::filter(
      combined_count >= min_sample_value,
      separate_count >= min_sample_value
    ) |>
    dplyr::ungroup() |>
    # create summary table for plotting
    dplyr::summarise(
      .by = c(event_type, label),
      # count number of events in each event type and annotation category
      total = dplyr::n(),
      shared_count = sum(shared_event),
      combined_only_count = sum(combined_event) - shared_count,
      separate_only_count = sum(separate_event) - shared_count,
      shared_percent = shared_count / total * 100,
      combined_only_percent = combined_only_count / total * 100,
      separate_only_percent = separate_only_count / total * 100
    )
}

# Create summary table of number and percentage of events in each PSI-NA matchup category
na_comparison_summary <- function(long_df) {
  long_df |>
    # try to get rid of the NA NAs here
    tidyr::drop_na() |>
    dplyr::mutate(
      match_category =
        dplyr::case_when(
          separate == -1 & combined == -1 ~ "both Shiba NAs",
          separate == -2 & combined == -1 ~ "separate dropped, combined shiba NA",
          separate >= 0 & combined >= 0 ~ "both numeric PSIs",
          separate == -2 & combined >= 0 ~ "numeric PSI dropped in separate table"
        )
    ) |>
    dplyr::summarise(
      .by = c(match_category),
      count = dplyr::n()
    ) |>
    dplyr::mutate(
      total = sum(count),
      percent = count / total * 100
    )
}

# plot PSI distributions of events with numeric PSI values in combined tables but are dropped in separate tables
psi_values_lost_in_separate <- function(na_comparison_df) {
  na_comparison_df |>
    # drop NAs from putting together the separate and combined tables
    tidyr::drop_na() |>
    # assign match categories
    dplyr::mutate(
      match_category =
        dplyr::case_when(
          separate == -1 & combined == -1 ~ "both Shiba NAs",
          separate == -2 & combined == -1 ~ "separate dropped, combined shiba NA",
          separate >= 0 & combined >= 0 ~ "both numeric PSIs",
          separate == -2 & combined >= 0 ~ "numeric PSI dropped in separate table"
        )
    ) |>
    # plot only events in combined table with numeric PSIs that are dropped in the separate table
    dplyr::filter(match_category == "numeric PSI dropped in separate table",) |>
      ggplot(aes(combined, fill = event_type)) +
      geom_histogram(bins = 20) +
      facet_wrap(vars(event_type, label)) +
    scale_fill_manual(values = cbPalette) +
      plot_theme +
      # make facet labels bigger
      theme(strip.text.x = element_text(size = global_size),
            # rotate x axis labels so they don't overlap
            axis.text.x = element_text(angle =45))
}
```

## Directories and files

```{r}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# define the data directories
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merging-psi-tables")
target_pilot_dir <- file.path(merge_exploration_dir, "target_pilot")

# directory of shiba results produced from the same shiba run
combined_dir <- file.path(target_pilot_dir, "shiba_combined", "results")

# psi table directory
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# Directory of PSI table, merged from separate shiba runs
separate_psi_table_dir <- file.path(target_pilot_dir, "merged_results")

# merged PSI file of two samples obtained from merge-shiba-psi-tables.qmd
separate_psi_file <- file.path(separate_psi_table_dir, "merged_psi_table.tsv")

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of shiba results of two samples run together
shiba_psi_paths <-file.path(combined_splice_results_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

```

Read in files

```{r}
# read merged splice table from separate runs of Shiba on the TARGET pilot samples
separate_psi_table <- readr::read_tsv(separate_psi_file, col_types=readr::cols(.default = "c")) |>
  # replace remaining NAs from this table representing dropped genes from one sample with no detected splicing for that gene with -2
  dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -2)))

# read in combined splice results to compare against separate results
combined_splice_results <- shiba_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against separate PSI dataframe
combined_psi_table <- purrr::list_rbind(combined_splice_results, names_to = "event_type")
```

### Obtain dimensions of each PSI table

```{r}
# dimensions of separate PSI table
dim(separate_psi_table)
```

```{r}
# dimensions of combined PSI table
dim(combined_psi_table)
```

There are 71,811 more splice events in the separate PSI table than the combined table.

## What splice event types do we miss out on with the separate tables method?

To help us prioritize what splice event types we may be the most confident in after merging separate Shiba PSI tables, we are interested in seeing a breakdown of what event types are most represented in PSI tables constructed from different methods

```{r}
# create df of splice events that are shared between both combined and separate tables
# This is an event-level dataframe (each row is one unique pos_id/event_id)
# columns are each samples' PSI value (with -1/-2/NA values indicating different NA categories)
# combined_event, separate_event, and shared_event columns label what tables the event is present in
all_events <- dplyr::full_join(
  combined_psi_table,
  separate_psi_table,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_combined", "_separate")) |>
  # categorize events by whether they are in the combined or separate tables
  dplyr::mutate(
    # combined events are counted if the values are not all NA
    combined_event = ! dplyr::if_all(ends_with("_combined"), is.na),
    # separate events are counted if the values for the event are not all NA (indicating the specific position ID is only found in the combined or separate table)
    separate_event = ! dplyr::if_all(ends_with("_separate"), is.na),
    shared_event = combined_event & separate_event
  )

```

Print summary of counts across each method, across all event types

```{r}
# number of total events shared or present only in the combined and separate tables
# this table counts events with -1 and -2 PSI values

all_methods_summary <- all_events |> dplyr::summarise(
    # count number of events in each event type and annotation category
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    separate_only_count = sum(separate_event) - shared_count
)

all_methods_summary
```

Print summary of counts and percentages across each method, broken down by event type

```{r}
# event-level summary of the number of each splice event type in each splice table
all_events_summary <- make_event_summary(all_events)
all_events_summary
```

Plot event-level (one count per unique event) summary of event frequency across methods and event types

```{r}
#| label: fig-splice_events_breakdown
#| fig-width: 11
#| fig-height: 8

plot_event_summary(all_events_summary)
```
The majority of annotated event types are shared between both tables, although there are more splice events unique to the separate table (green) than the combined table (red).
Inflated values in the separate tables may be due to -2 NAs which arise in the Shiba processing of separate samples when when genes without multiple transcripts are dropped.

### What specific event types are impacted (missed) by the separate tables method?
These following plots answer the question for me better than the above summary plots.

```{r}
#| label: fig-combined_only_splice_events
#| fig-width: 11
#| fig-height: 8

combined_only_summary <- all_events_summary |>
  dplyr::select(event_type, label, combined_only_count, total, combined_only_percent)

long_summary_df <- combined_only_summary |>
    tidyr::pivot_longer(
      cols = c(
        combined_only_count,
        combined_only_percent
      ),
      # extract percents and counts into separate columns
      names_to = c("category", ".value"),
      names_pattern = "(.+)_(percent|count)"
    )

# create raw counts grouped barplot
counts_plot <- ggplot(long_summary_df, aes(fill = event_type, x = event_type, y = count)) +
    geom_bar(stat = "identity") +
      scale_fill_manual(values = cbPalette) +
    plot_theme +
    labs(
      title = "# annotated and unannotated splice events only in combined table",
      x = "Event Type",
      y = "Number of events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    theme(strip.text = element_text(size = 15))

# create percents grouped barplot
percents_plot <- ggplot(long_summary_df, aes(fill = event_type, x = event_type, y = percent)) +
    geom_bar(position = "dodge", stat = "identity") +
      scale_fill_manual(values = cbPalette) +
    plot_theme +
    labs(
      title = "% annotated and unannotated splice events only in combined table",
      x = "Annotation status of event",
      y = "Percent of all events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    theme(strip.text = element_text(size = 15))


percents_plot /
  counts_plot + plot_layout(guides = "collect")
```

I think of these as "percent of each event category missed in the separate tables".
In the worst cases, 30% of unannotated events might be only in the combined table (FIVE events).
Losing 30% of unannotated real events seems rough.

### How many events remain if we filter for events with PSI values in a minimum number of samples?

-   Of splice events with high numbers of NA values, how many of these events only have a numeric PSI value in one (or a very low amount of) samples?

Add summary columns to dataframe
```{r}
all_events <- all_events |>
  dplyr::mutate(
    ## count PSI value types in separate table ##
    # count the number of PSI values that are not one of our NA types
    n_quantified_separate = rowSums(
      # select sample PSI values that are not one of our NA types
      dplyr::pick(dplyr::ends_with("_PSI_separate")) >= 0, 
      # ignore NA values (events only in separate or combined tables)
      na.rm = TRUE),
    # count the number of PSI values dropped due to insufficient reads
    n_shiba_na_separate = rowSums(
      # select sample PSI values that have -1 NA type
      dplyr::pick(dplyr::ends_with("_PSI_separate")) == -1,
      na.rm = TRUE),
    # count number of PSI values dropped due to lack of transcript diversity in single samples
    n_dropped_separate = rowSums(
      # elect sample PSI values that have -2 NA type
      dplyr::pick(dplyr::ends_with("_PSI_separate")) == -2
    ),
    ## count PSI value types in combined table ##    
    # count the number of PSI values that are not one of our NA types
    n_quantified_combined = rowSums(
      # select sample PSI values that are not one of our NA types
      dplyr::pick(dplyr::ends_with("_PSI_combined")) >= 0, 
      # ignore NA values (events only in separate or combined tables)
      na.rm = TRUE),
    # count the number of PSI values dropped due to insufficient reads
    n_shiba_na_combined = rowSums(
      # select sample PSI values that have -1 NA type
      dplyr::pick(dplyr::ends_with("_PSI_combined")) == -1,
      na.rm = TRUE),
    # count number of PSI values dropped due to lack of transcript diversity in single samples
    n_dropped_combined = rowSums(
      # elect sample PSI values that have -2 NA type
      dplyr::pick(dplyr::ends_with("_PSI_combined")) == -2
    )
  )

# check what summary looks like
all_events |>
  dplyr::select(pos_id, n_quantified_separate, n_shiba_na_separate, n_dropped_separate, 
                n_quantified_combined, n_shiba_na_combined, n_dropped_combined) |>
  head()
```

Filter for minimum samples with numeric PSI values in the separate table and check n_quantified.
The resulting summaries of this filtered table will be incomplete, as it will not contain information of events in the combined table that are present in min_samples but are missed completely in the separate tables.
```{r}
# event-level summary of the number of each splice event type in each splice table
all_events_min_samples <- all_events |>
  # filter for events where there are numeric PSI values in at least 5 samples
  dplyr::filter(n_quantified_separate >= min_samples | n_quantified_combined >= min_samples)

# check what summary columns look like
all_events_min_samples  |>
  dplyr::select(pos_id, n_quantified_separate, n_shiba_na_separate, n_dropped_separate, 
                n_quantified_combined, n_shiba_na_combined, n_dropped_combined) |>
  dplyr::arrange((n_quantified_separate)) |>
  head()
```

#### Plot event distributions of the filtered separate table

```{r}
#| label: fig-splice_events_breakdown_5_samples_separate
#| fig-width: 11
#| fig-height: 8

# event-level summary of the number of each splice event type in min_sample filtered separate table
events_summary_min_separate_samples <- all_events_min_samples |>
  # this data frame still may include -1 and -2 NA values for other splice events but the sample-level analysis later will tell us more
  # filter for events where min_samples are in separate table only
  dplyr::filter(n_quantified_separate >= min_samples) |> 
  make_event_summary()

plot_event_summary(events_summary_min_separate_samples )
```

Print table of splice event counts of filtered combined table
```{r}
events_summary_min_separate_samples |>
  dplyr::select(event_type, label, shared_count, separate_only_count, combined_only_count)
```

Observations:
-   Although most events are shared after applying this filter and we still have unannotated events, 
we also still have events only found in the separate table.
These events are harder to deal with because they may be real events mislabeled by Shiba, or they may be unreal events (we cannot easily tell)

#### Plot event distributions of the filtered combined table
```{r}
#| label: fig-splice_events_breakdown_5_samples_combined
#| fig-width: 11
#| fig-height: 8

# event-level summary of the number of each splice event type in min_sample filtered combined table
events_summary_min_combined_samples <- all_events_min_samples |>
  # this data frame still may include -1 and -2 NA values for other splice events but the sample-level analysis later will tell us more
  # filter for events where min_samples are in separate table only
  dplyr::filter(n_quantified_combined >= min_samples) |> 
  make_event_summary()

plot_event_summary(events_summary_min_combined_samples)
```

Print table of splice event counts of filtered combined table
```{r}
events_summary_min_combined_samples |>
  dplyr::select(event_type, label, shared_count, separate_only_count, combined_only_count)
```

Observations:
* We can see that the filtered combined table results in more shared events that are unannotated, compared to the filtered separate table (by maybe 5x, looking at the Y axis scale).
* There are also a high number of combined_only events that are missed in the separate tables, even after filtering for combined events that are present in min_samples.
There are ~5,000 skipped exon events only in the combined table that are in min_samples that are missed in the separate tables.
That's more than the total number of skipped exon events remaining in the min_samples filtered separate table.

## How many of each splice event types are quantified?

Histogram of quantified events

```{r}
#| label: fig-num_quantified_psi
#| fig-width: 11
#| fig-height: 8

number_events_dist(all_events, "n_quantified_separate", "Number of quantified PSI values in separate table") /

number_events_dist(all_events, "n_quantified_combined", "Number of quantified PSI values in combined table") + plot_layout(guides = "collect")
```
Check skipped exon quantified PSI histograms
```{r}
#| label: fig-num_quantified_psi_se
#| fig-width: 11
#| fig-height: 8

all_events |>
  dplyr::filter(event_type == "se") |>
  number_events_dist("n_quantified_separate", "Number of quantified PSI values in separate table") /

all_events |>
  dplyr::filter(event_type == "se") |>
  number_events_dist("n_quantified_combined", "Number of quantified PSI values in combined table") + plot_layout(guides = "collect")
```

As expected, we have more observations of quantified unannotated PSI values in the combined table (higher bars at the end of the X axis). 
More critically, the unannotated observations in the combined table that we lose in the separate tables seem to be quantified in most of the TARGET pilot samples.

Histogram of events dropped by Shiba due to insufficient read counts

```{r}
#| label: fig-num_low_read_na
#| fig-width: 11
#| fig-height: 8

number_events_dist(all_events, "n_shiba_na_separate", "Number of PSIs dropped due to insufficient read counts in separate table") /

number_events_dist(all_events, "n_shiba_na_combined", "Number of PSIs dropped due to insufficient read counts in combined table") + plot_layout(guides = "collect")
```

These histograms are now counting events that are lost (number of -1 values).
In the histogram of PSI values dropped due to low read counts in the combined table, most events that are dropped are dropped in almost all the samples.
We see fewer events dropped due to low read counts in general in the separate tables, 
but this might be because the events that would have been dropped are missing from the separate are also missing due to the GTF (-2 NAs).

Check PSI values dropped from insufficient read histograms for skipped exons
```{r}
#| label: fig-num_low_read_na_se
#| fig-width: 11
#| fig-height: 8

all_events |>
  dplyr::filter(event_type == "se") |>
  number_events_dist("n_shiba_na_separate", "Number of PSIs dropped due to insufficient read counts in separate table") /

all_events |>
  dplyr::filter(event_type == "se") |>
  number_events_dist("n_shiba_na_combined", "Number of PSIs dropped due to insufficient read counts in separate table") + plot_layout(guides = "collect")
```

The same trend regarding unannotated events is seen here as in @fig-num_low_read_na.
We also see a higher number of annotated events dropped in all samples in the combined tables.
I suspect the reason for these two observations is because the combined tables have a larger dictionary of possible splice events due to the merged GTF.
So all these splice events not found in the separate tables have the potential to have -1 values.

Check histograms of events dropped in separate tables (-2 values).
These histograms plot the number of splice events dropped in splice tables made form the separate Shiba runs method.
These events should correspond to splice events that only have one transcript annotated in the sample's GTF and so are dropped by Shiba.

```{r}
#| label: fig-num_separate_table_na
#| fig-width: 11
#| fig-height: 8

number_events_dist(all_events, "n_dropped_separate", "# PSIs dropped due to insufficient transcripts in separate table")
  
# I do not plot this for the combined table because they are all 0 (we only define what is dropped in the separate tables based off what is missing when individual separate tables are merged)
```

Histogram of events missing in separate tables for skipped exon events

```{r}
#| label: fig-num_separate_table_na_se
#| fig-width: 11
#| fig-height: 8

all_events |>
  dplyr::filter(event_type == "se") |>
  number_events_dist("n_dropped_separate", "# PSIs dropped in separate table")
```

I mainly see that these missing values are present more in the unannotated events, where the event is only quantified in a small number of samples.
In both @fig-num_separate_table_na and @fig-num_separate_table_na_se, the events that are dropped are absent in most samples.