---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

As part of our pipeline, we plan to merge Shiba tables created for individual samples to obtain a final PSI table of all samples.
Before doing so, we use this notebook to test that merging PSI tables from separate runs will not introduce a large amount of untrustworthy splice events or junction counts.

Notes from code review/meeting
- Track different NA values separately.

NAs from the updated merging code (NAs in the separate table that was merged):
- -1: NA values from Shiba for the separate tables (may be due low coverage)
- -2: NA values from merging the separate tables (samples have a NA value after this step if there are genes dropped by Shiba for those samples because there is only one transcript detected for that gene)

NAs from the combined table
- -1: NA values from Shiba for the separate tables (may be due low coverage)

NAs from merging combined and separate tables
- NA: NAs coming from splice events that are only found in one analysis method (combined vs. separate)

## Set up

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)

# color palette for splice event types
cbPalette <- c(
afe = "#999999",
ale = "#E69F00",
five = "#56B4E9",
mse = "#009E73",
mxe = "#F0E442",
ri = "#0072B2",
se = "#D55E00",
three = "#CC79A7")
```

## Directories and files

```{r}

# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merging-psi-tables")

# define the data directories
# shiba results dir
shiba_dir <- file.path(merge_exploration_dir, "shiba_results")

# directory of shiba results produced from the same shiba run
combined_dir <- file.path(shiba_dir, "combined_run", "target_pilot", "results")

# psi table directory
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# Directory of PSI table, merged from separate shiba runs
separate_psi_table_dir <- file.path(merge_exploration_dir, "target_pilot", "merged_results")

# merged PSI file of two samples obtained from merge-shiba-psi-tables.qmd
separate_psi_file <- file.path(separate_psi_table_dir, "merged_psi_table.tsv")

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of shiba results of two samples run together
shiba_psi_paths <-file.path(combined_splice_results_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

```

Read in files

```{r}
# read merged splice table from separate runs of Shiba on the TARGET pilot samples
separate_psi_table <- readr::read_tsv(separate_psi_file) |>
  # replace remaining NAs from this table representing dropped genes from one sample with no detected splicing for that gene with -2
  dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -2)))

# read in combined splice results to compare against separate results
combined_splice_results_paths <- shiba_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against separate PSI dataframe
combined_psi_table <- purrr::list_rbind(combined_splice_results_paths, names_to = "event_type")
```
### Obtain dimensions of each PSI table

```{r}
# dimensions of separate PSI table
dim(separate_psi_table)
```

```{r}
# dimensions of combined PSI table
dim(combined_psi_table)
```

There are 71,811 more splice events in the separate PSI table than the combined table.

## Examine splice events that are consistently present in both combined and separate splice tables

To help us prioritize what splice event types we may be the most confident in after merging separate Shiba PSI tables, we are interested in seeing a breakdown of what event types are most represented in PSI tables constructed from different methods

```{r}
# obtain df of splice events that are shared between both combined and separate tables
all_events <- dplyr::full_join(
  combined_psi_table,
  separate_psi_table,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_combined", "_separate")) |>
  # categorize events by whether they are in the combined or separate tables
  dplyr::mutate(
    combined_event = ! dplyr::if_all(ends_with("_combined"), is.na),
    # separate events are counted if the values for the event are not all NA (indicating the splice event is only found in the combined or separate table)
    separate_event = ! dplyr::if_all(ends_with("_separate"), is.na),
    shared_event = combined_event & separate_event
  )
```

Print summary of percentage of each event types found in both methods, or in only one

```{r}
all_events_summary <- all_events |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    count = dplyr::n(),
    shared_percent = sum(shared_event) / count * 100,
    combined_only_percent = ( sum(combined_event) - sum(shared_event) ) / count * 100,
    separate_only_percent = ( sum(separate_event) - sum(shared_event) ) / count * 100,
)

all_events_summary
```

### Visualize breakdown of shared events with stacked bar plots

```{r}
#| label: fig-splice_events_breakdown
#| fig-width: 11
#| fig-height: 8

# pivot summary plot longer so we can make grouped bar plots of the percentages
long_all_events_summary <- all_events_summary |>
  tidyr::pivot_longer(
    cols = c(shared_percent, combined_only_percent, separate_only_percent),
    names_to = "category",
    values_to = "percent",
    # remove percent suffix for display
    names_pattern = "(.+)_percent"
  )

ggplot(long_all_events_summary, aes(fill = category, x = event_type, y = percent)) +
  geom_bar(position = "stack", stat = "identity") +
  plot_theme +
  labs(
    title = "% of splice events only in combined PSI tables",
    x = "Annotation status of event",
    y = "Percentage of events"
  ) +
  facet_wrap(vars(label)) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  plot_theme


```

The majority of annotated event types are shared between both tables, although there are more splice events unique to the separate table than combined table.
The capturing of unannotated event types suffers more from merging separate PSI tables.
In particular, there are more ALE and MXE events only found in the separate PSI tables.
For other unannotated events, there are 10-30% that are only found in the separate PSI.

### Obtain the number of splice events only found in each table

```{r}
# produce dataframe of elements in combined splice results reference dataframe
combined_only_list <- setdiff(combined_psi_table$pos_id, separate_psi_table$pos_id)

# produce dataframe of elements in merged splice results dataframe but not in the reference df
separate_only_list <- setdiff(separate_psi_table$pos_id, combined_psi_table$pos_id)

# events only present in separate tables
# Number of splice events only in separate PSI table
length(separate_only_list)

# Number of splice events only in combined PSI table
length(combined_only_list)
```


### Check how similar PSI values of shared events are

Pivot longer to create scatter plots

```{r}
# Pivot all events df longer
long_all_events <- all_events |>
  dplyr::slice_sample(n = 2000) |>
  tidyr::pivot_longer(
    cols = matches("_combined|_separate"),
    names_to = c("sample", "method"),
    values_to = "PSI",
    names_pattern = "(.*)_(combined|separate)"
  )

# convert negative PSI values to NA so they will be dropped in the scatterplots and correlation calculation
long_all_events_na <- long_all_events |>
  dplyr::mutate(
    PSI = dplyr::na_if(PSI, -1),
    PSI = dplyr::na_if(PSI, -2)
      )

# pivot wider for scatterplots

all_events_by_method <- long_all_events_na |> tidyr:: pivot_wider(
    names_from = method,
    values_from = PSI
  )
```

Make scatter plot of unannotated events

```{r}
#| label: fig-unannotated_psi_correlation
#| fig-width: 11
#| fig-height: 8

long_unannotated <- all_events_by_method |>
  dplyr::filter(label == "unannotated")

  ggplot(long_unannotated) +
    aes(
      # read in columns to use for x and y from input
      x = combined,
      y = separate,
    ) +
    geom_point(size = 0.5, alpha = 0.5) +
    labs(
      title = "Spearman correlation of unannotated shared splice events",
      x = "Combined run PSI values",
      y = "Separate run PSI values") +
    facet_wrap(vars(event_type)) +

    ggpubr::stat_cor(method = "spearman", label.x = 0.2, label.y = -0.1, color = "blue")
```

Make scatterplot of annotated events

```{r}
#| label: fig-annotated_psi_correlation
#| fig-width: 11
#| fig-height: 8

long_annotated <- all_events_by_method |>
  dplyr::filter(label == "annotated")


  ggplot(long_annotated) +
    aes(
      # read in columns to use for x and y from input
      x = combined,
      y = separate,
    ) +
    geom_point(size = 0.5, alpha = 0.5) +
    labs(
      title = "Spearman correlation of annotated shared splice events",
      x = "Combined run PSI values",
      y = "Separate run PSI values") +
    facet_wrap(vars(event_type)) +

    ggpubr::stat_cor(method = "spearman", label.x = 0.2, label.y = -0.1, color = "blue")
```

### Takeaways so far

-   In the worst case scenario, about 40% of some splice event types across all unique events found in both the combined and separate (merged) tables are only found in the separate tables.
This is concerning if we assume all events found only in the separate tables are false.

-   Most of the events found only in the separate tables are unannotated events, which are the event types we are hoping to capture most in the splice compendium

-   If we only look at splice events shared between both separate and combined PSI tables, the PSI values appear identical.
We may be able to be more confident in our splice table if we filter out everything with NA values (limit ourselves to PSI values of events found in all samples),
but this may severely shrink down the number of splice events in the compendium.

A next step for this notebook will be to compare the number of NA values in each table (I suspect there will be many more NAs in the separate tables)

Another note here is that we assume the splice events in the combined table are the ground truth, but a separate kind of evaluation would be needed to truly ask how many "real" splice events are in each of the tables.
For instance, we maybe would need to do this experiment on simulated reads where we know going in what all the transcripts are, which may be beyond the scope of this project.
