---
title: Merge TARGET Pilot Shiba PSI tables
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

Shiba is run one sample at a time using our snakemake workflow.
We then merge PSI tables produced for each sample.

## Set up

## Directories and files

```{r define_paths}
# load list of TARGET pilot accession IDs from config file of the run done with pilot samples together
samples_file <- file.path("experiment.tsv")

# find project root to access separate results dir
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# splice results directory of separate runs
group_dir <- file.path(repo_root, "results", "target", "shiba")

# shiba splice results path
splice_results_path <- "results/splicing"

# output directory and file of merged results
out_dir <- file.path("merged_results")
out_file <- file.path(out_dir, "merged_psi_table.tsv")

# create output directory if it does not exist
if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}
```

Construct paths to separate TARGET pilot shiba output files

```{r}
# read in experiment.tsv and create list of samples from samples column
samples <- readr::read_tsv(samples_file, col_types = list(.default = "c")) |>
  dplyr::pull(sample) |>
  as.list()

# make sample paths
sample_paths <- file.path(
  group_dir,
  samples,
  splice_results_path
)

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
psi_file_list <- c(
  se = "PSI_SE.txt.gz",
  afe = "PSI_AFE.txt.gz",
  ale = "PSI_ALE.txt.gz",
  five = "PSI_FIVE.txt.gz",
  three = "PSI_THREE.txt.gz",
  mse = "PSI_MSE.txt.gz",
  mxe = "PSI_MXE.txt.gz",
  ri = "PSI_RI.txt.gz"
)
```

## Define functions

```{r}
# Make function for reading all PSI tables for a sample (or set of samples run together)
read_psi <- function(psi_path, psi_files = psi_file_list) {
  # returns a data frame with the following columns: pos_id, gene_id, label, event_type, sample, psi
  # construct file path for event type
  file_paths <- file.path(psi_path, psi_files)
  # name each PSI table file path by event type
  names(file_paths) <- names(psi_files)

  # read in files
  purrr::map(file_paths, \(file) {
    readr::read_tsv(file, col_types = readr::cols(.default = "c")) |>
      # select for columns that will be used downstream in analysis so all dataframes have uniform column names
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # transform so we get a long table of PSI values for each event type, with one row per position and sample
      tidyr::pivot_longer(
        cols = contains("_PSI"),
        names_to = "sample",
        names_pattern = "(.+)_PSI",
        values_to = "psi",
        values_transform = as.numeric
      )
  }) |>
    # merge individual event type PSI tables to get one PSI table per sample
    dplyr::bind_rows(.id = "event_type")
}
```

## Read and combine PSI tables

Now we can read in all of the files and create a single long table with all samples.
We will also replace the `NA` values with `-1` to indicate NA PSI values assigned by Shiba (e.g., due to low coverage), so that we can distinguish these from `NA` values that will be introduced by the merge when a position is missing from a given sample's PSI table.

```{r read_in_files}
# read in all samples' PSI tables into a list of data frames, one per sample,
# then combine them into a single large table
sample_psi_long <- purrr::map(sample_paths, read_psi) |>
  # combine (sample info is already in the table from the
  # `_PSI` column names and the pivot_longer step in the read_psi functiont)
  purrr::bind_rows() |>
  # replace NA values with -1 to indicate NA PSI values assigned by shiba (e.g., due to low coverage)
  tidyr::replace_na(list(psi = -1))
```

To make a table with one row per event, we can now use `pivot_wider()`.
Any events sample combinations that were not assigned a value by Shiba will now be given an `NA` value so we can distinguish `NA` values from the merge from the ones that Shiba assigned, as the Shiba-assigned values will be `-1`.

```{r}
# Make a wide table of psi values
merged_psi <- sample_psi_long |>
  tidyr::pivot_wider(
    id_cols = c(pos_id, gene_id, label, event_type),
    names_from = sample,
    values_from = psi
  )
```

Save output as a table.

```{r}
readr::write_tsv(merged_psi, out_file)
```
