---
title: Check Merged vs. Combined TARGET Pilot Junction files
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  min_samples: 5
  chromosome: "chr1"
---

## Background

This notebook checks junctions.bed files from two Shiba run methods for differences.
The goal is to get a sense of how close files that go into calculating PSI values for events are to each other in the merged and combined Shiba methods.
Junctions files that are different would impact PSI calculation, so I am mainly looking to see if mismatched junctions are present and if present, how much of the data do they represent.

Files compared in this notebook are junction count bedfiles of 88 TARGET pilot samples, produced from running Shiba with two different methods.

"Combined" refers to shiba splice analysis run with shiba in a canonical way.
Inputs of this method are .bam files of each sample and a reference annotation GTF.
In the combined method, unmodified Shiba scripts are run on the input files using shiba.py, which calls on Shiba scripts that generate intermediate files and perform the final PSI calculation.
The intermediate files that go into PSI calculation from this method are:

- GTF, which is made with \[bam2gtf.py\](https://github.com/Sika-Zheng-Lab/Shiba/blob/v0.8.1/src/bam2gtf.py. This script uses a stringtie merge command as part of that merges GTFs made from individual samples together with the reference GTF

- junctions.bed, which is produced by [bam2junc.py](https://github.com/Sika-Zheng-Lab/Shiba/blob/v0.8.1/src/bam2junc.py).
  This script counts exon-exon (regtools) and exon-intron (featureCounts) junctions from each bam file given as input.

  - To count exon-intron junctions, Shiba uses `create_saf_file()` within `bam2junc.py` to create annotations of intron/exon boundaries to pass to featureCounts. The exon-intron boundaries are taken from the `ri_event` file generated from `gtf2events.py`.

  Then, `bam2junc.py` checks if exon-exon junctions are duplicated and deduplicates them.
  There is a bug in the deduplication step, as this notebook will later show junctions with invalid coordinates in the combined junctions file.
  Finally, junctions files containing exon-exon and exon-intron junctions from all samples are merged together.

"Merged" refers to shiba splice analysis run with `merge_results.smk`.
The inputs of this method are junctions.bed and GTF files created by Shiba run on each sample separately.
`merge_results.smk` merges the GTFs using the same stringtie merge command Shiba uses, then manually combines the junction files after deduplicating any invalid junction entries resulkting from the `bam2junc.py` bug.
The jobs for the merged Shiba method were run in the following order, to generate the merged GTF and junctions.bed files:

```
[Mon Jun  1 03:46:24 2026]
Finished jobid: 4 (Rule: merge_junctions)
1 of 5 steps (20%) done
[Mon Jun  1 04:07:03 2026]
Finished jobid: 3 (Rule: merge_gtfs)
2 of 5 steps (40%) done
Select jobs to execute...
Execute 1 jobs...
[Mon Jun  1 04:07:03 2026]
localrule gtf_to_events:
    input: results/merged_shiba/target_pilot/merged_gtf.gtf, references/gencode.v47.primary_assembly.annotation.gtf
    output: results/merged_shiba/target_pilot/events
    jobid: 2
    reason: Missing output files: results/merged_shiba/target_pilot/events; Input files updated by another job: results/merged_shiba/target_pilot/merged_gtf.gtf
    priority: 1
    threads: 10
    resources: tmpdir=/tmp
[Mon Jun  1 04:43:48 2026]
Finished jobid: 2 (Rule: gtf_to_events)
```
Because the junctions.bed files have exon-intron coordinates defined by retained intron events files obtained from one sample only, some samples will not have exon-intron counts from exon-intron boundaries that are only present in other samples.

I'm not sure if it 100% makes sense to me to try to count exon-intron junctions in sample A that are only annotated in sample B.
I don't understand how we would know the sample B junction counts in sample A are real, if sample A does not have those intron-exon boundaries in its GTF.

### Analysis outline:

Check differences in full junctions files - Check how many rows (junction IDs) are in each bedfile.
Also calculate fraction of junctions in the merged junctions table that are not in the combined table; the fraction of junctions in the combined table that are missed in the merged junctions table

Check differences in junctions with at least 10 junctions in min_samples - Check if differences change if we filter for junctions with higher counts in at least 5 samples.
Filter merged and combined junctions tables for junctions with counts of at least 10 in at least 5 samples.
Calculate fraction of junctions in the merged junctions table that are not in the combined table; the fraction of junctions in the combined table that are missed in the merged junctions table

## Read in directories and files

Define directories and file paths

```{r directories_and_files}
### Directories ###

# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## combined shiba run results on target pilot samples ##
exploration_dir <- file.path(repo_root, "exploration")

# merged table eval dir
exploration_eval_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")

# shiba results dir
# target pilot exploration dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# junctions dir
combined_junctions_dir <- file.path(combined_results_dir, "junctions")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")

### Files ###
## combined shiba target pilot files
target_pilot_sample_sheet <- file.path(target_pilot_dir, "experiment.tsv")
# combined junctions bedfile from combined Shiba method with known invalid junctions
combined_junctions_file <- file.path(combined_junctions_dir, "junctions.bed")
# deduplicated combined junctions file - made by running bash scripts/deduplicated_shiba_junctions.sh
deduplicated_combined_junctions_file <- file.path(combined_junctions_dir, "deduplicated_junctions.bed")

## merged shiba target pilot files
merged_junctions_file <- file.path(merged_results_dir, "merged_junctions.bed")

```

Read in files

```{r read_in_files}
# list of target pilot sample IDs
target_pilot_samples <- readr::read_tsv(target_pilot_sample_sheet) |>
  dplyr::pull(sample)

# read in combined junctions table that have been deduplicated
deduplicated_combined_junctions <- readr::read_tsv(
  deduplicated_combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in files generated from merged shiba run
merged_junctions <- readr::read_tsv(
  merged_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

```

## Check tables for parsing errors

### Print parsing errors in combined junctions table

```{r check_parsing_errors}
combined_junctions_dat <- vroom::vroom(combined_junctions_file)
combined_junctions_problems <- vroom::problems(combined_junctions_dat)
combined_junctions_problems
```

We have `r nrow(unique(combined_junctions_problems$row))` junctions in the combined junctions file that are impacted by the concatenation bug in Shiba.
PSI files for events involving these junctions may be problematic in the combined PSI file.

### Check for parsing errors in the merged and deduplicated junctions tables

```{r}
merged_junctions_dat <- vroom::vroom(merged_junctions_file)
```

```{r}
deduplicated_comb_junctions_dat <- vroom::vroom(deduplicated_combined_junctions_file)
deduplicated_comb_junctions_problems <- vroom::problems(deduplicated_combined_junctions)
```

No problems are present in merged_junctions file or deduplicated junctions file

## Check differences in junctions bedfiles

### Subset junctions files
This portion is not 100% modular as I can't easily supply a field in params$chromosome that will allow all chromosomes to be passed in (unless I really do give it the full list of chromosomes)
Maybe add in an if statement to only do this filter if there is a value in params$chromosome ?

```{r}
deduplicated_combined_junctions <- deduplicated_combined_junctions |>
  dplyr::filter(chr %in% params$chromosome) |>
  dplyr::relocate(all_of(target_pilot_sample_sheet))

merged_junctions <- merged_junctions |>
  dplyr::filter(chr %in% params$chromosome) |>
  dplyr::relocate(all_of(target_pilot_sample_sheet))

```

### Check differences in junctions files

```{r count_differences}

# print number of junctions in only in combined junctions bedfile
combined_only <- dplyr::setdiff(deduplicated_combined_junctions, merged_junctions)
nrow(combined_only)

# check fraction of junctions that are only in the combined junctions bedfile
frac_combined_only <- nrow(combined_only) / nrow(deduplicated_combined_junctions)
frac_combined_only

# print number of junctions only in merged junctions bedfile
merged_only <- dplyr::setdiff(merged_junctions, deduplicated_combined_junctions)
nrow(merged_only)

# check fraction of junctions that are only in merged junctions
frac_merged_only <- nrow(merged_only) / nrow(merged_junctions)
frac_merged_only

```

There are `r nrow(merged_junctions) - nrow(deduplicated_combined_junctions)` more junctions in the merged junction file compared to the combined junction file.

`r frac_combined_only * 100`% of the combined method junctions are missed in the merged junctions file.

`r frac_merged_only * 100`% of the merged method junctions are not in the combined junctions file.

### Check differences in junctions with at least 10 junctions in min_samples

```{r filter_junction_counts_merged}
filtered_merged_junctions <- merged_junctions |> dplyr::filter(
  # count number of columns per row with junction count of at least 10
  rowSums(
    dplyr::across(
      dplyr::all_of(target_pilot_samples), ~ . >= 10
    )
    # filter for rows where at least 5 columns (samples) have counts of at least 10
  ) >= params$min_samples
  )

filtered_merged_junctions |>
  dplyr::arrange(start) |>
  head()
```

```{r filter_junction_counts_combined}

filtered_combined_junctions <- deduplicated_combined_junctions |> dplyr::filter(
  # count number of columns per row with junction count of at least 10
  rowSums(
    dplyr::across(
      dplyr::all_of(target_pilot_samples), ~ . >= 10
    )
    # filter for rows where at least 5 columns (samples) have counts of at least 10
  ) >= params$min_samples
  )

filtered_combined_junctions |>
  dplyr::arrange(start) |>
  head()
```

Count differences between counts-filtered junction files

Count number of junctions in each filtered file

```{r count_differences_counts_filtered}
nrow(filtered_combined_junctions)
nrow(filtered_merged_junctions)
```

Calculate fraction of filtered junctions that are only in the combined or merged files

```{r count_differences_filtered}

# print number of junctions only in combined junctions bedfile after filtering for junctions with at least 10 counts in at least 5 samples
filtered_combined_only <- dplyr::setdiff(filtered_combined_junctions, filtered_merged_junctions)
nrow(filtered_combined_only)

# check fraction of filtered_combined_only junctions
nrow(filtered_combined_only) / nrow(deduplicated_combined_junctions)

# print number of junctions only in merged junctions bedfile after filtering for junctions with at least 10 counts in at least 5 samples
filtered_merged_only <- dplyr::setdiff(filtered_merged_junctions, filtered_combined_junctions)
nrow(filtered_merged_only)

# check fraction of junctions different in the merged table > 10 counts
nrow(filtered_merged_only) / nrow(merged_junctions)
```

### Print examples of junctions only in merged or combined tables

#### Check junctions only in combined junctions file

Print examples of junctions only in the combined junctions file

```{r junctions_missed_in_merged}
filtered_combined_only |>
  dplyr::arrange(start) |>
  head()
```

For comparison, examine merged chromosome 1 counts-filtered junction table for junctions near chr1:18369-18370

```{r}
filtered_merged_junctions |>
  dplyr::arrange(start) |>
  dplyr::filter((18369 -  10) < start &
                  start < (18369 + 10) ) |>
  head()
```

This junction is completely missed, with the closest coordinate being chr1:18369-18913.

#### Check junctions only in merged junctions file

Print examples of junctions only in the merged junctions file

```{r junctions_only_in_merged}
filtered_merged_only |>
  dplyr::arrange(start) |>
  head()
```

For comparison, examine combined counts-filtered junction table for junctions near chr1:268020-268021

```{r}
filtered_combined_junctions |>
  dplyr::arrange(start) |>
  dplyr::filter((268020 -  10) < start &
                  start < (268020 + 10) ) |>
  head()
```

Position chr1:268020-268021 is actually a junction recorded in the combined junctions table.
The difference come in the counts going into each sample:

| sample | junction count in merged table | junction count in combined table |
|----|----|----|
| **SRR1712457** | 0 | 3 |
| **SRR1712459** | 0 | 3 |
| **SRR1712460** | 0 | 8 |
| **SRR1712461** | 0 | 3 |
| **SRR1712462** | 0 | 4 |
| ... | ... | ... |
| **SRR1799041** | 0 | 16 |

From scanning chr1:268020-268021, the counts differences arise from cases where the merged junctions counts have a value of 0 for a sample and the combined file is not zero for the same sample.
For all cases except for SRR1799041, the counts that are dropped in the merged table are < 10.

Because the values in the merged table are 0 for the counts that differ from the combined table, I suspect these differences stem from junctions that are not shared across samples (these appear as NA values which we replace with 0 in the final merged bedfile).
In the merged method, these 0 counts arise because exon-intron boundaries are defined from an intron events coordiantes file from gtf2event.py that was constructed from the annotations of one sample only.
