---
title: Check Merged vs. Combined TARGET Pilot Junction files
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  min_counts: 0
  chromosome: "chr1"
output: md_document
---

## Background

This notebook checks junctions.bed files from two Shiba run methods for differences.
The goal is to get a sense of how close files that go into calculating PSI values for events are to each other in the merged and combined Shiba methods.
Junctions files that are different would impact PSI calculation, so I am mainly looking to see if mismatched junctions are present and if present, how much of the data do they represent.

Files compared in this notebook are junction count bedfiles of 88 TARGET pilot samples, produced from running Shiba with two different methods.

"Combined" refers to shiba splice analysis run with shiba in a canonical way.
Inputs of this method are .bam files of each sample and a reference annotation GTF.
In the combined method, unmodified Shiba scripts are run on the input files using shiba.py, which calls on Shiba scripts that generate intermediate files and perform the final PSI calculation.
The intermediate files that go into PSI calculation from this method are:

- GTF, which is made with \[bam2gtf.py\](https://github.com/Sika-Zheng-Lab/Shiba/blob/v0.8.1/src/bam2gtf.py. This script uses a stringtie merge command as part of that merges GTFs made from individual samples together with the reference GTF

- junctions.bed, which is produced by [bam2junc.py](https://github.com/Sika-Zheng-Lab/Shiba/blob/v0.8.1/src/bam2junc.py).
  This script counts exon-exon (regtools) and exon-intron (featureCounts) junctions from each bam file given as input.

  - To count exon-intron junctions, Shiba uses `create_saf_file()` within `bam2junc.py` to create annotations of intron/exon boundaries to pass to featureCounts. The exon-intron boundaries are taken from the `ri_event` file generated from `gtf2events.py`.

  Then, `bam2junc.py` checks if exon-exon junctions are duplicated and deduplicates them.
  There is a bug in the deduplication step, as this notebook will later show junctions with invalid coordinates in the combined junctions file.
  Finally, junctions files containing exon-exon and exon-intron junctions from all samples are merged together.

"Merged" refers to shiba splice analysis run with `merge_results.smk`.
The inputs of this method are junctions.bed and GTF files created by Shiba run on each sample separately.
`merge_results.smk` merges the GTFs using the same stringtie merge command Shiba uses, then manually combines the junction files after deduplicating any invalid junction entries resulting from the `bam2junc.py` bug.
The jobs for the merged Shiba method were run in the following order, to generate the merged GTF and junctions.bed files:

```         
[Mon Jun  1 03:46:24 2026]
Finished jobid: 4 (Rule: merge_junctions)
1 of 5 steps (20%) done
[Mon Jun  1 04:07:03 2026]
Finished jobid: 3 (Rule: merge_gtfs)
2 of 5 steps (40%) done
Select jobs to execute...
Execute 1 jobs...
[Mon Jun  1 04:07:03 2026]
localrule gtf_to_events:
    input: results/merged_shiba/target_pilot/merged_gtf.gtf, references/gencode.v47.primary_assembly.annotation.gtf
    output: results/merged_shiba/target_pilot/events
    jobid: 2
    reason: Missing output files: results/merged_shiba/target_pilot/events; Input files updated by another job: results/merged_shiba/target_pilot/merged_gtf.gtf
    priority: 1
    threads: 10
    resources: tmpdir=/tmp
[Mon Jun  1 04:43:48 2026]
Finished jobid: 2 (Rule: gtf_to_events)
```

Because the junctions.bed files have exon-intron coordinates defined by retained intron events files obtained from one sample only, some samples will not have exon-intron counts from exon-intron boundaries that are only present in other samples.

## Analysis outline:

We discovered that the largest source of difference between the merged and combined method lies in how exon-intron junctions are defined and counted.
Shiba uses the retained intron events coordinates generated from gtf2events.py to define intron-exon coordinates for counting.
Since the merged method only uses the events coordinates defined from one sample at a time, junctions counted with this method are expected to miss all exon-intron junctions that are not shared between samples.
Coming into this analysis, I expect all junction IDs that are different between the methods to be exon-intron junctions.

Analysis questions in this notebook:

1.  How many junction positions (IDs) are different between bedfiles from each analysis method?

    1.  How many and what fraction of all IDs in each method are different?

    2.  Of IDs that are different, how many samples have nonzero junction counts in each analysis method?

2.  Of junction positions that are the same, how many IDs have different junction counts in the same samples in each analysis method?

    1.  How many and what fraction of all IDs have samples with different junction counts?

    2.  How many samples in each method have different junction counts for the same IDs?

    3.  Are differences in junction counts always due to counts being 0 in the merged method (due to us replacing junctions not shared between samples with 0)?

    4.  Are there samples that consistently have different junction counts in each method?

3.  What kind of junction IDs are different between bedfiles from each analysis method?

    1.  Are junction IDs that are only found in one analysis always 1bp in length (exon-intron)?

    2.  Are shared junction IDs with different counts in each analysis always 1bp in length?

## Read in directories and files

Define directories and file paths

```{r directories_and_files}
### Directories ###

# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## combined shiba run results on target pilot samples ##
exploration_dir <- file.path(repo_root, "exploration")

# merged table eval dir
exploration_eval_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")

# shiba results dir
# target pilot exploration dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# junctions dir
combined_junctions_dir <- file.path(combined_results_dir, "junctions")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")

### Files ###
## combined shiba target pilot files
target_pilot_sample_sheet <- file.path(target_pilot_dir, "experiment.tsv")
# deduplicated combined junctions file - made by running bash scripts/deduplicated_shiba_junctions.sh
deduplicated_combined_junctions_file <- file.path(combined_junctions_dir, "deduplicated_junctions.bed")

## merged shiba target pilot files
merged_junctions_file <- file.path(merged_results_dir, "merged_junctions.bed")

```

Read in files

```{r read_in_files}
# list of target pilot sample IDs
target_pilot_samples <- readr::read_tsv(target_pilot_sample_sheet) |>
  dplyr::pull(sample)

# read in combined junctions table that have been deduplicated
deduplicated_combined_junctions <- readr::read_tsv(
  deduplicated_combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in files generated from merged shiba run
merged_junctions <- readr::read_tsv(
  merged_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

```
## Define functions
```{r}
# pivot junctions dataframe long and add summary column of number of samples with counts exceeding min_counts
pivot_junctions_long <- function(df) {
 df |>
  tidyr::pivot_longer(
    cols = dplyr::starts_with("SRR"),
    names_to = "sample",
    values_to = "count"
  )
}

# create histogram of the number of samples in each junction ID that has a count over 0
num_nonzero_samples_hist <- function(df, id_vector) {
  summary_df <- df |>
  # filter for IDs only present in combined table
  dplyr::filter(ID %in% id_vector) |>
  # pivot longer for ease of filtering by samples
  pivot_junctions_long() |>
  # summarize by how many samples in each junction ID have counts over 0
  dplyr::summarize(
    .by = ID,
    total = dplyr::n(),
    n_counts_over_zero = sum(count > 0) 
  )
  
  # create histogram to spot-check
  hist(summary_df$n_counts_over_zero,
       main = paste("Distribution of samples with counts > 0 per junction ID"),
       xlab = "Number of samples with counts > 0 for junction IDs")
}

# print the top 6 samples with the most zero counts in a given set of junctions
samples_with_most_zeros <- function(df, id_vector) {
  # filter dataframe of junction IDs only present in the combined method
  df |>
    dplyr::filter(ID %in% id_vector) |>
    pivot_junctions_long() |>
    # add a column that labels if the sample has a count of 0
    dplyr::mutate(is_zero = dplyr::if_else(count == 0, TRUE, FALSE)) |>
    # filter for samples with a count of zero for each junction
    dplyr::filter(is_zero == TRUE) |>
    # for each sample, count how many times it has zero junction counts
    dplyr::summarize(
      .by = sample,
      n_zero_junction_counts = dplyr::n()
    ) |>
    # arrange table so that sample with most zero counts is printed first, then print head
    dplyr::arrange(desc(n_zero_junction_counts)) |>
    head()
  }
```

## Check tables for parsing errors

### Check for parsing errors in the merged and deduplicated junctions tables

```{r check_problems}
readr::problems(merged_junctions)
readr::problems(deduplicated_combined_junctions)
```

No problems are present in merged_junctions file or deduplicated junctions file (readr::problems prints out nothing)

## Subset junctions files for analysis

```{r subset_junctions}
# Subset junctions files if a list of chromosomes are provided, or if one chromosome is provided in params

if(length(params$chromosome) > 1 || tolower(params$chromosome) != "all"){
  deduplicated_combined_junctions <- deduplicated_combined_junctions |>
    dplyr::filter(chr %in% params$chromosome)
}

if(length(params$chromosome) > 1 || tolower(params$chromosome) != "all"){
  merged_junctions <- merged_junctions |>
    dplyr::filter(chr %in% params$chromosome)
}

# Ensure sample columns are in the same order
merged_junctions <- merged_junctions |> 
  dplyr::relocate(chr, start, end, ID, dplyr::all_of(target_pilot_samples))

# Check that columns are the same between both dataframes
setequal(colnames(deduplicated_combined_junctions), colnames(merged_junctions))
```

## Check differences in junctions files

### How many junction positions (IDs) are different between bedfiles from each analysis method?

#### How many and what fraction of all IDs in each method are different?

```{r count_diff_ids}
# save vectors of junction IDs in each junction file being compared
combined_ids <- deduplicated_combined_junctions$ID
merged_ids <- merged_junctions$ID

# Obtain vector of junction IDs only in combined junctions bedfile
combined_only_ids <- setdiff(combined_ids, merged_ids)
# Save number of junctions only in the combined dataframe
num_combined_only <- length(combined_only_ids)

# Calculate fraction of junction IDs in the combined method that are only found in the combined method
frac_combined_only <- num_combined_only / length(combined_ids)

# Obtain vector of junction IDs only in combined junctions bedfile
merged_only_ids <- setdiff(merged_ids, combined_ids)
# Save number of junctions only in the merged dataframe
num_merged_only <- length(merged_only_ids)

# Calculate fraction of junction IDs in the merged method that are only found in the merged method
frac_merged_only <- num_merged_only / length(merged_ids)

# print results
summary_message <- c(
  paste0("Number of rows different in combined junctions: ", num_combined_only),
  paste0("Fraction of rows different in combined junctions: ", frac_combined_only),
  paste0("Number of rows different in merged junctions: ", num_merged_only),
  paste0("Fraction of rows different in merged junctions: ", frac_merged_only)
  )

summary_message
```
`r frac_combined_only * 100`% of the combined method junction IDs are missed in the merged junctions file.

`r frac_merged_only * 100`% of the merged method junction IDs are not in the combined junctions file.

#### Of IDs that are different, how many samples have nonzero junction counts in each analysis method?

##### Junction IDs only present in combined method

```{r}
#| label: fig-samples_with_nonzero_counts_combined_only_ids

num_nonzero_samples_hist(deduplicated_combined_junctions, combined_only_ids)
```
Most (more than half) of the junction IDs that are only found in the combined junctions table have nonzero counts in over half of the samples.

What samples have the highest number of nonzero junction counts?

```{r what_samples_have_zero_counts_combined_only_junction_ids}
samples_with_most_zeros(deduplicated_combined_junctions, combined_only_ids)
```

SRR2042845 has the highest number of zero counts in junction IDs only found in the combined method

##### Junction IDs only present in merged method

```{r}
#| label: fig-samples_with_nonzero_counts_merged_only_ids

num_nonzero_samples_hist(merged_junctions, merged_only_ids)
```
In IDs only found in the merged junctions table, most samples have count values of 0

What samples have the highest number of nonzero junction counts?

```{r what_samples_have_zero_counts_merged_only_junction_ids}
samples_with_most_zeros(merged_junctions, merged_only_ids)
```
SRR2042845 still has the highest number of junction counts that are 0 in junction IDs only found in the merged method.
So there are some samples that consistently have unusually low numbers of counts.

### Of junction positions that are the same, how many IDs have different junction counts in the same samples in each analysis method?

```{r}

```


### What kind of junction IDs are different between bedfiles from each analysis method?

Check if junctions only found in combined or merged dataframes are always 1bp in length (are they always exon-intron junctions?)

```{r check_1bp_junctions}
filtered_combined_only_ids <- setdiff(filtered_long_combined$ID, filtered_long_merged$ID)
long_combined_only_ids <- filtered_long_combined |>
  dplyr::filter(ID %in% filtered_combined_only_ids) |>
  # count length of junctions only found in combined table
  dplyr::mutate(junction_length = abs(end - start))

# check if the junctions only in combined df are all always 1bp in length (are they all exon-intron junctions?)
unique(long_combined_only_ids$junction_length)

filtered_merged_only_ids <- setdiff(filtered_long_merged$ID, filtered_long_combined$ID)
long_merged_only_ids <- filtered_long_merged |>
  dplyr::filter(ID %in% filtered_merged_only_ids) |>
  # count length of junctions only found in combined table
  dplyr::mutate(junction_length = abs(end - start))

# check if the junctions only in merged df are all always 1bp in length
unique(long_combined_only_ids$junction_length)

```

Junctions that are only found in either the combined or merged dataframes are always 1bp in length, supporting our hypothesis that the difference stems from exon-intron junctions

### Print examples of junctions only in merged or combined tables

#### Check junctions only in combined junctions file

Print examples of junctions only in the combined junctions file

```{r junctions_missed_in_merged}
filtered_combined_only |>
  dplyr::arrange(start) |>
  head()
```

For comparison, examine merged chromosome 1 counts-filtered junction table for junctions near chr1:18369-18370

```{r}
merged_junctions |>
  dplyr::arrange(start) |>
  dplyr::filter(
    # make sure position filtered is chromosome 1 even if we analyze full bedfile
    chr == "chr1",
    (18369 -  10) < start &
      start < (18369 + 10) )
```

This junction is completely missed, with the closest coordinate being chr1:18369-18913.

Check if chr1:18369-18913 is only in the merged table and not in the combined table

```{r}
"chr1:18369-18913" %in% merged_only_ids
```

chr1:18369-18913 is found in both the merged and combined junctions tables.

#### Check junctions only in merged junctions file

Print examples of junctions only in the merged junctions file

```{r junctions_only_in_merged}
filtered_merged_only |>
  dplyr::arrange(start) |>
  head()
```

For comparison, examine combined junction table for junctions near chr1:268020-268021

```{r}
deduplicated_combined_junctions |>
  dplyr::arrange(start) |>
  dplyr::filter(
    chr == "chr1",
    (268020 -  10) < start &
    start < (268020 + 10) ) |>
  head()
```

Position chr1:268020-268021 is actually a junction recorded in the combined junctions table.
The difference come in the counts going into each sample:

| sample | junction count in merged table | junction count in combined table |
|------------------------|------------------------|------------------------|
| **SRR1712457** | 0 | 3 |
| **SRR1712459** | 0 | 3 |
| **SRR1712460** | 0 | 8 |
| **SRR1712461** | 0 | 3 |
| **SRR1712462** | 0 | 4 |
| ... | ... | ... |
| **SRR1799041** | 0 | 16 |

From scanning chr1:268020-268021, the counts differences arise from cases where the merged junctions counts have a value of 0 for a sample and the combined file is not zero for the same sample.
For all cases except for SRR1799041, the counts that are dropped in the merged table are \< 10.

Because the values in the merged table are 0 for the counts that differ from the combined table, I suspect these differences stem from junctions that are not shared across samples (these appear as NA values which we replace with 0 in the final merged bedfile).
In the merged method, these 0 counts arise because exon-intron boundaries are defined from an intron events coordinates file from gtf2event.py that was constructed from the annotations of one sample only.

## Check if differences flagged in junction IDs that are identical stem from differences in 0 counts

```{r are_same_id_differences_from_zeros}
# Exclude IDs that are only found in each dataframe
# The "_shared" dataframe contain rows corresponding to junctions that are shared between the merged and combined methods
# The differences within these dataframes should come from the values of the samples' junction counts for these IDs

merged_junctions_shared <- merged_junctions |>
  dplyr::filter(!ID %in% merged_only_ids)

combined_junctions_shared <- deduplicated_combined_junctions |>
  dplyr::filter(!ID %in% combined_only_ids)

# check that resulting length of dataframes excluding differing IDs are the same (they should be)
nrow(merged_junctions_shared) == nrow(combined_junctions_shared)

# pull out dataframe of junction IDs with values in merged_junctions_shared not in combined_junctions_shared
merged_shared_id_differences <- dplyr::setdiff(merged_junctions_shared, combined_junctions_shared)

combined_shared_id_differences <- combined_junctions_shared |> dplyr::filter(
  ID %in% merged_shared_id_differences$ID
)

```

Examine junction counts differences in one sample only.
Check if there are any junctions that are different because the junction counts of the merged and combined dataframes are nonzero and different

```{r}
# Pull one sample ID to examine
# This should be in params but I think ideally I would want to do this for all samples
# Not sure of the best way to do that at the moment - maybe do this on the long dataframe?

sample <- "SRR1559043"

# select for only this sample in the combined and merged dataframes
merged_shared_id_differences <- merged_shared_id_differences |>
  dplyr::select(chr, start, end, ID, dplyr::all_of(sample))
combined_shared_id_differences <- combined_shared_id_differences |>
  dplyr::select(chr, start, end, ID, dplyr::all_of(sample))

# Combine the dataframes and label each row by whether the junction count for the example sample are zero-to-nonzero, the same, or different due to nonzero-to-nonzero reasons

compare_junction_values_df <- dplyr::left_join(
  merged_shared_id_differences,
  combined_shared_id_differences,
  by = c("chr", "start", "end", "ID"),
  suffix = c("_merged", "_combined")
) |>
  # label junctions by nature of difference
  dplyr::mutate(
    difference_type = dplyr::case_when(SRR1559043_merged == SRR1559043_combined ~ "same",
                                       SRR1559043_merged != 0 & SRR1559043_combined != 0 ~ "nonzero-to-nonzero",
                                       SRR1559043_merged == 0 & SRR1559043_combined != 0 ~ "zero-to-nonzero"
    )
  )

# spot-check
compare_junction_values_df |> head()
```

Summarize fraction of junctions in sample `r target_pilot_samples[1]` in each category

```{r}
compare_junction_values_df |>
  dplyr::summarise(
    total = dplyr::n(),
    n_same = sum(difference_type == "same"),
    n_zero_to_nonzero = sum(difference_type == "zero-to-nonzero"),
    n_nonzero_to_nonzero = sum(difference_type == "nonzero-to-nonzeo"),
    frac_same = n_same / total,
    frac_zero_to_nonzero = n_zero_to_nonzero / total,
    frac_nonzero_to_nonzero = n_nonzero_to_nonzero / total
  )
```
