---
title: Check Merge vs. Combined TARGET Pilot Junction files (chromosome 1)
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Read in directories and files

Define directories and file paths

```{r directories_and_files}
### Directories ###

# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## combined shiba run results on target pilot samples ##
exploration_dir <- file.path(repo_root, "exploration")

# merged table eval dir
exploration_eval_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")

# shiba results dir
# target pilot explroation dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# junctions dir
combined_junctions_dir <- file.path(combined_results_dir, "junctions")
# gtf dir
combined_gtf_dir <- file.path(combined_results_dir, "annotation")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")

### Files ###
## combined shiba target pilot files
target_pilot_sample_sheet <- file.path(target_pilot_dir, "experiment.tsv")
# combined target pilot outputs
combined_gtf_file <- file.path(combined_gtf_dir, "assembled_annotation.gtf")
combined_junctions_file <- file.path(combined_junctions_dir, "junctions.bed")
# deduplicated combined junctions file - made by running bash scripts/deduplicated_shiba_junctions.sh
deduplicated_combined_junctions_file <- file.path(combined_junctions_dir, "deduplicated_junctions.bed")

## merged shiba target pilot files
merged_gtf_file <- file.path(merged_results_dir, "merged_gtf.gtf")
merged_junctions_file <- file.path(merged_results_dir, "merged_junctions.bed")

## output files
only_in_comb_junc <- file.path(exploration_eval_dir, "jcn_only_in_combined.bed")
only_in_merged_junc <- file.path(exploration_eval_dir, "jcn_only_in_merged.bed")

```

Create output dir if it does not exist

```{r create_dirs}
# Check if it exists; if not, create it
if (!dir.exists(exploration_eval_dir)) {
  dir.create(exploration_eval_dir, recursive = TRUE)
}
```

Read in files

```{r read_in_files}
# list of target pilot sample IDs
target_pilot_samples <- readr::read_tsv(target_pilot_sample_sheet) |>
  dplyr::pull(sample)

# minimum number of samples to filter minimum number of junctions by
min_samples <- 5

# read in files generated from combined shiba run
combined_junctions <- readr::read_tsv(
  combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in combined junctions table that have been deduplicated
deduplicated_combined_junctions <- readr::read_tsv(
  deduplicated_combined_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)

# read in files generated from merged shiba run
merged_junctions <- readr::read_tsv(
  merged_junctions_file,
  col_types = readr::cols(
    .default = "d",
    ID = "c",
    chr = "c",
    start = "d",
    end = "d"
  )
)


```

## Check tables for parsing errors

### Print parsing errors in combined junctions table

```{r check_parsing_errors}
combined_junctions_dat <- vroom::vroom(combined_junctions_file)
combined_junctions_problems <- vroom::problems(combined_junctions_dat)
combined_junctions_problems
```

We have `r nrow(unique(combined_junctions_problems$row))` junctions in the combined junctions file that are impacted by the concatenation bug in Shiba.
PSI files for events involving these junctions are expected to be problematic in the combined PSI file.

### Check for parsing errors in the merged and deduplicated junctions tables

```{r}
merged_junctions_dat <- vroom::vroom(merged_junctions_file)
```

```{r}
deduplicated_comb_junctions_dat <- vroom::vroom(deduplicated_combined_junctions_file)
deduplicated_comb_junctions_problems <- vroom::problems(deduplicated_combined_junctions)
```

No problems are present in merged_junctions file or deduplicated junctions file

## Check if files are identical

```{r}
# ensure column order is the same before checking if data frames are identical
deduplicated_combined_junctions <- deduplicated_combined_junctions |>
  dplyr::relocate(dplyr::all_of(target_pilot_samples), .after = dplyr::last_col())

merged_junctions <- merged_junctions |>
  dplyr::relocate(dplyr::all_of(target_pilot_samples), .after = dplyr::last_col())

identical(deduplicated_combined_junctions, merged_junctions)
```

Deduplicated and merged junctions are still not identical, so need to drill down on what junctions are different

## Check differences in junctions bedfiles

Dimensions of each bedfile

```{r}
dim(deduplicated_combined_junctions)
```

```{r}
dim(merged_junctions)
```

There are `r nrow(merged_junctions) - nrow(deduplicated_combined_junctions)` more junctions in the merged junction file compared to the combined junction file.

## Subsample junctions to those in chromosome 1

To allow for running on my laptop, subsample junctions files for those in chromosome 1 prior to checking for differences

```{r subsample_junctions}
deduplicated_combined_junctions_chr1 <- deduplicated_combined_junctions |> dplyr::filter(chr == "chr1")
merged_junctions_chr1 <- merged_junctions |> dplyr::filter(chr == "chr1")
```

Check head of deduplicated combined chromosome 1 junctions

```{r head_dedup_combined_junctions}
deduplicated_combined_junctions_chr1 |>
  dplyr::arrange(start) |>
  head()
```

Check head of merged chromosome 1 junctions

```{r head_merged_junctions}
merged_junctions_chr1 |>
  dplyr::arrange(start) |>
  head()
```

Many junctions also have counts below 10 across all samples.
These junctions are not expected to make it into PSI calculation based off default Shiba filters (\> 10 junctions).

Check number of junctions in each subsetted dataframe

```{r check_num_junctions_chr1}
nrow(deduplicated_combined_junctions_chr1)
nrow(merged_junctions_chr1)
```

There are `r nrow(merged_junctions_chr1) - nrow(deduplicated_combined_junctions_chr1)` more junctions in the merged bedfile, compared to the combined bedfile.
The chromosome 1-subsetted junction bedfiles are not identical.

### Filter for junctions with at least 10 junctions in min_samples

Filter for junctions in chromosome 1 whose counts are \< 10 in `r min_samples` samples and check for differences

```{r filter_junction_counts_merged}
merged_junctions_chr1_10 <- merged_junctions_chr1 |> dplyr::filter(
  # count number of columns per row with junction count of at least 10
  rowSums(
    dplyr::across(
      dplyr::all_of(target_pilot_samples), ~ . >= 10
    )
    # filter for rows where at least 5 columns (samples) have counts of at least 10
  ) >= 5
  )

merged_junctions_chr1_10 |>
  dplyr::arrange(start) |>
  head()
```

```{r filter_junction_counts_combined}

deduplicated_combined_junctions_chr1_10 <- deduplicated_combined_junctions_chr1 |> dplyr::filter(
  # count number of columns per row with junction count of at least 10
  rowSums(
    dplyr::across(
      dplyr::all_of(target_pilot_samples), ~ . >= 10
    )
    # filter for rows where at least 5 columns (samples) have counts of at least 10
  ) >= 5
  )

deduplicated_combined_junctions_chr1_10 |>
  dplyr::arrange(start) |>
  head()
```

Check for differences between counts-filtered junction files

```{r check_differences_counts_filtered}
nrow(deduplicated_combined_junctions_chr1_10)
nrow(merged_junctions_chr1_10)
```

There are more junctions in the combined bedfile, compared to the merged bedfile, after filtering for junctions with at least 10 counts in `r min_samples` samples.

### Check what's different in chromosome 1 junctions

```{r count_differences_chr1}

# print number of junctions in chromosome 1 only in combined junctions bedfile
combined_only_chr1 <- dplyr::setdiff(deduplicated_combined_junctions_chr1, merged_junctions_chr1)
nrow(combined_only_chr1)

# check fraction of junctions that are only in chr1 of combined junctions bedfile
frac_combined_only_chr1 <- nrow(combined_only_chr1) / nrow(deduplicated_combined_junctions_chr1)
frac_combined_only_chr1

# print number of junctions in chromosome 1 only in merged junctions bedfile
merged_only_chr1 <- dplyr::setdiff(merged_junctions_chr1, deduplicated_combined_junctions_chr1)
nrow(merged_only_chr1)

# check fraction of merged_only_chr1 junctions that are only in chr1 of merged combined junctions
frac_merged_only_chr1 <- nrow(merged_only_chr1) / nrow(merged_junctions_chr1)
frac_merged_only_chr1

```

Overall, only `r frac_combined_only_chr1 * 100`% of all chromosome 1 junctions are only found in the combined junctions file and `r frac_merged_only_chr1 * 100`% of all chromosome 1 junctions are only found in the in the merged junctions file.

```{r count_differences_chr1_countsfiltered}

# print number of junctions in chromosome 1 only in combined junctions bedfile after filtering for junctions with at least 10 counts in at least 5 samples
combined_only_chr1_10 <- dplyr::setdiff(deduplicated_combined_junctions_chr1_10, merged_junctions_chr1_10)
nrow(combined_only_chr1_10)

# check fraction of combined_only_chr1 junctions > 10 counts
nrow(combined_only_chr1_10) / nrow(deduplicated_combined_junctions_chr1)

# print number of junctions in chromosome 1 only in merged junctions bedfile after filtering for junctions with at least 10 counts in at least 5 samples
merged_only_chr1_10 <- dplyr::setdiff(merged_junctions_chr1_10, deduplicated_combined_junctions_chr1_10)
nrow(merged_only_chr1_10)

# check fraction of merged_chr1 junctions > 10 counts
nrow(merged_only_chr1_10) / nrow(merged_junctions_chr1)
```

If we filter for junctions with more than 10 counts, the fraction of events we miss in the merged junctions file is even smaller in chromosome 1.

### Print examples of junctions only in merged or combined tables

Print examples of junctions only in the combined junctions file

```{r junctions_missed_in_merged}
combined_only_chr1_10 |>
  dplyr::arrange(start) |>
  head()
```

For comparison, examine merged chromosome 1 counts-filtered junction table for junctions near chr1:18369-18370

```{r}
merged_junctions_chr1_10 |>
  dplyr::arrange(start) |>
  dplyr::filter((18369 -  10) < start &
                  start < (18369 + 10) ) |>
  head()
```

This junction is completely missed, with the closest coordinate being chr1:18369-18913.

Print examples of junctions only in the merged junctions file

```{r junctions_only_in_merged}
merged_only_chr1_10 |>
  dplyr::arrange(start) |>
  head()
```

For comparison, examine combined chromosome 1 counts-filtered junction table for junctions near chr1:268020-268021

```{r}
deduplicated_combined_junctions_chr1_10 |>
  dplyr::arrange(start) |>
  dplyr::filter((268020 -  10) < start &
                  start < (268020 + 10) ) |>
  head()
```

Position chr1:268020-268021 is actually a junction recorded in the combined junctions table.
The difference come in the counts going into each sample:

| sample | junction count in merged table | junction count in combined table |
|----|----|----|
| **SRR1712457** | 0 | 3 |
| **SRR1712459** | 0 | 3 |
| **SRR1712460** | 0 | 8 |
| **SRR1712461** | 0 | 3 |
| **SRR1712462** | 0 | 4 |
| ... | ... | ... |
| **SRR1799041** | 0 | 16 |

From scanning chr1:268020-268021, the counts differences arise from cases where the merged junctions counts have a value of 0 for a sample and the combined file is not zero for the same sample.
For all cases except for SRR1799041, the counts that are dropped in the merged table are < 10.
Because the values in the merged table are 0 for the counts that differ from the combined table, I suspect these differences stem from junctions that are not shared across samples (these appear as NA values which we replace with 0 in the final merged bedfile).
I think this means that junctions defined differently when Shiba runs on one vs. multiple samples in a way that I don't understand yet.
But since this difference impacts a very small fraction of junctions in chromosome 1, maybe it is not something worth trying to fix.

Another note - because the concatenated junctions bug impacted one junction with reads \> 10 in the combined junctions bedfile, I expect this will result in differences in the PSI table
