---
title: Check Merged vs. Combined TARGET Pilot GTFs
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

## Background

This notebook checks junctions.bed files from two Shiba run methods for differences.
The goal is to get a sense of how close files that go into calculating PSI values for events are to each other in the merged and combined Shiba methods.
Junctions files that are different would impact PSI calculation, so I am mainly looking to see if mismatched junctions are present and if present, how much of the data do they represent.

Files compared in this notebook are GTFs containing transcript annotations of 88 TARGET pilot samples, produced from running Shiba with two different methods.

"Combined" refers to shiba splice analysis run with shiba in a canonical way.
Inputs of this method are .bam files of each sample and a reference annotation GTF.
In the combined method, unmodified Shiba scripts are run on the input files using shiba.py, which calls on Shiba scripts that generate intermediate files and perform the final PSI calculation.
The relevant intermediate files that go into PSI calculation from this method are:

- GTF, which is made with \[bam2gtf.py\](https://github.com/Sika-Zheng-Lab/Shiba/blob/v0.8.1/src/bam2gtf.py. This script uses a stringtie merge command as part of that merges GTFs made from individual samples together with the reference GTF

"Merged" refers to shiba splice analysis run with `merge_results.smk`. 
The inputs of this method are junctions.bed and GTF files created by Shiba run on each sample separately.
`merge_results.smk` merges the GTFs using the same stringtie merge command Shiba uses:

The snakemake GTF merge command:
`stringtie --merge -p {threads} -G {input.reference_gtf} -o {output.merged_gtf} $manifest`
Where `$manifest` refers to the list of GTFs to merge

The canonical/combined Shiba GTF merge command, for reference (from [bam2gtf.py](https://github.com/Sika-Zheng-Lab/Shiba/blob/v0.8.1/src/bam2gtf.py)):
```
		"stringtie",
		"--merge",
		"-p", str(num_processors),
		"-G", reference_gtf,
		"-o", assembled_gtf
	] + gtf_list
```

### Analysis outline:
- Check if GTFs are identical

## Setup
```{r setup}
# import libraries
library(rtracklayer)
```

## Read in directories and files

Define directories and file paths

```{r directories_and_files}
### Directories ###

# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## combined shiba run results on target pilot samples ##
exploration_dir <- file.path(repo_root, "exploration")

# merged table eval dir
exploration_eval_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")

# shiba results dir
# target pilot exploration dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# gtf dir
combined_gtf_dir <- file.path(combined_results_dir, "annotation")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")

### Files ###
# combined GTF
combined_gtf_file <- file.path(combined_gtf_dir, "assembled_annotation.gtf")

## merged shiba target pilot files
merged_gtf_file <- file.path(merged_results_dir, "merged_gtf.gtf")

```

Read in files

```{r read_in_gtfs}

combined_gtf <- import(combined_gtf_file)
merged_gtf <- import(merged_gtf_file)

# quick and dirty check if they are identical
# this is a strict check and will fail if row order is different
identical(combined_gtf, merged_gtf)

```

## Sort GTFs for comparison

The GTFs are not identical - first check if row/column orders are different, or if there is another easy explanation why they are different by looking at the head

```{r}
head(combined_gtf)
```

```{r}
head(merged_gtf)
```
It looks like the row order is not the same in each GTF.
Attempt to sort the GTFs so that the rows appear in the same order, prior to comparison

```{r}
sorted_combined_gtf <- sort(combined_gtf)
sorted_merged_gtf <- sort(merged_gtf)

identical(sorted_combined_gtf, sorted_merged_gtf)
```

GTFs are still not identical, so print head to see if there is a clear reason

```{r}
head(sorted_merged_gtf)
```
```{r}
head(sorted_combined_gtf)
```

## Check differences in junctions bedfiles

```{r check_differences}
# convert sorted gtfs into dataframes
combined_df <- as.data.frame(sorted_combined_gtf)
merged_df <- as.data.frame(sorted_merged_gtf)

merged_gtf_only <- dplyr::setdiff(combined_df, merged_df)
merged_gtf_only |> head()
```

Check for closest match to chr1 start position 11121 in combined gtf
```{r}
combined_df |>
  dplyr::filter(
    seqnames == "chr1",
    start == "11121"
  )
```
Both combined GTF entries for chr1:11121-any match the merged GTF entries, so I'm unsure why it was flagged as different

Check if column names are the same between each dataframe
```{r}
dplyr::setdiff(names(merged_df), names(combined_df))
```
They are the same (setdiff returns 0)

Check if classes in each column are the same between data frames (they are)

```{r}
sapply(merged_df, class) == sapply(combined_df, class)
```
Check for differences column by column

```{r}
print("seqnames")

dplyr::setdiff(merged_df$seqnames, combined_df$seqnames)

print("start")

dplyr::setdiff(merged_df$start, combined_df$start) |> head()
dplyr::setdiff(combined_df$start, merged_df$start) |> head()

print("end")

dplyr::setdiff(merged_df$end, combined_df$end) |> head()
dplyr::setdiff(combined_df$end, merged_df$end) |> head()

print("width")

dplyr::setdiff(merged_df$width, combined_df$width) |> head()
dplyr::setdiff(combined_df$width, merged_df$width) |> head()

print("strand")

dplyr::setdiff(merged_df$strand, combined_df$strand) |> head()
dplyr::setdiff(combined_df$strand, merged_df$strand) |> head()

print("source")

dplyr::setdiff(merged_df$source, combined_df$source) |> head()
dplyr::setdiff(combined_df$source, merged_df$source) |> head()

print("type")

dplyr::setdiff(merged_df$type, combined_df$type) |> head()
dplyr::setdiff(combined_df$type, merged_df$type) |> head()

print("score")

dplyr::setdiff(merged_df$score, combined_df$score) |> head()
dplyr::setdiff(combined_df$score, merged_df$score) |> head()

print("phase")

dplyr::setdiff(merged_df$phase, combined_df$phase) |> head()
dplyr::setdiff(combined_df$phase, merged_df$phase) |> head()

print("gene_id")

dplyr::setdiff(merged_df$gene_id, combined_df$gene_id) |> head()
dplyr::setdiff(combined_df$gene_id, merged_df$gene_id) |> head()

print("transcript_id")

dplyr::setdiff(merged_df$transcript_id, combined_df$transcript_id) |> head()
dplyr::setdiff(combined_df$transcript_id, merged_df$transcript_id) |> head()

print("exon_number")

dplyr::setdiff(merged_df$exon_number, combined_df$exon_number) |> head()
dplyr::setdiff(combined_df$exon_number, merged_df$exon_number) |> head()

print("gene_name")

dplyr::setdiff(merged_df$gene_name, combined_df$gene_name) |> head()
dplyr::setdiff(combined_df$gene_name, merged_df$gene_name) |> head()

print("ref_gene_id")

dplyr::setdiff(merged_df$ref_gene_id, combined_df$ref_gene_id ) |> head()
dplyr::setdiff(combined_df$ref_gene_id, merged_df$ref_gene_id ) |> head()
```

Some of the differences seem like they come from stringtie-assigned transcripts that don't otherwise have transcript IDs

Investigate ELOA3CP - Does it really only appear in the merged dataframe and not the combined?

```{r}
merged_df |>
  dplyr::filter(gene_name == "ELOA3CP")
```


```{r}
combined_df |>
  dplyr::filter(gene_name == "ELOA3CP")
```
Yes, this transcript is only in the merged dataframe for some reason
