---
title: Compare non-RI events coordinates between merged and combined Shiba runs
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Background

Junctions files differ in intron-exon junctions.

GTFs differ, but we suspect the source of difference to come from the order in which files were passed into stringtie --merge and the use of multithreading.

We suspect that the junctions difference will only impact retained intron events.

If so, we would exclude this event type from the compendium.

If there are differences in other event types in the event coordinate files, we would need to change our merged compendium pipeline.

## Analysis outline

- Calculate Jaccard similarity score between events coordinates created from the merged and combined methods each event type

## Setup

### Define functions

```{r define_functions}

# read in list of splice event coordinate files and bind them into one dataframe
read_in_events <- function(file_paths) {
event_coords <- file_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  })
}

calculate_jaccard_for_ids <- function(event_type) {
  combined_set <- combined_event_coords_df[[event_type]]$pos_id
  merged_set <- merged_event_coords_df[[event_type]]$pos_id
  
  length(intersect(combined_set, merged_set))/ length(union(combined_set, merged_set))
}

# # combine event coordinates into one table, labeled by event type 
# all_events_table <- dplyr::bind_rows(event_coords, .id = "event_type") |>
#   # drop event_id column that assigns coordinates IDs like "SE_1" "SE_2"
#   # I suspect how these positions are numbered may be arbitrary and impacted by the order in which events are defined
#   # in other words I don't care if these IDs are different, so long as the pos_id are the same
#   dplyr::select(! event_id)
# }

```

### Read in file paths and files

```{r define_paths_and_files}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## combined shiba run results on target pilot samples ##
# parent directories
exploration_dir <- file.path(repo_root, "exploration")
# merged table eval dir
exploration_eval_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")

# shiba results dir
# target pilot combined shiba results dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# events dir
combined_events_dir <- file.path(combined_results_dir, "events")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")
# events dir
merged_events_dir <- file.path(merged_results_dir, "events")

### Files ###

# list of events coordinates files to compare between the merged and combined methods
event_files <- c(
  se = "EVENT_SE.txt",
  afe = "EVENT_AFE.txt",
  ale = "EVENT_ALE.txt",
  five = "EVENT_FIVE.txt",
  three = "EVENT_THREE.txt",
  mse = "EVENT_MSE.txt",
  mxe = "EVENT_MXE.txt",
  ri = "EVENT_RI.txt"
)

# construct events coordinate file paths of shiba results for combined run
combined_events_paths <-file.path(combined_events_dir, event_files)
names(combined_events_paths) <- names(event_files)

merged_events_paths <-file.path(merged_events_dir, event_files)
names(merged_events_paths) <- names(event_files)
```

Read in files

```{r read_in_files}
# read in combined splice event coordinates into one table
combined_event_coords_df <- read_in_events(combined_events_paths)

# read in merged splice event coordinates into one table
merged_event_coords_df <- read_in_events(merged_events_paths)

```

## Quantify similarity of position IDs in each event coordinate file

Check each event type one at a time to see how much of each type is different between the event coordinate files\

Outline of event types checked (calculate Jaccard similarity index for each event type in this order)

- se

<!-- -->

- afe

- ale

- five

- three

- mse

- mxe

- ri

#### Skipped exons (se)

```{r skipped_exon_jaccard}

# define sets to compare
# set 1 will be the combined method's event coordinates of the event type we are calcualting for
# set 2 will be the merged method's event coordinates

combined_set <- combined_event_coords_df$se$pos_id
merged_set <- merged_event_coords_df$se$pos_id

length(intersect(combined_set, merged_set))/ length(union(combined_set, merged_set))
```

Test function

```{r}
calculate_jaccard_for_ids(se)
```

#### Alternative first exons (afe)

```{r}

```

#### Alternative last exons (ale)

```{r}
combined_set <- combined_event_coords_df$ale$pos_id
merged_set <- merged_event_coords_df$ale$pos_id

length(intersect(combined_set, merged_set))/ length(union(combined_set, merged_set))
```

#### Alternative 5' splice site (five)

```{r}


```

#### Alternative 3' splice site (three)

```{r}

```

### mse

#### mxe

#### ri

```{r}
combined_set <- combined_event_coords_df$ri$pos_id
merged_set <- merged_event_coords_df$ri$pos_id

length(intersect(combined_set, merged_set))/ length(union(combined_set, merged_set))
```
### Quantify what proportion of all event types are different between the two datasets
This might be better done with a Jaccard index as well?


```{r check_pos_ids}
combined_only_pos_ids <- setdiff(combined_event_coords_df$pos_id, merged_event_coords_df$pos_id)
# count how many IDs are only in the combined file
length(combined_only_pos_ids)
paste0(length(combined_only_pos_ids) / length(combined_event_coords_df$pos_id) * 100, "% of position IDs are only in the combined events tables")

merged_only_pos_ids <- setdiff(merged_event_coords_df$pos_id, combined_event_coords_df$pos_id)
length(merged_only_pos_ids)
paste0(length(merged_only_pos_ids) / length(merged_event_coords_df$pos_id) * 100, "% of position IDs are only in the merged events tables")
```
## Conclusions

- Retained intron position IDs are more different between the events files produced by the merged and combined methods compared to other splice event types

- Differences in other event types are more minor and may be due to smaller differences stemming from multithreading or different merge order creating different transcript models in the GTFs used to create transcript coordinates


Although we expected the retained introns to be the only event type impacted by how junctions.bed is generated in the merged method, differences in all event types' position IDs are present.
Additionally, there are more AFE and ALE event types impacted by these differences than retained intron events for both merged and combined sets of event coordinates.

For the event coordinate files, the only input file is the merged GTF.
So these differences would come from the differences in GTF input file order and multithreading in the stringtie --merge step.
In other words, merging GTFs on different computing systems does alter the coordinates of splice events that are detected.

Do we want to consider re-running the pilot Shiba "combined" and "merged" runs on mustard with the same file order and on only one thread?
