---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  use_cache: TRUE
---

## Introduction

Compendium workflow v1 (workflow with updated GTF generation and merging rules) has been run on the TARGET pilot samples.
To understand what differences remain between PSI results generated from this workflow and the canonical Shiba run, a portion of the `compared_merged_psi_table` notebook is run on the PSI tables generated from the v1 workflow and the unaltered Shiba scripts.

**v1**: Refers to PSI results generated from the most recent compendium workflow method

**Combined**: Refers to PSI results generated from the unaltered Shiba script.

**Combined_dedup**: PSI valiues generated from running only the Shiba PSI.py script on the following inputs and command:

- event coordinates files generated from the Combined method

- Junctions.bed files generated from Combined, then duplicated with the deduplication commands used in the v1 workflow

```         
python $CONDA_PREFIX/share/shiba-0.8.1-0/src/psi.py -m \
10 -p \
30 -v \
--onlypsi \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/junctions/deduplicated_junctions.bed \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/events \
v1_dedup_target_pilot_combined
```

**Question:** Does merging v1 splice tables cause us to lose out on the trustworthiness of unannotated events to an extent that it impacts many samples?

As part of our pipeline, we plan to merge Shiba tables created for individual samples to obtain a final PSI table of all samples.
Before doing so, we use this notebook to test that merging PSI tables from v1 runs will not introduce a large amount of untrustworthy splice events.

As part of tracking differences between the two tables, we have labeled NA values introduced at various steps of the data processing pipeline with different negative values:

- -1: NA values assigned by Shiba (splice events with low coverage)

- NA: NA values from merging the v1 and combined tables.
  These NAs represent splice events that are only found in one analysis method (combined vs. v1)

We are particularly interested in the impact merging v1 tables has on unannotated event detection, since this is one novelty of using Shiba.

**Targeted questions this notebook is trying to answer:**

- What splice event types do we miss out on with the v1 tables method?
- Of splice events with high numbers of NA values, how many of these events only have a numeric PSI value in one (or a very low amount of) samples?
- How many NA values of different NA types are there in each type of splice table? How are they distributed across events? I am more concerned with the merge NAs (-2), as there is no good way to replace them in the v1 tables (they could either be 0 or 1). Another concern is events that are only found in the combined table, or only found in the v1 table, as there is no easy way to correct for those.
- Can the number of NA values be reduced in the v1 tables method if we filter for events with numeric PSI values in a minimum number of samples?
- What is the relationship between PSI values and NAs in each splice table type?
- What is the relationship between PSI values and NAs in each splice table type after filtering for events with numeric PSI values in a minimum number of samples? Assuming the Shiba NAs (-1) are from low read support, do the v1 and combined tables always call the same events as a Shiba NA? Merge NAs (-2) should arise from events dropped by Shiba due to lack of alternative transcripts for a gene, so we expect these NAs to be 0 or 1.
- How well correlated are the number of NA values in each event type for each PSI table type?
- How well correlated are the PSI values of events that are shared?

## Set up

```{r, setup, include=FALSE}
# load libraries
library(ggplot2)
library(patchwork)

# define global size for plots
global_size <- 20

# define plot theme (size) presets
plot_theme = list(
 theme_bw(),
 theme(
   legend.position = "right",
   axis.title = element_text(size = global_size),
   plot.title = element_text(size = global_size),
   axis.text = element_text(size = global_size),
   axis.text.x = element_text(
     hjust = 1,
     margin = margin(t = 0, r = 80, b = 0, l = 0),
     size = global_size
   ),
   legend.title = element_text(size=global_size),
   legend.text=element_text(size=global_size - 8)
 )
)

# color palette for splice event types
cbPalette <- c(
  afe = "#999999",
  ale = "#E69F00",
  five = "#56B4E9",
  mse = "#009E73",
  mxe = "#F0E442",
  ri = "#0072B2",
  se = "#D55E00",
  three = "#CC79A7"
)

# define minimum number of samples to filter splice table by
min_samples <- 5
```

### Define functions

```{r functions}

# PSI distribution plots of events in each method
psi_distributions <- function(
    df,
    event_type_value) {
  df |>
    dplyr::filter(event_type == event_type_value) |>
    ggplot(aes(PSI, fill = event_status)) +
    geom_histogram(bins = 20) +
    facet_wrap(vars(event_type, label, event_status), scales = "free_y") +
    plot_theme +
    # make facet labels bigger
    theme(strip.text.x = element_text(size = global_size),
          # rotate x axis labels so they don't overlap
          axis.text.x = element_text(angle =45))
}

# create summary table of events detected
make_event_summary <- function (events_table) {
  events_table |>
    dplyr::summarise(
      .by = c(event_type, label),
      # count number of events in each event type and annotation category
      total = dplyr::n(),
      shared_count = sum(shared_event),
      combined_only_count = sum(combined_event) - shared_count,
      v1_only_count = sum(v1_event) - shared_count,
      shared_percent = shared_count / total * 100,
      combined_only_percent = combined_only_count / total * 100,
      v1_only_percent = v1_only_count / total * 100,
    )
}

# plot event summary frequencies as percent and raw counts bar plots
plot_event_summary <- function(
    summary_df) {
  # pivot summary df longer for plotting
  long_summary_df <- summary_df |>
    tidyr::pivot_longer(
      cols = c(
        shared_count,
        shared_percent,
        combined_only_count,
        combined_only_percent,
        v1_only_count,
        v1_only_percent
      ),
      # extract percents and counts into v1 columns
      names_to = c("category", ".value"),
      names_pattern = "(.+)_(percent|count)"
    )

  # create percent stacked barplots
  percent_plot <-
    ggplot(long_summary_df, aes(fill = category, x = event_type, y = percent)) +
    geom_bar(position = "stack", stat = "identity") +
    plot_theme +
    labs(
      title = "% of splice events in combined, v1, or shared groups",
      x = "Annotation status of event",
      y = "Percentage of events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    plot_theme

  # create raw counts grouped barplot
  counts_plot <-
    ggplot(long_summary_df, aes(fill = category, x = event_type, y = count)) +
    geom_bar(position = "dodge", stat = "identity") +
    plot_theme +
    labs(
      title = "Number of splice events only in combined, v1, or shared groups",
      x = "Annotation status of event",,
      y = "Number of events"
    ) +
    facet_wrap(vars(label)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    plot_theme

  # arrange plots for printing
  percent_plot /
    counts_plot

}

# Histogram of quantified events
number_events_dist <- function(
    df,
    psi_type, title) {
    # construct histogram
    ggplot(df,  aes(x = .data[[psi_type]], fill = event_type)) +
    geom_histogram(binwidth = 1) +
    facet_wrap(vars(label)) +
    scale_fill_manual(values = cbPalette) +
    plot_theme +
    # make facet labels bigger
    theme(strip.text.x = element_text(size = global_size),
          # rotate x axis labels so they don't overlap
          axis.text.x = element_text(angle =45)) +
    ggtitle(title)
}

## Sample-level functions

# pivot event-level df long
pivot_long <- function (wide_event_df) {
  wide_event_df |>
    dplyr::select(starts_with("SRR"), event_type, pos_id, label) |>
    tidyr::pivot_longer(
      cols = matches("_PSI_(combined|v1)$"),
      names_to = c("sample", ".value"),
      names_pattern = "(.*)_PSI_(combined|v1)"
  )}

# event summary plots of frequency of events in each category
event_summary <- function(
    long_df,
    min_sample_value) {
  # filter samples by complete PSI value in min number of samples
  all_events_n_filtered <- long_df |>
    dplyr::group_by(pos_id) |>
    # count number of events are present in each shared status
    dplyr::mutate(
      combined_count = sum(method == "combined"),
      v1_count = sum(method == "v1")
    ) |>
    # filter for events with complete PSI values for minimum number of samples
    dplyr::filter(
      combined_count >= min_sample_value,
      v1_count >= min_sample_value
    ) |>
    dplyr::ungroup() |>
    # create summary table for plotting
    dplyr::summarise(
      .by = c(event_type, label),
      # count number of events in each event type and annotation category
      total = dplyr::n(),
      shared_count = sum(shared_event),
      combined_only_count = sum(combined_event) - shared_count,
      v1_only_count = sum(v1_event) - shared_count,
      shared_percent = shared_count / total * 100,
      combined_only_percent = combined_only_count / total * 100,
      v1_only_percent = v1_only_count / total * 100
    )
}

# Create summary table of number and percentage of events in each PSI-NA matchup category
na_comparison_summary <- function(long_df) {
  long_df |>
    dplyr::mutate(
      match_category =
        dplyr::case_when(
          v1 == -1 & combined == -1 ~ "both Shiba NAs",
          v1 == -1 & is.na(combined) ~ "no event in combined, Shiba NA in v1",
          is.na(v1) & combined >= 0 ~ "PSI only quantified in combined table",
          is.na(v1) & combined == -1 ~ "no event in v1, Shiba NA in combined",
          is.na(v1) & is.na(combined) ~ "both NA",
          v1 >= 0 & is.na(combined) ~ "PSI only quantified in v1 table",
          v1 >= 0 & combined >= 0 ~ "PSI quantified in both"
        )
    ) |>
    dplyr::summarise(
      .by = c(match_category),
      count = dplyr::n()
    ) |>
    dplyr::mutate(
      total = sum(count),
      percent = count / total * 100
    )
}

# plot PSI distributions of events with numeric PSI values in combined tables but are dropped in v1 tables
psi_values_lost_in_tables <- function(na_comparison_df, match_queried) {
  na_comparison_df |>
    # drop NAs from putting together the v1 and combined tables
    tidyr::drop_na() |>
    # assign match categories
    dplyr::mutate(
      match_category =
        dplyr::case_when(
          v1 == -1 & combined == -1 ~ "both Shiba NAs",
          v1 == -1 & combined >=0 ~ "v1 Shiba NA",
          v1 >= 0 & combined == -1 ~ "combined Shiba NA",
          v1 >= 0 & combined >= 0 ~ "PSI quantified in both",
          is.na(v1) & combined >= 0 ~ "PSI only quantified in combined table",
          v1 >= 0 & is.na(combined) ~ "PSI only quantified in v1 table"
        )
    ) |>
    # plot only events in combined table with numeric PSIs that are dropped in the v1 table
    dplyr::filter(match_category == match_queried,) |>
      ggplot(aes(combined, fill = event_type)) +
      geom_histogram(bins = 20) +
      facet_wrap(vars(label)) +
      scale_fill_manual(values = cbPalette) +
      plot_theme +
      # make facet labels bigger
      theme(strip.text.x = element_text(size = global_size),
            # rotate x axis labels so they don't overlap
            axis.text.x = element_text(angle =45))
}
```

## Directories and files

```{r}
## directories ##
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)
# define the data directories
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")
target_pilot_workflow_results_dir <- file.path(exploration_dir, "v1_workflow_target_pilot_results", "psi")
#output dir for long tables
target_pilot_output <- file.path(exploration_dir, "v1_psi_tables")

## combined / canonical shiba run results
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
# directory of shiba results produced from the same shiba run
combined_dir <- file.path(target_pilot_dir, "shiba_combined", "results")
# psi table directory
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# Check output dir exists; if not, create it
if (!dir.exists(target_pilot_output)) {
  dir.create(target_pilot_output)
}

## files ##
# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
v1_psi_files <- c(
  se = "PSI_SE.txt.gz",
  afe = "PSI_AFE.txt.gz",
  ale = "PSI_ALE.txt.gz",
  five = "PSI_FIVE.txt.gz",
  three = "PSI_THREE.txt.gz",
  mse = "PSI_MSE.txt.gz",
  mxe = "PSI_MXE.txt.gz",
  ri = "PSI_RI.txt.gz"
)

psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of shiba results of two samples run together
shiba_psi_paths <-file.path(combined_splice_results_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

v1_psi_paths <-file.path(target_pilot_workflow_results_dir, v1_psi_files)
names(v1_psi_paths) <- names(v1_psi_files)

## output ##
# long df output file of combined and v1 PSI tables
long_df_output <- file.path(target_pilot_output, "long_psi_df.rds")
```

Read in files

```{r}
# read merged splice table from v1 workflow on the TARGET pilot samples
v1_splice_results <- v1_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against v1 PSI dataframe
v1_psi_table <- purrr::list_rbind(v1_splice_results, names_to = "event_type")

# read in combined splice results to compare against v1 results
combined_splice_results <- shiba_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against v1 PSI dataframe
combined_psi_table <- purrr::list_rbind(combined_splice_results, names_to = "event_type")
```

### Obtain dimensions of each PSI table

```{r}
# dimensions of v1 PSI table
dim(v1_psi_table)
```

```{r}
# dimensions of combined PSI table
dim(combined_psi_table)
```

There are \~400 more splice events in the combined PSI table than the v1 table.

## What splice event types do we miss out on with the v1 tables method?

To help us prioritize what splice event types we may be the most confident in after merging v1 Shiba PSI tables, we are interested in seeing a breakdown of what event types are most represented in PSI tables constructed from different methods

```{r}
# create df of splice events that are shared between both combined and v1 tables
# This is an event-level dataframe (each row is one unique pos_id/event_id)
# columns are each samples' PSI value (with -1/-2/NA values indicating different NA categories)
# combined_event, v1_event, and shared_event columns label what tables the event is present in
all_events <- dplyr::full_join(
  combined_psi_table,
  v1_psi_table,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_combined", "_v1")) |>
  # categorize events by whether they are in the combined or v1 tables
  dplyr::mutate(
    # combined events are counted if the values are not all NA
    combined_event = ! dplyr::if_all(ends_with("_combined"), is.na),
    # v1 events are counted if the values for the event are not all NA (indicating the specific position ID is only found in the combined or v1 table)
    v1_event = ! dplyr::if_all(ends_with("_v1"), is.na),
    shared_event = combined_event & v1_event
  )

```

Print summary of counts across each method, across all event types

```{r}
# number of total events shared or present only in the combined and v1 tables
# this table counts events with -1 and -2 PSI values

all_methods_summary <- all_events |> dplyr::summarise(
    # count number of events in each event type and annotation category
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    v1_only_count = sum(v1_event) - shared_count
)

all_methods_summary
```

Although the majority of splice events (inclusive of NA values) are shared between methods, there are about twice as many splice events only in the v1 method's table than the combined method.

Print summary of counts and percentages across each method, broken down by event type

```{r}
# event-level summary of the number of each splice event type in each splice table
all_events_summary <- all_events |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    total = dplyr::n(),
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    v1_only_count = sum(v1_event) - shared_count,
    shared_percent = shared_count / total * 100,
    combined_only_percent = combined_only_count / total * 100,
    v1_only_percent = v1_only_count / total * 100,
)

all_events_summary |> dplyr::select(
  event_type,
  label,
  total,
  shared_percent,
  combined_only_percent,
  v1_only_percent)
```

Out of all the event types, annotated skipped exons and annotated multiple skipped exon types are the most shared between the two methods (above 90%).
In contrast, unannotated events of these types are only shared at 65% and 58% respectively.

The majority of annotated event types are shared between both tables, although there are more splice events unique to the v1 table (green) than the combined table (red).
Inflated values in the v1 tables may be due to -2 NAs which arise in the Shiba processing of v1 samples when when genes without multiple transcripts are dropped.

### How many events remain if we filter for events with PSI values in a minimum number of samples?

- Of splice events with high numbers of NA values, how many of these events only have a numeric PSI value in one (or a very low amount of) samples?

Add summary columns to dataframe

```{r}
all_events <- all_events |>
  dplyr::mutate(
    ## count PSI value types in v1 table ##
    # count the number of PSI values that are not one of our NA types
    n_quantified_v1 = rowSums(
      # select sample PSI values that are not one of our NA types
      dplyr::pick(dplyr::ends_with("_PSI_v1")) >= 0,
      # ignore NA values (events only in v1 or combined tables)
      na.rm = TRUE),
    # count the number of PSI values dropped due to insufficient reads
    n_shiba_na_v1 = rowSums(
      # select sample PSI values that have -1 NA type
      dplyr::pick(dplyr::ends_with("_PSI_v1")) == -1,
      na.rm = TRUE),
    # count number of PSI values dropped due to lack of transcript diversity in single samples
    n_dropped_v1 = rowSums(
      # elect sample PSI values that have -2 NA type
      dplyr::pick(dplyr::ends_with("_PSI_v1")) == -2
    ),
    ## count PSI value types in combined table ##
    # count the number of PSI values that are not one of our NA types
    n_quantified_combined = rowSums(
      # select sample PSI values that are not one of our NA types
      dplyr::pick(dplyr::ends_with("_PSI_combined")) >= 0,
      # ignore NA values (events only in v1 or combined tables)
      na.rm = TRUE),
    # count the number of PSI values dropped due to insufficient reads
    n_shiba_na_combined = rowSums(
      # select sample PSI values that have -1 NA type
      dplyr::pick(dplyr::ends_with("_PSI_combined")) == -1,
      na.rm = TRUE),
    # count number of PSI values dropped due to lack of transcript diversity in single samples
    n_dropped_combined = rowSums(
      # elect sample PSI values that have -2 NA type
      dplyr::pick(dplyr::ends_with("_PSI_combined")) == -2
    )
  )

```

Filter for minimum samples with numeric PSI values in the v1 table and check n_quantified.
The resulting summaries of this filtered table will be incomplete, as it will not contain information of events in the combined table that are present in min_samples but are missed completely in the v1 tables.

```{r}
# event-level summary of the number of each splice event type in each splice table
all_events_min_samples <- all_events |>
  # filter for events where there are numeric PSI values in at least 5 samples
  dplyr::filter(n_quantified_v1 >= min_samples)

# check what summary columns look like
all_events_min_samples  |>
  dplyr::select(pos_id, n_quantified_v1, n_shiba_na_v1, n_dropped_v1,
                n_quantified_combined, n_shiba_na_combined, n_dropped_combined) |>
  dplyr::arrange((n_quantified_v1)) |>
  head()
```

`n_quantified_v1` and `n_quantified_combined` columns have min values of 5, so filtering appears to be correct.

Observations:

- Although most events are shared after applying this filter and we still have unannotated events, we also still have events only found in the v1 table.
  These events are harder to deal with because they may be real events mislabeled by Shiba, or they may be unreal events (we cannot easily tell)

- Additionally, we still see that most unannotated events are no longer present after this filtering.

Observations:

- Once the "splice event with numeric value in at least 5 samples" filter is applied, there are no longer splice events only in the combined method's PSI table.

- The combined method is expected capture more events found in fewer samples, due to calculating splice events usage defined from a more complete set of events.

## How many of each splice event types are quantified?

This section plots the distributions of splice events quantified in each method.

### Histogram of quantified events

```{r}
#| label: fig-num_quantified_psi
#| fig-width: 11
#| fig-height: 8

number_events_dist(all_events, "n_quantified_v1", "Number of quantified PSI values in v1 table") /

number_events_dist(all_events, "n_quantified_combined", "Number of quantified PSI values in combined table") + plot_layout(guides = "collect")
```

Check skipped exon quantified PSI histograms

```{r}
#| label: fig-num_quantified_psi_se
#| fig-width: 11
#| fig-height: 8

all_events |>
  dplyr::filter(event_type == "se") |>
  number_events_dist("n_quantified_v1", "Number of quantified PSI values in v1 table") /

all_events |>
  dplyr::filter(event_type == "se") |>
  number_events_dist("n_quantified_combined", "Number of quantified PSI values in combined table") + plot_layout(guides = "collect")
```

As expected, we have more observations of quantified unannotated PSI values in the combined table (higher bars at the end of the X axis) in both sets of histograms.
More critically, the unannotated observations in the combined table that we lose in the v1 tables seem to be quantified in most of the TARGET pilot samples.

### Histogram of events dropped by Shiba due to insufficient read counts

The following histograms show the counts of PSI values that are DROPPED in each method.

```{r}
#| label: fig-num_low_read_na
#| fig-width: 11
#| fig-height: 8

number_events_dist(all_events, "n_shiba_na_v1", "Number of PSIs dropped due to insufficient read counts in v1 table") /

number_events_dist(all_events, "n_shiba_na_combined", "Number of PSIs dropped due to insufficient read counts in combined table") + plot_layout(guides = "collect")
```

There appear to be more events dropped across more samples in the combined results compared to the v1 results.
However, I think this is because the combined method captures more "rare" (e.g. present in only one sample) splice events.
This would increase the overall pool of possible loci that could be sequenced so sparsely that they do not actually make it into PSI calculation.

### Histogram of events dropped due to genes being thrown out in v1 method

I do not plot this for the combined table because they are all 0 (we only define what is dropped due to insufficient transcript representation based off what transcripts are missing when individual v1 tables are merged).

```{r}
#| label: fig-num_low_transcript_na
#| fig-width: 11
#| fig-height: 8

number_events_dist(all_events, "n_dropped_v1", "# PSIs dropped due to gene being thrown out in v1 table")
```

This category of missing values are inflated in unannotated events in the v1 table.

## How many of each splice event types are quantified?

## Examine sample-level splice events

### Make long df of events

Pivot longer for plotting sample-level splice event info

Check if result exists in cache

```{r}
use_cached_long <- file.exists(long_df_output) && params$use_cache
```

```{r}
# Pivot all events df longer
# Each row corresponds to a splice event found in one sample, so there are multiple rows of splice events with the same sample ID
if (use_cached_long) {
  long_all_events <- readr::read_rds(long_df_output)
} else {
long_all_events <- pivot_long(all_events)
}

# print column names
colnames(long_all_events)
```

### Write output

Write long table into output

```{r}
# long df
saveRDS(long_all_events, file = long_df_output)

```

### Examine relationships between of PSI and NA values of combined and v1 tables

#### Are PSI values of events that are shared between each method identical?

```{r}
long_psi_in_both_df <- long_all_events |> dplyr::filter(combined > 0 & v1 > 0)
identical(long_psi_in_both_df$combined, long_psi_in_both_df$combined)
```

Yes, PSI values of shared events in both methods are identical.

#### How well correlated are the PSI values and NAs in each splice table type?

To better understand patterns behind what PSI values are missing in each method, this notebook categorizes PSI values into "match categories", based on what the value (or NA type) for that position is in each method's PSI table.

Summarize how many of each of these match categories are present in the pilot tables

```{r}
important_categories <- c(
  "PSI quantified in both",
  "NA in v1, quantified in combined",
  "NA in combined, quantified in v1",
  "PSI quantified in combined, dropped in v1 table")

na_comparison_summary(long_all_events)
```

We care most about "PSI quantified in both", "NA in v1, quantified in combined", "NA in combined, quantified in v1", and "PSI quantified in combined, dropped in v1 table" categories.
These categories of match types have explanations for why PSI values may be missing.

- Events we expect are real and are captured in both methods: PSI quantified in both
- Events we expect are real and are missed or miscategorized in the v1 method: NA in v1, quantified in combined
- Events that may be real and miscategorized (but are difficult to confirm): NA in combined, quantified in v1
- Events we expect are real but are missed in the v1 method due to how events are defined in the GTF: PSI quantified in combined, dropped in v1 table

In this table, most mismatches (35%) stem from cases where the splice event is dropped in the v1 table due to only one gene being detected in the GTF, but there are too few junction counts for a splice event in that gene for the event to be assigned a PSI value in the combined table.
This category is not concerning, as these events would not appear in the "canonical" Shiba method.

The most concerning mismatch types in this table are "PSI quantified in combined, dropped in v1 table" (9%) and "PSI only quantified in combined table" (4%), as these represent missing splice events due to differences between each method.
Although the overall percent appears small, only 16% of PSI values are quantified in both tables, so we are missing about as many events as we can detect with this method.

#### Plot PSI distribution of PSI values lost in each method

This plot shows the distribution of the "true" values of -2 ("gene dropped") NAs in the combined table.

##### Quantified PSI values dropped in v1 tables due to lack of transcript diversity

```{r}
#| label: fig-psi_dist_of_combined_only_events
#| fig-width: 11
#| fig-height: 8

long_min_sample_all_events |>
    # drop NAs from putting together the v1 and combined tables
    tidyr::drop_na() |>
    # assign match categories
    dplyr::mutate(
      match_category =
        dplyr::case_when(
          v1 == -1 & combined == -1 ~ "both Shiba NAs",
          v1 == -1 & combined >=0 ~ "v1 Shiba NA",
          v1 >= 0 & combined == -1 ~ "combined Shiba NA",
          v1 >= 0 & combined >= 0 ~ "PSI quantified in both",
          is.na(v1) & combined >= 0 ~ "PSI only quantified in combined table",
          v1 >= 0 & is.na(combined) ~ "PSI only quantified in v1 table"
        )
    ) |>
    # plot only events in combined table with numeric PSIs that are dropped in the v1 table
    dplyr::filter(match_category == "PSI only quantified in v1 table") |>
      ggplot(aes(combined, fill = event_type)) +
      geom_histogram(bins = 20) +
      # facet_wrap(vars(label)) +
      scale_fill_manual(values = cbPalette) +
      plot_theme +
      # make facet labels bigger
      theme(strip.text.x = element_text(size = global_size),
            # rotate x axis labels so they don't overlap
            axis.text.x = element_text(angle =45))
```

The PSI values of the "gene dropped" NAs span the full range of PSI values, so we miss out entirely on a wide range of values using the v1 method.

Following this analysis, we concluded that the Shiba splice compendium workflow needed to be modified to allow for the "gene dropped" NA values to be quantified.
