---
title: Compare splice event coordinates between merged and combined Shiba runs
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Background

From analyzing junction count files produced by the merged and combined Shiba methods in `exploration/check_merged_junctions.qmd`, only exon-intron junctions were different.

Junctions.bed file exon-intron junctions are counted from exon-intron boundaries defined by `EVENT_RI.txt`, produced from the GTF of all samples' transcripts by `gtf2event.py` in Shiba.
Initial analysis of the GTFs produced by the merged and combined methods also revealed differences in the GTFs used to create the event coordinate files.
We suspect these differences stem from differences in the order in which files were passed into `stringtie --merge`, and the use of multithreading in merging.

We next wanted to answer how much GTF differences impacted similarity in the coordinates of splice events in the event coordinate files (like `EVENT_RI.txt`).
For instance, if only one event type's coordinates is disproportionately impacted by GTF differences, 
we may decide to not include quantification of that event type in the first release of the splice compendium.
Alternatively, if differences are present throughout all event types, 
we will need to alter how the GTF is processed to make it more similar to an unaltered Shiba run ("combined" method).

## Analysis outline

This notebook compares the similarity (Jaccard index) between splice event coordinates identified from different pairs of GTFs generated using different methods:

- Negative control (expected to be identical)

  - Two GTFs were generated with the unaltered Shiba v 0.8.1 `bam2gtf.py` script, with one thread, with the following commands:
    - `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py -i exploration/merging-psi-tables/target_pilot/experiment.tsv -r references/gencode.v47.primary_assembly.annotation.gtf -o results/merged_shiba/target_pilot_gtf_tests/one_thread_pilot_gtf.gtf -p 1 -v`
    - `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py -i exploration/merging-psi-tables/target_pilot/experiment.tsv -r references/gencode.v47.primary_assembly.annotation.gtf -o results/merged_shiba/target_pilot_gtf_tests_rep2/one_thread_pilot_gtf.gtf -p 1 -v`
  - Events coordinates files for both GTFs were generated with unaltered `gtf2events.py` using 10 threads (the amount passed to this script in the splice compendium merged snakemake workflow) with the following commands:
    - `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py -i target_pilot_gtf_tests/one_thread_pilot_gtf.gtf -r references/gencode.v47.primary_assembly.annotation.gtf -o shiba_pilot_gtf_events -p 10 -v`
    - `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py -i /scratch/celiang/target_pilot_gtfs/target_pilot_gtf_tests_rep2/one_thread_pilot_gtf.gtf -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf -o /scratch/celiang/shiba_event_to_gtf_tests/shiba_pilot_gtf_events_rep2 -p 10 -v`

- Merged vs. combined method: To what extent are event coordinates defined with the merged and combined methods different?

  - Merged GTF: Individual GTFs for each sample were first generated with `shiba-0.8.1-0/src/bam2gtf.py` with `Snakefile` by merging one sample's StringTie GTF and the reference GTF.
    Then, these GTFs were merged together, along with the reference with a final `stringtie —merge` command with `merge_results.smk`.
    4 threads were used in both merge steps (one-sample and 88 samples).

    - Events coordinates files were made using `$CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py` called by `merge_results.smk` , with 10 threads

  - Combined GTF: Separate GTFs for 88 TARGET pilot samples were greated with StringTie, then merged together with the reference using `stringtie -–merge` using the unaltered `shiba-0.8.1-0/src/bam2gtf` and 15 threads.

    - Events coordinates files were made using `$CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py` as part of the unaltered `shiba.py` workflow, with 15 threads.

- Multiple threads test: To what extent does using multiple threads change events coordinates defined by GTFs made with the same commands?

  -  Two GTFs were generated with the unaltered Shiba v 0.8.1 `bam2gtf.py` script, with 4 threads, with the following commands:

    - `time python $CONDA_PREFIX/share/shiba-0.8.1-0/src /bam2gtf.py -i exploration/merging-psi-tables/target_pilot/experiment.tsv -r references/gencode.v47.primary_assembly.a nnotation.gtf -o results/merged_shiba/target_pilot_gtf_tests_multithread/four_thread_pilot_gtf.gtf -p 4 -v`

    - `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py -i exploration/merging-psi-tables/target_pilot/experiment.tsv -r references/gencode.v47.primary_assembly.annotation.gtf -o results/merged_shiba/target_pilot_gtf_tests_multithread_rep2/four_thread_pilot_gtf.gtf -p 4 -v`

## Setup

### Define functions

```{r define_functions}
# read in list of splice event coordinate files and bind them into one list of dataframes
read_in_events <- function(file_paths) {
event_coords <- file_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  })
}

# Calculate list of Jaccard indices for event coordinate file pairs
# Similarity scores between events files made from combined vs. merged method
calculate_jaccard_indices_of_events <- function(
  list1,
  list2
) {
  purrr::map2(
  # read in lists of event coordinate dataframes to compare and iterate the two simultaneously
  list1,
  list2,
  # take the matching two dataframes from the input lists
  \(df1, df2){
    # obtain set of position IDs 
    set1 <- df1$pos_id
    set2 <- df2$pos_id
    
    # calculate Jaccard index of position IDs
    length(intersect(set1, set2)) / length(union(set1, set2))
  }
  )
  }

```

### Read in file paths and files

```{r define_paths_and_files}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## combined shiba run results on target pilot samples ##
# parent directories
exploration_dir <- file.path(repo_root, "exploration")
# merged table eval dir
exploration_eval_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")
# event coordinate file directory generated from TARGET pilot GTFs created with the same Shiba commands with one thread
# the events files themselves were generated with the same commands, but with 10 threads
events_dir_one_thread_bam2gtf <- file.path(exploration_dir, "merged_shiba_target_pilot", "shiba_gtf_to_event_pilot_tests")
# events files from replicate 1 of gtf generated from bam2gtf.py with one thread
events_dir_one_thread_bam2gtf_rep1 <- file.path(events_dir_one_thread_bam2gtf, "shiba_pilot_gtf_events")
# events files from replicate 2 of gtf generated from bam2gtf.py with one thread
events_dir_one_thread_bam2gtf_rep2 <- file.path(events_dir_one_thread_bam2gtf, "shiba_pilot_gtf_events_rep2")

# shiba results dir
# target pilot combined shiba results dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# events dir
combined_events_dir <- file.path(combined_results_dir, "events")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")
# events dir
merged_events_dir <- file.path(merged_results_dir, "events")

### Files ###

# list of events coordinates files to compare between the merged and combined methods
event_files <- c(
  se = "EVENT_SE.txt",
  afe = "EVENT_AFE.txt",
  ale = "EVENT_ALE.txt",
  five = "EVENT_FIVE.txt",
  three = "EVENT_THREE.txt",
  mse = "EVENT_MSE.txt",
  mxe = "EVENT_MXE.txt",
  ri = "EVENT_RI.txt"
)

# construct events coordinate file paths of shiba results for combined run
combined_events_paths <-file.path(combined_events_dir, event_files)
names(combined_events_paths) <- names(event_files)

merged_events_paths <-file.path(merged_events_dir, event_files)
names(merged_events_paths) <- names(event_files)

rep1_events_from_bam2gtf_one_thread_paths <- file.path(events_dir_one_thread_bam2gtf_rep1, event_files)
names(rep1_events_from_bam2gtf_one_thread_paths) <- names(event_files)

rep2_events_from_bam2gtf_one_thread_paths <- file.path(events_dir_one_thread_bam2gtf_rep2, event_files)
names(rep2_events_from_bam2gtf_one_thread_paths) <- names(event_files)

```

Read in files

```{r read_in_files}
# read in combined splice event coordinates into a list of dataframes
combined_events_list <- read_in_events(combined_events_paths)

# read in merged splice event coordinates into a list
merged_events_list <- read_in_events(merged_events_paths)

# replicate 1 of event coordinates of gtf generated from bam2gtf.py with one thread
rep1_events_from_bam2gtf_one_thread_list <- read_in_events(rep1_events_from_bam2gtf_one_thread_paths)

# replicate 2 of event coordinates of gtf generated from bam2gtf.py with one thread
rep2_events_from_bam2gtf_one_thread_list <- read_in_events(rep2_events_from_bam2gtf_one_thread_paths)

```

## Quantify similarity of position IDs in each event coordinate file pair

Calculate Jaccard similarity index for each set of position IDs for each splice event type identified from each Shiba run method

```{r}
jaccard_indices_comb_vs_merged <- calculate_jaccard_indices_of_events(combined_events_list, merged_events_list)

```


```{r calculate_jaccard}

# Similarity scores between events files made from combined vs. merged method
jaccard_indices_comb_vs_merged <- calculate_jaccard_indices_of_events(combined_events_list, merged_events_list, pos_id)

# Similarity scores between events files made with two identical runs of unaltered Shiba scripts (one thread for bam2gtf.py, 10 threads for gtf2events.py)
jaccard_indices_rep1_vs_rep2 <- calculate_jaccard_indices_of_events(rep1_events_from_bam2gtf_one_thread_list,
                                                                    rep2_events_from_bam2gtf_one_thread_list,
                                                                    pos_id)
```

Print results

```{r}
# convert list of jaccard indices into a dataframe
jaccard_df <- data.frame(
  event_type = names(jaccard_indices_comb_vs_merged),
  jaccard_index_combined_vs_merged = unlist(jaccard_indices_comb_vs_merged,
                         use.names = FALSE),
  jaccard_indices_rep1_vs_rep2 = unlist(jaccard_indices_rep1_vs_rep2,
                                        use.names = FALSE)
)

jaccard_df
```

## Conclusions

- Retained intron position IDs are the most different between the events files produced by the merged and combined methods, compared to other splice event types.

- Differences still exist in the other event types, where similarity scores range from 0.83-0.94 and likely stem from GTF differences from the different Shiba runs (different GTF merge order, multithreading).

- Events coordinates derived from GTF files made without multithreading with the same merge order are the same
