---
title: Compare splice event coordinates between merged and combined Shiba runs
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Background

From analyzing junction count files produced by the merged and combined Shiba methods in `exploration/check_merged_junctions.qmd`, only exon-intron junctions were different.

Junctions.bed file exon-intron junctions are counted from exon-intron boundaries defined by `EVENT_RI.txt`, produced from the GTF of all samples' transcripts by `gtf2event.py` in Shiba.
Initial analysis of the GTFs produced by the merged and combined methods also revealed differences in the GTFs used to create the event coordinate files.
We suspect these differences stem from differences in the order in which files were passed into `stringtie --merge`, the use of multithreading in merging, or the fact that the reference GTF is merged into the final GTF multiple times in the "merged" compendium workflow (`Snakefile` + `merge_results.smk`).

We next wanted to answer how much GTF differences impacted similarity in the coordinates of splice events in the event coordinate files (like `EVENT_RI.txt`).
For instance, if only one event type's coordinates is disproportionately impacted by GTF differences, we may decide to not include quantification of that event type in the first release of the splice compendium.
Alternatively, if differences are present throughout all event types, we will need to alter how the GTF is processed to make it more similar to an unaltered Shiba run ("combined" method).

Sources of differences we want to test in this notebook are:

- Use of multithreading in the GTF merge command

- Order of GTFs that are merged in

- Merging in of reference GTF multiple times in the workflow

## Analysis outline

This notebook compares the similarity (Jaccard index) between splice event coordinate position IDs from event coordinate files produced from GTFs run with different StringTie conditions:

- **Negative control comparison**: Comparing event coordinates generated from GTFs produced with the same commands (expected to be identical):

  - Two GTFs were generated with the unaltered Shiba v 0.8.1 `bam2gtf.py` script, with one thread, with the same commands
  - Events coordinates files for both GTFs were generated with unaltered `gtf2events.py` using 10 threads (the amount passed to this script in `merge_results.smk`) with the same commands. Position IDs from these coordinate files are compared for similarity with Jaccard indices in this notebook.
  - Replicate 1 files and commands used to generate them
    - **Input GTF replicate 1:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py   -i exploration/merging-psi-tables/target_pilot/experiment.tsv   -r references/gencode.v47.primary_assembly.annotation.gtf   -o results/merged_shiba/target_pilot_gtf_tests/one_thread_pilot_gtf.gtf   -p 1  -v`
    - **Events files generated from input GTF 1:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py   -i target_pilot_gtf_tests/one_thread_pilot_gtf.gtf   -r references/gencode.v47.primary_assembly.annotation.gtf   -o shiba_pilot_gtf_events   -p 10  -v`
  - Replicate 2 files and commands used to generate them
    - **Input GTF replicate 2:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py   -i exploration/merging-psi-tables/target_pilot/experiment.tsv   -r references/gencode.v47.primary_assembly.annotation.gtf   -o results/merged_shiba/target_pilot_gtf_tests_rep2/one_thread_pilot_gtf.gtf -p 1  -v`
    - **Events files generated from input GTF 2:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py   -i /scratch/celiang/target_pilot_gtfs/target_pilot_gtf_tests_rep2/one_thread_pilot_gtf.gtf   -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf   -o /scratch/celiang/shiba_event_to_gtf_tests/shiba_pilot_gtf_events_rep2   -p 10  -v`

- **Multiple threads test:** To what extent does using multiple threads change events coordinates defined by GTFs made with the same commands?

  - Two GTFs were generated with the unaltered Shiba v 0.8.1 `bam2gtf.py` script, with 15 threads, with the same commands

  - Events coordinates files for both GTFs were generated with unaltered `gtf2events.py` using 10 threads with the same commands.
    Position IDs from these coordinate files are compared for similarity with Jaccard indices in this notebook.

  - Replicate 1 files and commands used to generate them:

    - **Input GTF replicate 1:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py   -i experiment_scratch.tsv   -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf   -o target_pilot_gtfs/15_threads_rep1/15_threads_pilot_gtf.gtf   -p 15  -v`

    - **Events files generated from input GTF 1:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py   -i target_pilot_gtfs/15_threads_rep1/15_threads_pilot_gtf.gtf   -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf   -o shiba_gtf_to_event_pilot_tests/15threads_events_rep1   -p 10  -v`

  - Replicate 2 files and commands used to generate them:

    - **Input GTF replicate 2:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py   -i experiment_scratch.tsv   -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf   -o target_pilot_gtfs/15_threads_rep2/15_threads_pilot_gtf.gtf   -p 15  -v`

    - **Events files generated from input GTF 2:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py   -i target_pilot_gtfs/15_threads_rep2/15_threads_pilot_gtf.gtf   -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf   -o shiba_gtf_to_event_pilot_tests/15threads_events_rep2   -p 10  -v`

- **Different GTF merge order:** To what extent does merging GTFs in different orders (as defined by the sample order in `experiment.tsv`) change event coordinate definition?

  - Scrambled merge order files and commands used to generate them:
    - **Input GTF 1 (scrambled merge order):** Was generated by first randomly scrambling the row order of `experiment.tsv` with `exploration/merged-method-target-pilot-eval/scramble_experiment_file_sample_order.R` , then running the unaltered Shiba v 0.8.1 `bam2gtf.py` with the scrambled experiment.tsv file using the following command: `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/bam2gtf.py  -i scrambled_experiment.tsv  -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf  -o target_pilot_gtfs/15_threads_scrambled/15_threads_scrambled_gtf.gtf  -p 15 -v`
    - **Events files generated from input GTF 1:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py  -i target_pilot_gtfs/15_threads_scrambled/15_threads_scrambled_gtf.gtf  -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf  -o shiba_gtf_to_event_pilot_tests/15_threads_scrambled  -p 10 -v`
  - "Control" merge order files:
    - Input GTF replicate 1 from Multiple Threads Test
    - Events files derived from input GTF replicate 1 from Multiple Threads Test

- **Merging in of reference GTF multiple times in the workflow:** Another difference between the merged and combined method workflows lies in how many times the reference GTF is merged in to the final GTF.
  In the merged method, the reference GTF is merged in once per sample to make an intermediate GTF from each bam, then once again when all intermediate GTFs are merged together.
  In contrast, the combined method creates a GTF from each sample's bam without merging in the reference and only merges in the reference GTF once in the final merge step where all samples' GTFs are merged together.
  To generate position IDs to test differences resulting from these two methods, the following code is run:

  - **"Merge reference GTF once" test**

    - **Input GTF 1:** `merge_gtfs.smk` is used to create a merged GTF from 88 pilot TARGET samples that should replicate the "combined" method of merging GTFs, where the reference is only merged in once.

    - **Events files generated from input GTF 1:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py  -i /private/groups/treehouse/working-projects/celiang/splicing-compendium/results/main_merged_gtf_tests/target_pilot/merged_gtf.gtf  -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf  -o shiba_gtf_to_event_pilot_tests/main_merged_gtf_test_target_pilot  -p 10 -v`

  - **"Merge reference GTF multiple times" test**

    - **Input GTF 2:** `merge_gtf_multiple_refs.smk` is used to create GTF where the reference GTF is merged in multiple times (once, when the GTF is generated from a sample bam, and again when all sample GTFs are merged together)

    - **Events files generated from input GTF 1:** `python $CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py  -i /private/groups/treehouse/working-projects/celiang/splicing-compendium/results/multiple_refs_merged_gtf_tests/target_pilot/merged_gtf.gtf  -r /private/groups/treehouse/working-projects/celiang/splicing-compendium/references/gencode.v47.primary_assembly.annotation.gtf  -o shiba_gtf_to_event_pilot_tests/multiple_refs_merged_gtf_test_target_pilot  -p 10 -v`

- **Merged vs. combined method: To what extent are event coordinates defined with the merged and combined methods different?**

  - Merged method files and commands used to generate them:

    - **Merged input GTF:** Individual GTFs for each sample were first generated with `shiba-0.8.1-0/src/bam2gtf.py` with `Snakefile` by merging one sample’s StringTie GTF and the reference GTF.
      Then, these GTFs were merged together, along with the reference with a final `stringtie —merge` command with `merge_results.smk`.
      4 threads were used in the merge step.

    - **Events files generated from input merged GTF:** Events coordinates files were made using `$CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py` called by `merge_results.smk` , with 10 threads

  - Combined method files and commands used to generate them:

    - **Combined input GTF:** Separate GTFs for 88 TARGET pilot samples were greated with StringTie, then merged together with the reference using `stringtie -–merge` using the unaltered `shiba-0.8.1-0/src/bam2gtf` and 15 threads.

    - **Events files generated from input combined GTF:** Events coordinates files were made using `$CONDA_PREFIX/share/shiba-0.8.1-0/src/gtf2event.py` as part of the unaltered `shiba.py` workflow, with 15 threads.

## Setup

### Define functions

```{r define_functions}
# read in list of splice event coordinate files and bind them into one list of dataframes
read_in_events <- function(file_paths) {
event_coords <- file_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  })
}

# Calculate list of Jaccard indices for event coordinate file pairs
# Similarity scores between events files made from combined vs. merged method
calculate_jaccard_indices_of_events <- function(
  list1,
  list2
) {
  purrr::map2(
  # read in lists of event coordinate dataframes to compare and iterate the two simultaneously
  list1,
  list2,
  # take the matching two dataframes from the input lists
  \(df1, df2){
    # obtain set of position IDs 
    set1 <- df1$pos_id
    set2 <- df2$pos_id
    
    # calculate Jaccard index of position IDs
    length(intersect(set1, set2)) / length(union(set1, set2))
  }
  )
  }

```

### Read in file paths and files

```{r define_paths_and_files}
# find the root-level repo directory so results files can be accessed
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## parent directories ##
exploration_dir <- file.path(repo_root, "exploration")
method_tests_dir <- file.path(exploration_dir, "merged_shiba_target_pilot")
splice_event_results_dir <- file.path(method_tests_dir, "shiba_gtf_to_event_pilot_tests")
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
# merged vs. combined TARGET pilot result dirs
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")

# negative control event directories
one_thread_events_rep1 <- file.path(splice_event_results_dir, "shiba_pilot_gtf_events")
one_thread_events_rep2 <- file.path(splice_event_results_dir, "shiba_pilot_gtf_events_rep2")

# multithreading test event directories
multithread_events_rep1 <- file.path(splice_event_results_dir, "15threads_events_rep1")
multithread_events_rep2 <- file.path(splice_event_results_dir, "15threads_events_rep2")

# scrambled merge order test event directory
scrambled_merge_order_events <- file.path(splice_event_results_dir, "15_threads_scrambled")

# merging reference GTF once
merge_gtf_one_ref_events <- file.path(splice_event_results_dir, "main_merged_gtf_test_target_pilot")

# merging reference GTF multiple times
merge_gtf_multiple_refs_events <- file.path(splice_event_results_dir, "multiple_refs_merged_gtf_test_target_pilot")

# merged vs. combined method shiba run events directories
combined_events_dir <- file.path(combined_results_dir, "events")
merged_events_dir <- file.path(merged_results_dir, "events")

### Files ###

# list of events coordinates files to compare between the merged and combined methods
event_files <- c(
  se = "EVENT_SE.txt",
  afe = "EVENT_AFE.txt",
  ale = "EVENT_ALE.txt",
  five = "EVENT_FIVE.txt",
  three = "EVENT_THREE.txt",
  mse = "EVENT_MSE.txt",
  mxe = "EVENT_MXE.txt",
  ri = "EVENT_RI.txt"
)

## construct events coordinate file paths for event files to read in ##
# negative control
neg_ctrl_rep1_paths <- file.path(one_thread_events_rep1, event_files)
names(neg_ctrl_rep1_paths) <- names(event_files)

neg_ctrl_rep2_paths <- file.path(one_thread_events_rep2, event_files)
names(neg_ctrl_rep2_paths) <- names(event_files)

# multithreading
multithread_rep1_paths <- file.path(multithread_events_rep1, event_files)
names(multithread_rep1_paths) <- names(event_files)

multithread_rep2_paths <- file.path(multithread_events_rep2, event_files)
names(multithread_rep2_paths) <- names(event_files)

# scrambled merge order
scrambled_merge_order_paths <- file.path(scrambled_merge_order_events, event_files)
names(scrambled_merge_order_paths) <- names(event_files)

# merging refs once vs. multiple times
one_ref_merge_paths <- file.path(merge_gtf_one_ref_events, event_files)
names(one_ref_merge_paths) <- names(event_files)

multiple_refs_merge_paths <- file.path(merge_gtf_multiple_refs_events, event_files)
names(multiple_refs_merge_paths) <- names(event_files)

# merged vs. combined method 
merged_events_paths <-file.path(merged_events_dir, event_files)
names(merged_events_paths) <- names(event_files)

combined_events_paths <-file.path(combined_events_dir, event_files)
names(combined_events_paths) <- names(event_files)
```

Read in files

```{r read_in_files}
# read in each list of files into a list of dataframes

# negative control 
neg_ctrl_rep1_list <- read_in_events(neg_ctrl_rep1_paths)
neg_ctrl_rep2_list <- read_in_events(neg_ctrl_rep2_paths)

# multithreading
multithread_rep1_list <- read_in_events(multithread_rep1_paths)
multithread_rep2_list <- read_in_events(multithread_rep2_paths)

# scrambled gtf merge prder
scrambled_list <- read_in_events(scrambled_merge_order_paths)

# one ref vs multiple refs being merged
one_ref_merge_list <- read_in_events(one_ref_merge_paths)
multiple_refs_merge_list <- read_in_events(multiple_refs_merge_paths)

# merged vs. combined method
merged_events_list <- read_in_events(merged_events_paths)
combined_events_list <- read_in_events(combined_events_paths)
```

## Quantify similarity of position IDs in each event coordinate file pair

Calculate Jaccard similarity index for each set of position IDs for each splice event type identified from each Shiba run method

```{r calculate_jaccard}

# negative control comparison
jaccard_indices_neg_ctrl <- calculate_jaccard_indices_of_events(
  neg_ctrl_rep1_list,
  neg_ctrl_rep2_list)

# multithreading comparison
jaccard_indices_multithreading <- calculate_jaccard_indices_of_events(
  multithread_rep1_list,
  multithread_rep2_list
)

# scrambled GTF order comparison (same threads)
jaccard_indices_scrambled <- calculate_jaccard_indices_of_events(
  multithread_rep1_list,
  scrambled_list
)

# merging reference gtfs multiple times
jaccard_indices_multiple_refs <- calculate_jaccard_indices_of_events(
  one_ref_merge_list,
  multiple_refs_merge_list)

# merging reference once vs. combined method
jaccard_indices_one_ref <- calculate_jaccard_indices_of_events(
  one_ref_merge_list,
  combined_events_list
)

# merged vs. combined method
jaccard_indices_merged_vs_combined <- calculate_jaccard_indices_of_events(
  merged_events_list,
  combined_events_list)
```

Print results

```{r}
#| label: tbl-jaccard
#| tbl-cap: Jaccard indices for comparisons among methods, separated by event type. 

# convert list of jaccard indices into a dataframe
jaccard_df <- data.frame(
  event_type = names(jaccard_indices_merged_vs_combined),
  neg_ctrl = unlist(jaccard_indices_neg_ctrl, use.names = FALSE),
  same_threads = unlist(jaccard_indices_multithreading, use.names = FALSE),
  scrambled = unlist(jaccard_indices_scrambled, use.names = FALSE),
  multiple_refs_vs_one_ref = unlist(jaccard_indices_multiple_refs, use.names = FALSE),
  one_ref_merge_vs_combined = unlist(jaccard_indices_one_ref, use.names = FALSE),
  merged_vs_combined = unlist(jaccard_indices_merged_vs_combined, use.names = FALSE)
)

jaccard_df
```

## Conclusions

- Retained intron position IDs are the most different between the events files produced by the merged and combined methods, compared to other splice event types.

- Differences still exist in the other event types in the merged vs. combined method comparison, where similarity scores range from 0.83-0.94 and likely stem from GTF differences from the different Shiba runs.

- Events coordinates derived from GTF files made without multithreading with the same merge order, with multithreading and the same merge order, and with different merge orders are the same.

- The source of differences lies in merging in the reference GTF multiple times in the merged method.
  When GTFs are merged with the reference only once in a separate snakemake workflow, the event coordinates are identical to those generated from the canonical Shiba workflow.
