---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  use_cache: TRUE
---

## Introduction

Compendium workflow v1 (workflow with updated GTF generation and merging rules) has been run on the TARGET pilot samples.
To understand what differences remain between PSI results generated from this workflow and the canonical Shiba run, 
this notebook compares what percentage of events remain that can only be found in the current v1 workflow and the canonical Shiba run methods.

**v1**: Refers to PSI results generated from the most recent compendium workflow method

**rep2_v1**: Refers to PSI results generated from the most recent compendium workflow method, generated with the same commands and parameters as v1 but written in a different output directory.
rep2_v1 will be used to test whether identical runs of the workflow will result in the same proportion of different events, due to randomness from GTFs being generated from scratch in each workflow run.

**Combined**: Refers to PSI results generated from the unaltered Shiba script.

**Combined_dedup**: PSI valiues generated from running only the Shiba PSI.py script on the following inputs and command:

- event coordinates files generated from the Combined method

- Junctions.bed files generated from Combined, then duplicated with the deduplication commands used in the v1 workflow

```         
python $CONDA_PREFIX/share/shiba-0.8.1-0/src/psi.py -m \
10 -p \
30 -v \
--onlypsi \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/junctions/deduplicated_junctions.bed \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/events \
v1_dedup_target_pilot_combined
```

For all PSI tables loaded in, NA values assigned by Shiba within each method listed above are replaced with -1,
to distinguish NAs from splice events with low coverage apart from NAs from splice events that are not matched between each method.

## Set up
Define functions

```{r functions}
# read merged splice table from v1 workflow on the TARGET pilot samples
read_psi_tables <- function(psi_paths) {
  psi_results <- psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against other PSI tables
psi_table <- purrr::list_rbind(psi_results, names_to = "event_type")

# return the psi table
return(psi_table)
}
```


## Directories and files

```{r directories_and_files}
## directories ##
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)
# define the data directories
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")
# these contain zipped psi files
target_pilot_workflow_results_dir <- file.path(exploration_dir, "v1_workflow_target_pilot_results", "psi")
rep2_target_pilot_results_dir <- file.path(exploration_dir, "v1_target_pilot_rep2", "psi")

#output dir for long tables
target_pilot_output <- file.path(exploration_dir, "v1_psi_tables")

## combined / canonical shiba run results
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
# directory of shiba results produced from the same shiba run
combined_dir <- file.path(target_pilot_dir, "shiba_combined", "results")
# psi table directory
# these contain unzipped psi files
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# psi table dir of deduplicated combined pilot results
# these are unzipped files
dedup_combined_results_dir <- file.path(exploration_dir, "dedup_target_pilot_combined_results")

# Check output dir exists; if not, create it
if (!dir.exists(target_pilot_output)) {
  dir.create(target_pilot_output)
}

## files ##
# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
v1_psi_files <- c(
  se = "PSI_SE.txt.gz",
  afe = "PSI_AFE.txt.gz",
  ale = "PSI_ALE.txt.gz",
  five = "PSI_FIVE.txt.gz",
  three = "PSI_THREE.txt.gz",
  mse = "PSI_MSE.txt.gz",
  mxe = "PSI_MXE.txt.gz",
  ri = "PSI_RI.txt.gz"
)

psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of combined run shiba results
shiba_psi_paths <-file.path(combined_splice_results_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

# psi table paths of psi.py run on deduplicated combined run junctions.bed
dedup_shiba_psi_paths <-file.path(dedup_combined_results_dir, psi_files)
names(dedup_shiba_psi_paths) <- names(psi_files)

# psi table paths of v1 workflow on target pilot samples
# these psi tables are zipped so can use the 'v1_psi_files' names
v1_psi_paths <-file.path(target_pilot_workflow_results_dir, v1_psi_files)
names(v1_psi_paths) <- names(v1_psi_files)

rep2_v1_psi_paths <- file.path(rep2_target_pilot_results_dir, v1_psi_files)
names(rep2_v1_psi_paths) <- names(v1_psi_files)
```

Read in files
```{r read_in_files}
# read merged splice table from v1 workflow on the TARGET pilot samples
v1_psi_table <- read_psi_tables(v1_psi_paths)
rep2_v1_psi_table <- read_psi_tables(rep2_v1_psi_paths)

# read in 'negative control' canonical shiba run tables
combined_psi_table <- read_psi_tables(shiba_psi_paths)
dedup_combined_psi_table <- read_psi_tables(dedup_shiba_psi_paths)
```

### Obtain dimensions of each PSI table

```{r}
# dimensions of v1 PSI table
dim(v1_psi_table)
# dimensions of rep2 v1 PSI table
dim(rep2_v1_psi_table)
# dimensions of combined PSI table
dim(combined_psi_table)
# dimensions of deduplicated combined PSI table
dim(dedup_combined_psi_table)
```

There are \~400 more splice events in the combined PSI table than the v1 table.
Check if PSI tables are identical

```{r check_identical}
# check if 
identical(v1_psi_table, dedup_combined_psi_table)

identical(combined_psi_table, dedup_combined_psi_table)
```

The combined PSI tables and PSI tables from running psi.py on the combined run's event files and deduplicated junctions are identical
This kind of makes sense to me because the final PSI tables don't have duplicated junction pos_ids; so I think the bug didn't impact the PSI calculation since shiba just "picked" the valid junction rows?

## What splice event types do we miss out on with the v1 tables method?

To help us prioritize what splice event types we may be the most confident in after merging v1 Shiba PSI tables, we are interested in seeing a breakdown of what event types are most represented in PSI tables constructed from different methods

```{r}
# create df of splice events that are shared between both combined and v1 tables
# This is an event-level dataframe (each row is one unique pos_id/event_id)
# columns are each samples' PSI value with -1 indicating low read NA; NA indicating event not shared between methods
# combined_event, v1_event, and shared_event columns label what tables the event is present in
all_events <- dplyr::full_join(
  dedup_combined_psi_table,
  v1_psi_table,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_combined", "_v1")) |>
  # categorize events by whether they are in the combined or v1 tables
  dplyr::mutate(
    # combined events are counted if the values are not all NA
    combined_event = ! dplyr::if_all(ends_with("_combined"), is.na),
    # v1 events are counted if the values for the event are not all NA (indicating the specific position ID is only found in the combined or v1 table)
    v1_event = ! dplyr::if_all(ends_with("_v1"), is.na),
    shared_event = combined_event & v1_event
  )

```

Print summary of counts across each method, across all event types

```{r}
# number of total events shared or present only in the combined and v1 tables
# this table counts events with -1 and -2 PSI values

all_methods_summary <- all_events |> dplyr::summarise(
    # count number of events in each event type and annotation category
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    v1_only_count = sum(v1_event) - shared_count
)

all_methods_summary
```

Although the majority of splice events (inclusive of NA values) are shared between methods, some events remain that are only in the combined or v1 method

Print summary of counts and percentages across each method, broken down by event type

```{r}
# event-level summary of the number of each splice event type in each splice table
all_events_summary <- all_events |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    total = dplyr::n(),
    shared_count = sum(shared_event),
    combined_only_count = sum(combined_event) - shared_count,
    v1_only_count = sum(v1_event) - shared_count,
    shared_percent = shared_count / total * 100,
    combined_only_percent = combined_only_count / total * 100,
    v1_only_percent = v1_only_count / total * 100,
)

all_events_summary |> dplyr::select(
  event_type,
  label,
  total,
  shared_percent,
  combined_only_percent,
  v1_only_percent)
```

Check what the unshared events look like
```{r}
unshared_events <- all_events |> dplyr::filter(
  shared_event == FALSE,
  event_type == "se"
) |>
  dplyr::arrange(pos_id)

head(unshared_events)
```

From looking at chr 10 iof the unshared skipped exon events, these are actually all the same position ID; they have just been assigned a different gene
So the error likely comes from random things different in the GTF


