---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  use_cache: TRUE
---

## Introduction

Treehouse Compendium version 1 has been run on the TARGET pilot samples.
This workflow includes generating GTFs from alignments, merging GTFs of all samples, merging separately-produced junction bedfiles for each sample, and running Shiba on the merged GTF and junction file as inputs to produce event coordinate and PSI files.

To understand what differences remain between PSI results generated from Treehouse Compendium v1 and the canonical Shiba run, this notebook compares what percentage of events remain that can only be found in the workflow and the canonical Shiba method.

Below are names used to refer to different workflow runs on TARGET pilot samples:

### PSI tables produced using canonical (unaltered) Shiba scripts

**Combined**: Refers to PSI results generated from the unaltered Shiba scripts, using `shiba.py`:

```         
shiba.py -p 15 exploration/merging-psi-tables/target_pilot/shiba_config.yaml
```

There is a bug in Shiba v0.8.1 that causes some `junctions.bed` position IDs (coordinates) to be duplicated.
To check if duplicated junctions impact PSI calculation, I also compare the PSI tables below:

**Combined_dedup**: PSI values generated from running only the Shiba `PSI.py` script on the following inputs and command:

- event coordinates files generated from the Combined (canonical Shiba) method above

- `junctions.bed` files generated from Combined, then duplicated with the deduplication commands used in the v1 snakemake workflow in the `merge_jucntions` rule

```         
python $CONDA_PREFIX/share/shiba-0.8.1-0/src/psi.py -m \
10 -p \
30 -v \
--onlypsi \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/junctions/deduplicated_junctions.bed \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/events \
v1_dedup_target_pilot_combined
```

### PSI table results produced by running Treehouse Compendium v1:

Two replicates of Treehosue Compendium v1 workflow were run to check whether repeated runs of the workflow (same command, different output directory) produce different PSI tables.

**rep1**: Refers to PSI results generated from the most recent compendium v1 workflow method **rep2**: Refers to PSI results generated from the most recent compendium v1 workflow method, generated with the same commands and parameters as rep1 but written in a different output directory.

### Comparisons in notebook

Are combined and deduplicated results identical?

- combined vs. combined_dedup

Are there different events in Compendium vs. canonical Shiba methods?

- rep1 vs. combined

Are there differences in PSI tables if compendium workflow is run twice (are two replicate runs different?)

- rep1 vs. rep2

Are GTFs generated from the same sample in rep1 and rep2 identical?

For all PSI tables loaded in, NA values assigned by Shiba within each method listed above are replaced with -1, to distinguish NAs from splice events with low coverage apart from NAs from splice events that are not matched between each method.

## Set up

### Define functions

```{r functions}

## Reading in PSI tables ##

# construct paths to psi tables
construct_psi_paths <- function(psi_dir, psi_file_list) {
  # construct paths to individual psi files
  psi_paths <-file.path(psi_dir, psi_file_list)
  # name each path according to event type
  names(psi_paths) <- names(psi_file_list)
  
  # returned the named vector of psi paths
  return(psi_paths)
}

# read in and merge splice tables
read_psi_tables <- function(psi_paths) {
  psi_results <- psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against other PSI tables
psi_table <- purrr::list_rbind(psi_results, names_to = "event_type")

# return the psi table
return(psi_table)
}

## Manipulating PSI tables for comparison ##

# merge tables by all identifying fields for a splice event ID (event_type", "pos_id", "gene_id", "label")
join_tables_incl_genes <- function(table1, table2) {
  
  # create the merged table of PSI tables to compare
  joined_table <- dplyr::full_join(
  table1,
  table2,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_table1", "_table2")) |>
  # categorize events by whether they are in the v1 or rep2 tables
  dplyr::mutate(
    # Count events in one table or other if the values are not all NA
    table1_event = ! dplyr::if_all(ends_with("_table1"), is.na),
    # rep2 events are counted if the values for the event are not all NA
    table2_event = ! dplyr::if_all(ends_with("_table2"), is.na),
    shared_event = table1_event & table2_event
  )
  
  return(joined_table)

}

# merge tables by event_type", "pos_id", and "label" only (consider event the same if position is the same regardless of gene)
join_tables_no_genes <- function(table1, table2) {
  
  # create the merged table of PSI tables to compare
  joined_table <- dplyr::full_join(
  table1,
  table2,
  by = c("event_type", "pos_id", "label"),
  # label PSI values by table they came from
  suffix = c("_table1", "_table2")) |>
  # categorize events by whether they are in the v1 or rep2 tables
  dplyr::mutate(
    # Count events in one table or other if the values are not all NA
    table1_event = ! dplyr::if_all(ends_with("_table1"), is.na),
    # rep2 events are counted if the values for the event are not all NA
    table2_event = ! dplyr::if_all(ends_with("_table2"), is.na),
    shared_event = table1_event & table2_event
  )
  
  return(joined_table)

}

# Summarize events shared and unshared between tables
summarize_shared_events <- function(joined_table) {
  # calculate fraction of events only present in either table and fraction events that are shared
  shared_events_summary <- joined_table |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    total = dplyr::n(),
    shared_count = sum(shared_event),
    table1_only_count = sum(table1_event) - shared_count,
    table2_only_count = sum(table2_event) - shared_count,
    shared_percent = shared_count / total * 100,
    table1_only_percent = table1_only_count / total * 100,
    table2_only_percent = table2_only_count / total * 100,
  )
  
  # return summary table
  return(shared_events_summary)
}

```

## Directories and files

```{r directories_and_files}
## directories ##
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)
# define the data directories
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")
# these contain zipped psi files
target_pilot_workflow_results_dir <- file.path(exploration_dir, "v1_workflow_target_pilot_results", "psi")
rep2_target_pilot_results_dir <- file.path(exploration_dir, "v1_target_pilot_rep2", "psi")

#output dir for long tables
target_pilot_output <- file.path(exploration_dir, "v1_psi_tables")

## combined / canonical shiba run results
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
# directory of shiba results produced from the same shiba run
combined_dir <- file.path(target_pilot_dir, "shiba_combined", "results")
# psi table directory
# these contain unzipped psi files
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# psi table dir of deduplicated combined pilot results
# these are unzipped files
dedup_combined_results_dir <- file.path(exploration_dir, "dedup_target_pilot_combined_results")

# Check output dir exists; if not, create it
if (!dir.exists(target_pilot_output)) {
  dir.create(target_pilot_output)
}

## files ##
# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
workflow_psi_files <- c(
  se = "PSI_SE.txt.gz",
  afe = "PSI_AFE.txt.gz",
  ale = "PSI_ALE.txt.gz",
  five = "PSI_FIVE.txt.gz",
  three = "PSI_THREE.txt.gz",
  mse = "PSI_MSE.txt.gz",
  mxe = "PSI_MXE.txt.gz",
  ri = "PSI_RI.txt.gz"
)

psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of combined run shiba results
combined_psi_paths <- construct_psi_paths(combined_splice_results_dir, psi_files)

# psi table paths of psi.py run on deduplicated combined run junctions.bed
dedup_comb_psi_paths <- construct_psi_paths(dedup_combined_results_dir, psi_files)

# psi table paths of v1 workflow on target pilot samples
# these psi tables are zipped so can use the 'v1_psi_files' names
rep1_psi_paths <- construct_psi_paths(target_pilot_workflow_results_dir, workflow_psi_files)
rep2_psi_paths <- construct_psi_paths(rep2_target_pilot_results_dir, workflow_psi_files)

```

Read in files

```{r read_in_files}
# read merged splice table from v1 workflow on the TARGET pilot samples
rep1_psi_table <- read_psi_tables(rep1_psi_paths)
rep2_psi_table <- read_psi_tables(rep2_psi_paths)

# read in 'negative control' canonical shiba run tables
combined_psi_table <- read_psi_tables(combined_psi_paths)
dedup_combined_psi_table <- read_psi_tables(dedup_comb_psi_paths)
```

## Compare PSI tables

### Combined vs. combined_dedup
Are these PSI tables the same? 

#### Compare dimensions

```{r combined_vs_dedup_dims}
# dimensions of combined PSI table
dim(combined_psi_table)
# dimensions of deduplicated combined PSI table
dim(dedup_combined_psi_table)
```

Dimensions between tables are identical.

#### Check whether combined and deduplicated combined tables are identical

```{r}
identical(combined_psi_table, dedup_combined_psi_table)
```

`combined` and `deduplicated_combined` results are identical (duplicated junctions in the canonical junctions.bed file do not alter PSI calculations). 
Moving forward, the combined table is compared to the other methods.

### Treehouse Compendium workflow rep1 vs. Combined table

#### Compare dimensions
```{r rep1_vs_combined_dim}
dim(rep1_psi_table)
dim(combined_psi_table)
```

There are ~400 more unique splice events (rows) in the combined table, compared to the rep1 table;   
the canonical Shiba and Treehouse Compendium methods produce different results.

#### Check fraction of splice events that are unshared between methods

First, quantify differences present in each table.
To compare differences, rep1 and combined tables are merged on "event_type", "pos_id", "gene_id", and "label" columns.
(data tables are joined so each row corresponds to a splice event, defined by the combination of event_type, pos_id, gene_id, and label columns)
Columns correspond to PSI values for a sample and are suffixed by whether they appear in the rep1 table, combined table, or both.
-1 PSI values indicate missing PSI values from low junction readsl. 
An NA indicates that a splice event is not shared between the tables (splice event is only detected in one method).

```{r rep1_vs_combined_frac_unshared}
#| label: tbl-rep1_vs_combined_event_comparison
#| tbl-cap: "Percentage of events only in the rep1 (table1) or combined (table2) PSI tables"

# merge rep1 and combined tables and label columns by whether the event is shared between tables
rep1_vs_combined_table <- join_tables_incl_genes(rep1_psi_table, combined_psi_table)

# summarize table
summary_with_genes_rep1_vs_combined <- summarize_shared_events(rep1_vs_combined_table)

summary_with_genes_rep1_vs_combined |> 
  # no need to look at event types that are all the same
  dplyr::filter(
    shared_percent != 100
  ) |>
  dplyr::select(
  event_type,
  label,
  total,
  shared_percent,
  table1_only_percent,
  table2_only_percent)


```
A small fraction of splice events are still only found in the `rep1` or `combined` table.
Except for retained intron events, unshared events are all in annotated event types.
Retained introns are the event type category with the highest fraction of unshared events.
However, this is expected based on how exon-intron junctions (which are used to calculate these event types' PSI values) are handled in the compendium workflow.
Moving forward, I will focus on non-RI event types.

Spot-check what examples of unshared event IDs look like
```{r rep1_vs_combined_unshared_event_examples}
#| label: tbl-rep1_vs_combined_unshared_events
#| tbl-cap: "Example features of splice events not shared between rep1 and combined tables"

rep1_vs_combined_table |>
  dplyr::filter(
      event_type == "se",
      shared_event == FALSE
    ) |>
  # arrange by pos_id
  dplyr::arrange(pos_id) |>
  # only print PSI columns of two samples for ease of viewing
  dplyr::select(
    event_type, 
    pos_id, 
    gene_id, 
    label,
    SRR1559043_PSI_table1,
    SRR1559044_PSI_table1
  ) |>
  # print head
  head()

```

Many position IDs that are not shared between tables have the same position ID (see top two rows of table) but different gene IDs.

If we exclude gene ID from the position IDs, what is the fraction of events that remain unshared?
```{r no_gene_id_rep1_vs_combined_frac_unshared}
#| label: tbl-no_gene_id_rep1_vs_combined_event_comparison
#| tbl-cap: "Percentage of events only in the rep1 (table1) or combined (table2) PSI tables if gene_id is excluded from the definition of a unique event ID"

# merge rep1 and combined tables and label columns by whether the event is shared between tables
nogene_rep1_vs_combined_table <- join_tables_no_genes(rep1_psi_table, combined_psi_table)

# summarize table
summary_without_genes_rep1_vs_combined <- summarize_shared_events(nogene_rep1_vs_combined_table)

summary_without_genes_rep1_vs_combined |> 
  # no need to look at event types that are all the same
  dplyr::filter(
    shared_percent != 100
  ) |>
  dplyr::select(
  event_type,
  label,
  total,
  shared_percent,
  table1_only_percent,
  table2_only_percent)

```

If gene ID is excluded from the definition of a unique splice event, only retained intron event types are unshared between tables,
at a slightly lower fraction compared to the fraction of retained introns unshared when gene_id is included in the definition of a unique splice event.

#### Do position IDs that are shared between tables (including same gene ID) have the same PSI values?

```{r rep1_vs_combined_pos_ids_identical}
# First check PSI values between pos_ids that are shared in all aspects (including gene_id)

# extract position ids of events that are shared between tables
shared_ids_rep1_vs_combined <- rep1_vs_combined_table |>
  dplyr::filter(
    shared_event == TRUE,
    # exclude retained intron event types
    event_type != "ri"
    ) |>
  dplyr::pull(pos_id)

# filter tables for events in shared tables
shared_only_rep1_table <- rep1_psi_table |>
  dplyr::filter(pos_id %in% shared_ids_rep1_vs_combined) |>
  # filter only for fields to compare (pos_id, psi values of samples)
  dplyr::select(pos_id, ends_with("_PSI")) |>
  # sort by pos_id so order is the same
  dplyr::arrange(pos_id)

shared_only_combined_table <- combined_psi_table |>
  dplyr::filter(pos_id %in% shared_ids_rep1_vs_combined) |>
  dplyr::select(pos_id, ends_with("_PSI")) |>
  # sort by pos_id so order is the same
  dplyr::arrange(pos_id)

all.equal(shared_only_rep1_table, shared_only_combined_table)

```
PSI values are identical in events where pos_id, event_type, gene_id, and label are the same.


