---
title: Compare non-RI events coordinates between merged and combined Shiba runs
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Background

Junctions files differ in intron-exon junctions. 

GTFs differ, but we suspect the source of difference to come from the order in which files were passed into stringtie --merge and the use of multithreading.

We suspect that the junctions difference will only impact retained intron events.

If so, we would exclude this event type from the compendium.

If there are differences in other event types in the event coordinate files, we would need to change our merged compendium pipeline.

## Analysis outline

First pass: Just check if position IDs are all identical in all event types.
If not identical, check what event types are impacted by differences - is it just one or two we can easily exclude from the compendium?

## Setup

### Define functions
```{r define_functions}

# read in list of splice event coordinate files and bind them into one dataframe
read_in_events <- function(file_paths) {
event_coords <- file_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  })

# combine event coordinates into one table, labeled by event type 
all_events_table <- dplyr::bind_rows(event_coords, .id = "event_type") |>
  # drop event_id column that assigns coordinates IDs like "SE_1" "SE_2"
  # I suspect how these positions are numbered may be arbitrary and impacted by the order in which events are defined
  # in other words I don't care if these IDs are different, so long as the pos_id are the same
  dplyr::select(! event_id)
}

```


### Read in file paths and files

```{r define_paths_and_files}
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

## combined shiba run results on target pilot samples ##
# parent directories
exploration_dir <- file.path(repo_root, "exploration")
# merged table eval dir
exploration_eval_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")

# shiba results dir
# target pilot combined shiba results dir
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
combined_results_dir <- file.path(target_pilot_dir, "shiba_combined")
# events dir
combined_events_dir <- file.path(combined_results_dir, "events")

## merged shiba run results on target pilot samples ##
# shiba results dir
merged_results_dir <- file.path(repo_root, "results", "merged_shiba", "target_pilot")
# events dir
merged_events_dir <- file.path(merged_results_dir, "events")

### Files ###

# list of events coordinates files to compare between the merged and combined methods
event_files <- c(
  se = "EVENT_SE.txt",
  afe = "EVENT_AFE.txt",
  ale = "EVENT_ALE.txt",
  five = "EVENT_FIVE.txt",
  three = "EVENT_THREE.txt",
  mse = "EVENT_MSE.txt",
  mxe = "EVENT_MXE.txt",
  ri = "EVENT_RI.txt"
)

# construct events coordinate file paths of shiba results for combined run
combined_events_paths <-file.path(combined_events_dir, event_files)
names(combined_events_paths) <- names(event_files)

merged_events_paths <-file.path(merged_events_dir, event_files)
names(merged_events_paths) <- names(event_files)
```

Read in files 
```{r read_in_files}
# read in combined splice event coordinates into one table
combined_event_coords_df <- read_in_events(combined_events_paths)

# read in merged splice event coordinates into one table
merged_event_coords_df <- read_in_events(merged_events_paths)

```

## Check differences
first check: see what position IDs differ and how many IDs are only found in each dataframe

```{r check_pos_ids}
combined_only_pos_ids <- setdiff(combined_event_coords_df$pos_id, merged_event_coords_df$pos_id)
# count how many IDs are only in the combined file
length(combined_only_pos_ids)
paste0(length(combined_only_pos_ids) / length(combined_event_coords_df$pos_id) * 100, "% of position IDs are only in the combined events tables")

merged_only_pos_ids <- setdiff(merged_event_coords_df$pos_id, combined_event_coords_df$pos_id)
length(merged_only_pos_ids)
paste0(length(merged_only_pos_ids) / length(merged_event_coords_df$pos_id) * 100, "% of position IDs are only in the merged events tables")
```

Check what event types these different position IDs impact 
```{r}
# filter dataframes for events only in these position IDs

events_only_in_combined_df <- combined_event_coords_df |>
  dplyr::filter(pos_id %in% combined_only_pos_ids)

events_only_in_merged_df <- merged_event_coords_df |>
  dplyr::filter(pos_id %in% merged_only_pos_ids)

```

summarize number of event types with position IDs only in combined DF
```{r}
events_only_in_combined_df |>
  dplyr::summarise(.by = event_type,
                   n = dplyr::n())
```
summarize number of event types with position IDs only in merged df
```{r}
events_only_in_merged_df |>
  dplyr::summarise(.by = event_type,
                   n = dplyr::n())
```

Although we expected the retained introns to be the only event type impacted by how junctions.bed is generated in the merged method, differences in all event types' position IDs are present.
Additionally, there are more AFE and ALE event types impacted by these differences than retained intron events for both merged and combined sets of event coordinates.

For the event coordinate files, the only input file is the merged GTF. 
So these differences would come from the differences in GTF input file order and multithreading in the stringtie --merge step.
In other words, merging GTFs on different computing systems does alter the coordinates of splice events that are detected. 

Do we want to consider re-running the pilot Shiba "combined" and "merged" runs on mustard with the same file order and on only one thread?