---
title: Compare merged PSI table from TARGET pilot samples
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  use_cache: TRUE
---

## Introduction

Treehouse Compendium version 1 has been run on the TARGET pilot samples.
This workflow includes generating GTFs from alignments, merging GTFs of all samples, merging separately-produced junction bedfiles for each sample, and running Shiba on the merged GTF and junction file as inputs to produce event coordinate and PSI files.

To understand what differences remain between PSI results generated from Treehouse Compendium v1 and the canonical Shiba run, this notebook compares what percentage of events remain that can only be found in the workflow and the canonical Shiba method.

Below are names used to refer to different workflow runs on TARGET pilot samples:

### PSI tables produced using canonical (unaltered) Shiba scripts

**Combined**: Refers to PSI results generated from the unaltered Shiba scripts, using `shiba.py`:

```         
shiba.py -p 15 exploration/merging-psi-tables/target_pilot/shiba_config.yaml
```

There is a bug in Shiba v0.8.1 that causes some `junctions.bed` position IDs (coordinates) to be duplicated.
To check if duplicated junctions impact PSI calculation, I also compare the PSI tables below:

**Combined_dedup**: PSI values generated from running only the Shiba `PSI.py` script on the following inputs and command:

- event coordinates files generated from the Combined (canonical Shiba) method above

- `junctions.bed` files generated from Combined, then duplicated with the deduplication commands used in the v1 snakemake workflow in the `merge_jucntions` rule

```         
python $CONDA_PREFIX/share/shiba-0.8.1-0/src/psi.py -m \
10 -p \
30 -v \
--onlypsi \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/junctions/deduplicated_junctions.bed \
/private/groups/treehouse/working-projects/celiang/splicing-compendium/exploration/merging-psi-tables/target_pilot/shiba_combined/events \
v1_dedup_target_pilot_combined
```

### PSI table results produced by running Treehouse Compendium v1:

Two replicates of Treehosue Compendium v1 workflow were run to check whether repeated runs of the workflow (same command, different output directory) produce different PSI tables.

**rep1**: Refers to PSI results generated from the most recent compendium v1 workflow method

**rep2**: Refers to PSI results generated from the most recent compendium v1 workflow method, generated with the same commands and parameters as rep1 but written in a different output directory.

### Comparisons in notebook

Are combined and deduplicated results identical?

- combined vs. combined_dedup

Are there different events in Compendium vs. canonical Shiba methods?

- rep1 vs. combined

Are there differences in PSI tables if compendium workflow is run twice (are two replicate runs different?)

- rep1 vs. rep2

Are GTFs generated from the same sample in rep1 and rep2 identical?

For all PSI tables loaded in, NA values assigned by Shiba within each method listed above are replaced with -1, to distinguish NAs from splice events with low coverage apart from NAs from splice events that are not matched between each method.

## Set up

### Define functions

```{r functions}

## Reading in files tables ##

# read in PSI tables

# construct paths to psi tables
construct_psi_paths <- function(psi_dir, psi_file_list) {
  # construct paths to individual psi files
  psi_paths <-file.path(psi_dir, psi_file_list)
  # name each path according to event type
  names(psi_paths) <- names(psi_file_list)
  
  # returned the named vector of psi paths
  return(psi_paths)
}

# read in and merge splice tables
read_psi_tables <- function(psi_paths) {
  psi_results <- psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      # convert PSI values to numeric and replace NA values from shiba representing low coverage with -1
      dplyr::mutate(across(contains("_PSI"), \(x) tidyr::replace_na(as.numeric(x), -1)))
  })

# combine PSI values of samples run together into one dataframe to compare against other PSI tables
psi_table <- purrr::list_rbind(psi_results, names_to = "event_type")

# return the psi table
return(psi_table)
}

# read in event tables
read_in_events <- function(file_paths) {
event_coords <- file_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  })
}

## Manipulating PSI tables for comparison ##

# merge tables by all identifying fields for a splice event ID (event_type", "pos_id", "gene_id", "label")
join_tables_incl_genes <- function(table1, table2) {
  
  # create the merged table of PSI tables to compare
  joined_table <- dplyr::full_join(
  table1,
  table2,
  by = c("event_type", "pos_id", "gene_id", "label"),
  # label PSI values by table they came from
  suffix = c("_table1", "_table2")) |>
  # categorize events by whether they are in the v1 or rep2 tables
  dplyr::mutate(
    # Count events in one table or other if the values are not all NA
    table1_event = ! dplyr::if_all(ends_with("_table1"), is.na),
    # rep2 events are counted if the values for the event are not all NA
    table2_event = ! dplyr::if_all(ends_with("_table2"), is.na),
    shared_event = table1_event & table2_event
  )
  
  return(joined_table)

}

# merge tables by event_type", "pos_id", and "label" only (consider event the same if position is the same regardless of gene)
join_tables_no_genes <- function(table1, table2) {
  
  # create the merged table of PSI tables to compare
  joined_table <- dplyr::full_join(
  table1,
  table2,
  by = c("event_type", "pos_id", "label"),
  # label PSI values by table they came from
  suffix = c("_table1", "_table2")) |>
  # categorize events by whether they are in the v1 or rep2 tables
  dplyr::mutate(
    # Count events in one table or other if the values are not all NA
    table1_event = ! dplyr::if_all(ends_with("_table1"), is.na),
    # rep2 events are counted if the values for the event are not all NA
    table2_event = ! dplyr::if_all(ends_with("_table2"), is.na),
    shared_event = table1_event & table2_event
  )
  
  return(joined_table)

}

# Summarize events shared and unshared between tables
summarize_shared_events <- function(joined_table) {
  # calculate fraction of events only present in either table and fraction events that are shared
  shared_events_summary <- joined_table |> dplyr::summarise(
    .by = c(event_type, label),
    # count number of events in each event type and annotation category
    total = dplyr::n(),
    shared_count = sum(shared_event),
    table1_only_count = sum(table1_event) - shared_count,
    table2_only_count = sum(table2_event) - shared_count,
    shared_percent = shared_count / total * 100,
    table1_only_percent = table1_only_count / total * 100,
    table2_only_percent = table2_only_count / total * 100,
  )
  
  # return summary table
  return(shared_events_summary)
}

```

## Directories and files

```{r directories_and_files}
## directories ##
# find the root-level repo directory so we can access the other files
repo_root <- rprojroot::find_root(rprojroot::is_git_root)

# define the data directories
exploration_dir <- file.path(repo_root, "exploration")
merge_exploration_dir <- file.path(exploration_dir, "merged-method-target-pilot-eval")
# directories of rep1 and rep2 compendium Shiba results
target_pilot_workflow_results_dir <- file.path(exploration_dir, "withlogs_v1_pilot_rep1")
rep2_target_pilot_results_dir <- file.path(exploration_dir, "withlogs_v1_pilot_rep2")
# directory of GTFs produced by rep1 and rep2 of compendium workflow run
gtf_dir <- file.path(exploration_dir, "pilot_gtfs")
# these contain zipped psi files
rep1_psi_dir <- file.path(target_pilot_workflow_results_dir, "psi")
rep2_psi_dir <- file.path(rep2_target_pilot_results_dir, "psi")


## combined / canonical shiba run results
target_pilot_dir <- file.path(exploration_dir, "merging-psi-tables", "target_pilot")
# directory of shiba results produced from the same shiba run
combined_dir <- file.path(target_pilot_dir, "shiba_combined", "results")
# psi table directory
# these contain unzipped psi files
combined_splice_results_dir <- file.path(combined_dir, "splicing")

# psi table dir of deduplicated combined pilot results
# these are unzipped files
dedup_combined_results_dir <- file.path(exploration_dir, "dedup_target_pilot_combined_results")

## files ##

# PSI tables

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
workflow_psi_files <- c(
  se = "PSI_SE.txt.gz",
  afe = "PSI_AFE.txt.gz",
  ale = "PSI_ALE.txt.gz",
  five = "PSI_FIVE.txt.gz",
  three = "PSI_THREE.txt.gz",
  mse = "PSI_MSE.txt.gz",
  mxe = "PSI_MXE.txt.gz",
  ri = "PSI_RI.txt.gz"
)

psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of combined run shiba results
combined_psi_paths <- construct_psi_paths(combined_splice_results_dir, psi_files)

# psi table paths of psi.py run on deduplicated combined run junctions.bed
dedup_comb_psi_paths <- construct_psi_paths(dedup_combined_results_dir, psi_files)

# psi table paths of v1 workflow on target pilot samples
# these psi tables are zipped so can use the 'v1_psi_files' names
rep1_psi_paths <- construct_psi_paths(rep1_psi_dir, workflow_psi_files)
rep2_psi_paths <- construct_psi_paths(rep2_psi_dir, workflow_psi_files)

# GTFs

# GTF produced by rep 1 compendium TARGET pilot run
rep1_gtf_path <- file.path(gtf_dir, "rep1", "SRR1559043.gtf")
# GTF produced by rep 2 compendium TARGET pilot run
rep2_gtf_path <- file.path(gtf_dir, "rep2", "SRR1559043.gtf")

# event files

# rep1 event files path lists
rep1_events_dir <- file.path(target_pilot_workflow_results_dir, "events")
rep2_events_dir <- file.path(rep2_target_pilot_results_dir, "events")

# define list of event file paths
event_files <- c(
  se = "EVENT_SE.txt",
  afe = "EVENT_AFE.txt",
  ale = "EVENT_ALE.txt",
  five = "EVENT_FIVE.txt",
  three = "EVENT_THREE.txt",
  mse = "EVENT_MSE.txt",
  mxe = "EVENT_MXE.txt",
  ri = "EVENT_RI.txt"
)

```

Read in files

```{r read_in_files}

# PSI tables

# read merged splice table from v1 workflow on the TARGET pilot samples
rep1_psi_table <- read_psi_tables(rep1_psi_paths)
rep2_psi_table <- read_psi_tables(rep2_psi_paths)

# read in 'negative control' canonical shiba run tables
combined_psi_table <- read_psi_tables(combined_psi_paths)
dedup_combined_psi_table <- read_psi_tables(dedup_comb_psi_paths)

# GTFs

# read in GTFs from compendium TARGET runs
rep1_gtf_table <- rtracklayer::import(rep1_gtf_path) |>
  # convert to dataframe for ease of comparison
  as.data.frame(header = FALSE, sep = '\t')

# read in GTFs from compendium TARGET runs
rep2_gtf_table <- rtracklayer::import(rep2_gtf_path) |>
  # convert to dataframe for ease of comparison
  as.data.frame(header = FALSE, sep = '\t')

# event files

# construct event file paths
events_rep1_paths <- file.path(rep1_events_dir, event_files)
names(events_rep1_paths) <- names(event_files)
events_rep2_paths <- file.path(rep2_events_dir, event_files)
names(events_rep2_paths) <- names(event_files)

# read in events
events_rep1 <- read_in_events(events_rep1_paths)
events_rep2 <- read_in_events(events_rep2_paths)

```

## Compare PSI tables

### Combined vs. combined_dedup

Are these PSI tables the same?

#### Compare dimensions

```{r combined_vs_dedup_dims}
# dimensions of combined PSI table
dim(combined_psi_table)
# dimensions of deduplicated combined PSI table
dim(dedup_combined_psi_table)
```

Dimensions between tables are identical.

#### Check whether combined and deduplicated combined tables are identical

```{r}
identical(combined_psi_table, dedup_combined_psi_table)
```

`combined` and `deduplicated_combined` results are identical (duplicated junctions in the canonical junctions.bed file do not alter PSI calculations).
Moving forward, the combined table is compared to the other methods.

### Treehouse Compendium workflow rep1 vs. Combined table

#### Compare dimensions

```{r rep1_vs_combined_dim}
dim(rep1_psi_table)
dim(combined_psi_table)
```

There are \~400 more unique splice events (rows) in the combined table, compared to the rep1 table;\
the canonical Shiba and Treehouse Compendium methods produce different results.

#### Check fraction of splice events that are unshared between methods

First, quantify differences present in each table.

To compare differences, rep1 and combined tables are merged on "event_type", "pos_id", "gene_id", and "label" columns.
(data tables are joined so each row corresponds to a splice event, defined by the combination of event_type, pos_id, gene_id, and label columns) Columns correspond to PSI values for a sample and are suffixed by whether they appear in the rep1 table, combined table, or both.

`-1` PSI values indicate missing PSI values from low junction reads.
An `NA` indicates that a splice event is not shared between the tables (splice event is only detected in one method).

```{r rep1_vs_combined_frac_unshared}
#| label: tbl-rep1_vs_combined_event_comparison
#| tbl-cap: "Percentage of events only in the rep1 (table1) or combined (table2) PSI tables"

# merge rep1 and combined tables and label columns by whether the event is shared between tables
rep1_vs_combined_table <- join_tables_incl_genes(rep1_psi_table, combined_psi_table)

# summarize table
summary_with_genes_rep1_vs_combined <- summarize_shared_events(rep1_vs_combined_table)

summary_with_genes_rep1_vs_combined |> 
  # no need to look at event types that are all the same
  dplyr::filter(
    shared_percent != 100
  ) |>
  dplyr::select(
    event_type,
    label,
    total,
    shared_percent,
    table1_only_percent,
    table2_only_percent
  )


```

A small fraction of splice events are still only found in the `rep1` or `combined` table.
Except for retained intron events, unshared events are all in annotated event types.
Retained introns are the event type category with the highest fraction of unshared events.
However, this is expected based on how exon-intron junctions (which are used to calculate these event types' PSI values) are handled in the compendium workflow.
Moving forward, I will focus on non-RI event types.

If we exclude gene ID from the position IDs, what is the fraction of events that remain unshared?

```{r no_gene_id_rep1_vs_combined_frac_unshared}
#| label: tbl-no_gene_id_rep1_vs_combined_event_comparison
#| tbl-cap: "Percentage of events only in the rep1 (table1) or combined (table2) PSI tables if gene_id is excluded from the definition of a unique event ID"

# merge rep1 and combined tables and label columns by whether the event is shared between tables
nogene_rep1_vs_combined_table <- join_tables_no_genes(rep1_psi_table, combined_psi_table)

# summarize table
summary_without_genes_rep1_vs_combined <- summarize_shared_events(nogene_rep1_vs_combined_table)

summary_without_genes_rep1_vs_combined |> 
  # no need to look at event types that are all the same
  dplyr::filter(
    shared_percent != 100
  ) |>
  dplyr::select(
    event_type,
    label,
    total,
    shared_percent,
    table1_only_percent,
    table2_only_percent
  )

```

If gene ID is excluded from the definition of a unique splice event, only retained intron event types are unshared between tables, at a slightly lower fraction compared to the fraction of retained introns unshared when gene_id is included in the definition of a unique splice event.

#### Do position IDs that are shared between tables (including same gene ID) have the same PSI values?

```{r rep1_vs_combined_pos_ids_identical}
# First check PSI values between pos_ids that are shared in all aspects (including gene_id)

# extract position ids of events that are shared between tables
shared_ids_rep1_vs_combined <- rep1_vs_combined_table |>
  dplyr::filter(
    shared_event == TRUE,
    # exclude retained intron event types
    event_type != "ri"
    ) |>
  dplyr::pull(pos_id)

# filter tables for events in shared tables
shared_only_rep1_table <- rep1_psi_table |>
  dplyr::filter(pos_id %in% shared_ids_rep1_vs_combined) |>
  # filter only for fields to compare (pos_id, psi values of samples)
  dplyr::select(pos_id, ends_with("_PSI")) |>
  # sort by pos_id so order is the same
  dplyr::arrange(pos_id)

shared_only_combined_table <- combined_psi_table |>
  dplyr::filter(pos_id %in% shared_ids_rep1_vs_combined) |>
  dplyr::select(pos_id, ends_with("_PSI")) |>
  # sort by pos_id so order is the same
  dplyr::arrange(pos_id)

all.equal(shared_only_rep1_table, shared_only_combined_table)

```

PSI values are identical in events where pos_id, event_type, gene_id, and label are the same.

### Rep1 vs. rep2

Because all differences in the PSI tables stem from pos_ids being assigned different genes, I suspect the difference comes from the GTF generation or merge steps in the workflow.

To check this, see if the same kind of differences (same pos_ids present but assigned to different genes) are present in PSI tables generated from identical Compendium workflow runs (rep1 and rep2).

#### Check fraction of splice events that are unshared between methods

```{r rep1_vs_rep2_frac_unshared}
#| label: tbl-rep1_vs_rep2_event_comparison
#| tbl-cap: "Percentage of events only in the rep1 (table1) or rep2 (table2) PSI tables"

# merge rep1 and combined tables and label columns by whether the event is shared between tables
rep1_vs_rep2_table <- join_tables_incl_genes(rep1_psi_table, rep2_psi_table)

# summarize table
summary_with_genes_rep1_vs_rep2 <- summarize_shared_events(rep1_vs_rep2_table)

summary_with_genes_rep1_vs_rep2 |> 
  # no need to look at event types that are all the same
  dplyr::filter(
    shared_percent != 100
  ) |>
  dplyr::select(
  event_type,
  label,
  total,
  shared_percent,
  table1_only_percent,
  table2_only_percent)

```

Even in PSI tables generated from the same workflow commands, a similar fraction of unshared splice events are present in the tables.


Between rep1 and rep2, events that are unshared also appear to be events with the same position IDs, but assigned to different genes

Check if these differences are all from `pos_ids` being assigned to different genes

```{r no_gene_id_rep1_vs_rep2_frac_unshared}
#| label: tbl-no_gene_id_rep1_vs_rep2_event_comparison
#| tbl-cap: "Percentage of events only in the rep1 (table1) or rep2 (table2) PSI tables if gene_id is excluded from the definition of a unique event ID"

# merge rep1 and rep2 tables and label columns by whether the event is shared between tables
nogene_rep1_vs_rep2_table <- join_tables_no_genes(rep1_psi_table, rep2_psi_table)

# summarize table
summary_without_genes_rep1_vs_rep2 <- summarize_shared_events(nogene_rep1_vs_rep2_table)

summary_without_genes_rep1_vs_rep2 |> 
  # no need to look at event types that are all the same
  dplyr::filter(
    shared_percent != 100
  ) |>
  dplyr::select(
  event_type,
  label,
  total,
  shared_percent,
  table1_only_percent,
  table2_only_percent)

```

Similar to @tbl-no_gene_id_rep1_vs_combined_event_comparison, all event types except for retained introns become shared if `gene_id` is dropped from the definition of a unique event.

#### Do position IDs that are shared between tables (including same gene ID) have the same PSI values?

```{r rep1_vs_rep2_pos_ids_identical}
# First check PSI values between pos_ids that are shared in all aspects (including gene_id)

# extract position ids of events that are shared between tables
shared_ids_rep1_vs_rep2 <- rep1_vs_rep2_table |>
  dplyr::filter(
    shared_event == TRUE,
    # exclude retained intron event types
    event_type != "ri"
    ) |>
  dplyr::pull(pos_id)

# filter tables for events in shared tables
shared_only_rep1_table <- rep1_psi_table |>
  dplyr::filter(pos_id %in% shared_ids_rep1_vs_rep2) |>
  # filter only for fields to compare (pos_id, psi values of samples)
  dplyr::select(pos_id, ends_with("_PSI")) |>
  # sort by pos_id so order is the same
  dplyr::arrange(pos_id)

shared_only_rep2_table <- rep2_psi_table |>
  dplyr::filter(pos_id %in% shared_ids_rep1_vs_rep2) |>
  dplyr::select(pos_id, ends_with("_PSI")) |>
  # sort by pos_id so order is the same
  dplyr::arrange(pos_id)

all.equal(shared_only_rep1_table, shared_only_rep2_table)

```

PSI values are identical in events where pos_id, event_type, gene_id, and label are the same.

## Compare GTFs produced from rep1 and rep2 of compendium workflow runs

Since differences in the PSI tables from the `rep1 vs. combined` and `rep1 vs. rep2` comparisons stem from positions being assigned to different gene IDs in annotated events, first check what GTF entries for features in an example unshared pos_id look like.
Check features corresponding to the first pos_id in @tbl-rep1_vs_rep2_unshared_events: `SE@chr10@100523977-100524139@100516960-100526399`

This position corresponds to `ENSG00000075826.17` in rep1, but `ENSG00000255339.6` in rep2

```{r}
test_rep1 <- rep1_gtf_table |>
  dplyr::filter(
    ref_gene_id %in% c("ENSG00000075826.17", "ENSG00000255339.6")
  )

test_rep2 <- rep2_gtf_table |>
  dplyr::filter(
    ref_gene_id %in% c("ENSG00000075826.17", "ENSG00000255339.6")
  )

identical(test_rep1, test_rep2)
```

I would have expected these features in the two GTFs to be non-identical if the GTFs were the cause of pos_ids being assigned to different genes.
Contrary to this, they are identical.

Next, compare GTF `ref_gene_id` columns for a given feature (seqnames, start, end)

```{r combine_gtf_tables}
# merge gtf tables from rep1 and rep2 to compare
gtf1_gtf2_merged_df <- dplyr::full_join(
  rep1_gtf_table,
  rep2_gtf_table,
  by = c(
    "seqnames", 
    "start", 
    "end", 
    "width", 
    "cov", 
    "FPKM", 
    "TPM", 
    "exon_number", 
    "reference_id"
    ),
  # only row 45455 has a many-to-many relationship given these join columns
  relationship = "many-to-many",
  # label PSI values by table they came from
  suffix = c("_table1", "_table2")
  ) |>
  # add a column to label by whether the gene assigned to the feature is the same
  dplyr::mutate(
    gene_id_status = dplyr::case_when(
      ref_gene_id_table1 != ref_gene_id_table2 ~ "diff_gene",
      ref_gene_id_table1 == ref_gene_id_table2 ~ "same_gene",
      # if feature is unannotated it will not have gene ID 
      is.na(ref_gene_id_table1) | is.na(ref_gene_id_table2) ~ "unannotated"
    )
  )

```

If "many-to-many" relationship is not set above, the following warning is thrown:

```         
Warning in dplyr::full_join(rep1_gtf_table, rep2_gtf_table, by = c("seqnames",  :
  Detected an unexpected many-to-many relationship between `x` and `y`.
ℹ Row 45455 of `x` matches multiple rows in `y`.
ℹ Row 45455 of `y` matches multiple rows in `x`.
ℹ If a many-to-many relationship is expected, set `relationship = "many-to-many"` to silence this
  warning.
```

check row 45455 for why it has a many-to-many relationship

```{r}
# check row 45455 for why it has a many-to-many relationship
gtf1_gtf2_merged_df[45454, ]
```

I don't see any list-cols in this row, but notice that it corresponds to an unannotated feature (no ref_gene_id or ref_gene_name).
So there may be two STRG.5502.2 corresponding to unannotated events that matched to the other columns in this row.

Summarize fraction of features between `gtf1` and `gtf2` that have the same gene or have different genes

```{r compare_gtf_tables}

gtf_comparison_df <- gtf1_gtf2_merged_df |> 
  dplyr::summarise(
    .by = c(gene_id_status),
    # count number of events in each event type and annotation category
    total = dplyr::n()
  )

gtf_comparison_df

```

The GTFs here also do not differ in a way that I would have expected to make a difference in how position IDs are assigned to genes.
e.g.
I would have expected there to be nonzero GTF features assigned to different genes.

Next, I revisited the Shiba code to see what part is responsible for assigning splice event IDs to genes.

Shiba v0.8.1 [gtf2event.py](https://github.com/Sika-Zheng-Lab/Shiba/blob/v0.8.1/src/gtf2event.py) actually is what assigns event coordinates gene names, and includes code to deal with ties:

```         
for index in range(gtf_df.shape[0]):
        if gtf_gene_id[index] in gene_id_list_dic:
            l = gene_id_list_dic[gtf_gene_id[index]]
            c = collections.Counter(l)
            most_common = c.most_common()[0][0]
            gtf_df.at[index, "gene_id"] = most_common
        if gtf_gene_name[index] in gene_name_list_dic:
            l = gene_name_list_dic[gtf_gene_name[index]]
            c = collections.Counter(l)
            most_common = c.most_common()[0][0]
            gtf_df.at[index, "gene_name"] = most_common
        if ~(gtf_chr[index].startswith("chr")) and (len(gtf_chr[index]) <= 2):
            gtf_df.at[index, "chr"] = "chr" + gtf_chr[index]
```

The following explanation is from Claude -- from what I understand, multiprocessing in `gtf2events.py` results in how ties are resolved to be non-deterministic, which would explain why gene IDs in the PSI tables are different:

```         
2. The actual source of irreproducibility: multiprocessing + unstable sort + dedup
The gene_id resolved above gets carried into event rows purely by which gene's dictionary an event was discovered under:

python
event_l += [[exon, inc1, inc2, exc, strand, gene, gene_name]]

Event search itself, however, is parallelized:

python
with concurrent.futures.ProcessPoolExecutor(max_workers=num_process) as executor:
    futures = [executor.submit(se, gtf_dic_split[i]) for i in range(num_process)]
output_l = []
for future in concurrent.futures.as_completed(futures):
    output_l += future.result()
as_completed() yields futures in the order they finish, not the order they were submitted 

— this order depends on OS scheduling, per-process workload, and system load, so output_l's row order is genuinely non-deterministic across runs whenever num_process > 1.
That non-deterministic ordering then feeds directly into the dedup step used for every event type:

python
output_df = output_df.sort_values("exon")   # or "exon_a", etc.
output_df = output_df.drop_duplicates(subset = "pos_id", keep = "first")

Two things compound here:

sort_values() uses pandas' default kind='quicksort', which is not stable — it doesn't guarantee that rows with equal sort keys (e.g. the same exon coordinate) keep their original relative order.
drop_duplicates(keep="first") then keeps whichever row happens to land first after that sort.

Putting it together

If two different genes — whether genuinely overlapping loci, or exons that got resolved to different gene_ids via the majority-vote fallback — happen to produce an event with the identical pos_id string 
(same event type, chromosome, and junction coordinates), then:
Their rows arrive in output_l in whatever order the worker processes happened to finish (non-deterministic across runs).

The unstable sort doesn't reliably preserve any particular ordering for ties.
drop_duplicates(subset="pos_id", keep="first") silently keeps whichever gene's row happened to survive — which can differ from run to run.
So the same genomic position can end up labeled with different gene_id/gene_name values across two otherwise-identical runs of gtf2event.py, purely because of scheduling non-determinism in the parallel event search.

Practical implication: this only manifests when -p/--num-process is set above 1 (with a single process, there's only one future to complete, so ordering is trivially deterministic) and only for pos_ids that are genuinely ambiguous across genes — i.e., cases where two distinct genes generate an identical coordinate-based event signature.
```

I did indeed use more than one thread for these runs.
[compare_event_coords.md](https://github.com/UCSC-Treehouse/splicing-compendium/blob/main/exploration/merged-method-target-pilot-eval/compare_event_coords.md) compared similarity between event files produced by identical Shiba runs with multiple threads, but Jaccard indices were only compared using position IDs.
This notebook did not examine similarity in gene names assigned to position IDs.

## Compare event coordinates

Because the difference in PSI values stem from differences in gene IDs, compare gene IDs in event coordinates

```{r}

# lazily using jaccard index code from previous notebook
gene_id_jaccard <- purrr::map2(
  # read in lists of event coordinate dataframes to compare and iterate the two simultaneously
  events_rep1,
  events_rep2,
  # take the matching two dataframes from the input lists
  \(df1, df2){
    # obtain set of gene IDs 
    set1 <- df1$gene_id
    set2 <- df2$gene_id
    
    # calculate Jaccard index of position IDs
    length(intersect(set1, set2)) / length(union(set1, set2))
  }
  )

# print results
gene_id_jaccard

```

The event file gene IDs are not the same, supporting the hypothesis that inconsistent gene IDs come from `gtf2events.py`

Check features corresponding to the first pos_id in @tbl-rep1_vs_rep2_unshared_events: `SE@chr10@100523977-100524139@100516960-100526399`

```{r}
events_rep1$se |>
  dplyr::filter(pos_id == "SE@chr10@100523977-100524139@100516960-100526399")
```

```{r}
events_rep2$se |>
  dplyr::filter(pos_id == "SE@chr10@100523977-100524139@100516960-100526399")
```

In the event file for this position ID, we can see that the `gene_id` / `gene_name` columns are the only fields different between the two runs.

## Conclusions:

- PSI differences all stem from the same position ID being assigned to different genes

- Different gene assignment for a splice event position ID appears to come from Shiba's `gtf2event.py` code, not GTF generation (though row order of features in the GTF may play a role in how `gtf2event.py` breaks ties)

- A potential last step would be to rerun two replicates of TARGET pilot with one thread for `gtf2event.py` and compare Jaccard indices for both `pos_id` and `gene_id` between rep1 and rep2 events made with multithreading.
