---
title: Compare merged PSI table vs. Shiba PSI tables
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Set up
## Directories and files

```{r}
# find the project directory
base_dir <- here::here()

# define the data directories
# psi table comparison dir
merging_dir <- file.path(base_dir, "exploration/merging-psi-tables")

# shiba splice results directory of two samples run together
shiba_dir <- file.path(merging_dir, "shiba_results/combined_run/splicing")

# merged PSI table of two samples run separately with shiba
merged_psi_table_dir <- file.path(merging_dir, "merged_results")

# Read in file paths
# merged PSI file of two samples obtained from merge-shiba-psi-tables.qmd
merged_psi_file <- file.path(merged_psi_table_dir, "merged_psi_table.tsv")

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of shiba results of two samples run together
shiba_psi_paths <-file.path(shiba_dir, psi_files)
names(shiba_psi_paths) <- names(psi_files)

```

Read in files
```{r}
# read in manually combined splice table
merged_psi_table <- readr::read_tsv(merged_psi_file)

# read in splice results to compare against manually combined results
shiba_splice_results_paths <- shiba_psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      dplyr::mutate(across(contains("_PSI"), as.numeric))
  })

# combine psi values of samples run together into one dataframe to compare against manually combined dataframe
shiba_psi_table <- purrr::list_rbind(shiba_splice_results_paths, names_to = "event_type")
```
## Obtain dimensions of the merged and shiba tables to compare values
```{r}
paste0("Merged PSI table dimensions ", 
       dim(merged_psi_table)[1], 
       "x", 
       dim(merged_psi_table)[2])

paste0("Shiba PSI table dimensions ", 
       dim(shiba_psi_table)[1],
       "x",
       dim(shiba_psi_table)[2])
```

## Obtain splice events unique to each table

```{r}
# produce dataframe of elements in combined splice results reference dataframe but not in merged psi df
unique_shiba <- setdiff(shiba_psi_table$pos_id, merged_psi_table$pos_id)

# produce dataframe of elements in merged splice results dataframe but not in the reference df
unique_merged <- setdiff(merged_psi_table$pos_id, shiba_psi_table$pos_id)
```

## Obtain dimensions of the merged and shiba tables to compare values

```{r}
paste0("Number of total unique merged events: ", 
       length(unique_merged))

paste0("Number of total unique Shiba events: ", 
       length(unique_shiba))
```

Print the number of unique annotated vs. unannotated events in each event type for the merged table
```{r}
# filter for pos_ids unique to the shiba dataframe
unique_shiba_psi_table <- shiba_psi_table |>
  dplyr::filter(pos_id %in% unique_shiba,
                !is.na(SRR601500_PSI) & !is.na(SRR604528_PSI))

# print summary of how many unique events with values in both samples are novel vs. unannotated
unique_shiba_psi_table |> 
  dplyr::summarise(.by = c(event_type, label), n = dplyr::n())
```

Print the number of unique annotated vs. unannotated events in each event type for the merged table
```{r}
# filter for the pos_ids in the merged df that are unique
unique_merged_psi_table <- merged_psi_table |>
  dplyr::filter(pos_id %in% unique_merged,
                !is.na(SRR601500_PSI) & !is.na(SRR604528_PSI))

# print summary of how many unique events with values in both samples are novel vs unannotated
unique_merged_psi_table |> 
  dplyr::summarise(.by = c(event_type, label), n = dplyr::n())

```
For both tables, there are more annotated unique events than unannotated ones, which may be concerning.
These probably represent splice events that are annotated but only found in one of the two samples.

## Inspect individual position IDs of unique events

We are most concerned if unannotated events have altered positions in the merged table and are unmatched in the shiba table (is Shiba doing something strange to assign reads into position coordinates?)

Check for one event type. 
From the table above, RI events are the simplest to spot-check on IGV that have unannotated and annotated events with PSI values in both samples in the merged table.

```{r}
# Filter for splice events that have real PSI values in all samples; start with skipped exon events for now
unique_se_in_merged <- unique_merged_psi_table |>
  dplyr::filter(event_type == "ri",
                !is.na(SRR601500_PSI) & !is.na(SRR604528_PSI))

unique_se_in_merged
```

## Search for the closest RI events in the shiba PSI table

### Look for RI events in GL000219.1
```{r}
shiba_psi_table |>
  dplyr::filter(stringr::str_detect(pos_id, "GL000219.1@77669"))
```

There are no RI events in @GL000219.1 region in the shiba PSI table.
Instead, these reads might be assigned to ALE events (GL000219.1@77669-78842 is in three ALE events in the shiba table)

#### IGV screenshots of region 
Check on IGV - does this event look like an ALE or RI?

### Look for RI events in chr3@139355790

```{r}
shiba_psi_table |>
  dplyr::filter(stringr::str_detect(pos_id, "chr3@139355790"))
```

There are also no RI events in the chr3@139355790 region in the shiba PSI table.
Instead, these reads might be assigned to ALE events.

#### IGV screenshots of region 
Check on IGV - does this event look like an ALE or RI?

### Look for RI events in chr6@99400311
This is the only unmatched RI annotated event in the merged table.

```{r}
shiba_psi_table |>
  dplyr::filter(stringr::str_detect(pos_id, "chr6@99400311"))
```

There are no events in the Shiba PSI table with the start coordinate chr6@99400311.

Check junction counts as well or any other stats for these three events --> will the values even be kept after we run our distribution comparison?


