---
title: Compare merged PSI table vs. Shiba PSI tables
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable
editor_options:
  markdown:
    wrap: sentence
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Set up
## Directories and files

```{r}
# find the project directory
base_dir <- here::here()

# define the data directories
# splice results directory
results_dir <- file.path(base_dir, "results")

# merged shiba PSI table of two samples
merged_psi_table_dir <- file.path(results_dir, "merged_shiba_results_tables")

# merged PSI file of two samples
merged_psi_file <- file.path(merged_psi_table_dir, "merged_psi_table.tsv")

# general shiba results paths
shiba_gtex_path <- file.path(results_dir, "gtex/shiba")

# path to splice results of shiba run done on two samples at once
merged_splice_results_path <- "two_samples_together/results/splicing"

# define list of PSI results files corresponding to event types quantified by Shiba bulk analysis
psi_files <- c(
  se = "PSI_SE.txt",
  afe = "PSI_AFE.txt",
  ale = "PSI_ALE.txt",
  five = "PSI_FIVE.txt",
  three = "PSI_THREE.txt",
  mse = "PSI_MSE.txt",
  mxe = "PSI_MXE.txt",
  ri = "PSI_RI.txt"
)

# construct psi table paths of shiba results of two samples run together
psi_paths <-file.path(shiba_gtex_path, merged_splice_results_path, psi_files)
names(psi_paths) <- names(psi_files)

```

Read in files
```{r}

# read in manually combined splice table
test_psi_table <- readr::read_tsv(merged_psi_file)

# read in splice results to compare against manually combined results
splice_results <- psi_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c")) |>
      # select for columns that will be used in downstream analysis
      dplyr::select(pos_id, gene_id, label, contains("_PSI")) |>
      dplyr::mutate(across(contains("_PSI"), as.numeric))
  })

# combine psi values of samples run together into one dataframe to compare against manually combined dataframe
ref_splice_results <- purrr::list_rbind(splice_results, names_to = "event_type")
```
Use R setdiff
For annotated sites are they all the same
If differences are in the novel sites, how different are they

```{r}
# produce dataframe of elements in combined splice results reference dataframe but not in merged psi df
unique_reference <- setdiff(ref_splice_results$pos_id, test_psi_table$pos_id)

# produce dataframe of elements in merged splice results dataframe but not in the reference df
unique_test <- setdiff(test_psi_table$pos_id, ref_splice_results$pos_id)
```

1839 pos_ids unique in test df
657 pos_ids unique in ref df

Count number of unique events annotated vs. unannotated in each group
```{r}

print("Number of unique annotated and unannotated events in test table")

# filter for the pos_ids in the test df that are unique
unique_test_psi_table <- test_psi_table |>
  dplyr::filter(pos_id %in% unique_test)

# count the number of novel vs annotated events
table(unique_test_psi_table$label)

print("Number of unique annotated and unannotated events in reference table")

# filter for the pos_ids in the ref df that are unique
unique_reference_psi_table <- ref_splice_results |>
  dplyr::filter(pos_id %in% unique_reference)

# count the number of novel vs annotated events
table(unique_reference_psi_table$label)

```
Inspect individual position IDs of SE events in the merged (test) PSI table that have PSI values in both samples
```{r}
unique_se_test <- unique_test_psi_table |>
  dplyr::filter(event_type == "se",
                !is.na(SRR601500_PSI) & !is.na(SRR604528_PSI))
```

Spot-check how close pos_ids are by eye for one event type for now
```{r}
complete_unique_test_psi_table <- unique_test_psi_table |>
  dplyr::filter(
                # we're most interested in events that have PSI values in both samples
                !is.na(SRR601500_PSI) & !is.na(SRR604528_PSI), 
                )
```

There are no SE events in the pos-ids unique only to the test table.
Of the events unique only to the test table, that have values in both samples, only two are unannotated 

A lot of the unique events in the test table are AFE events in gene ENSG00000214655.11, so look for AFE events in this gene in the ref table

```{r}
afe_ref_table <- ref_splice_results |>
  dplyr::filter(gene_id == "ENSG00000214655.11",
                event_type == "afe")
```

We are more concerned by thigns that are missing in the test (merged) table  in 

There are surprisingly no AFE events in this gene in the ref table!
If we just search this gene in the ref table, there are only ALE, FIVE, THREE, and RI events

Notes:

We are most concerned if unannotated events have altered positions in the merged table and are unmatched

Check the unique unannotated events in the shiba ref table first
Are there things where it will change the position a little bit? or calls across samples are ending up in different rows?
Looking for unannotated events where neither samples have NA PSI values 

Keep in same notebook for now
Head tables (no screenshots)

New header every time you start another part of an analysis

Check junction counts as well or any other stats--> will the values even be kept after we tun our distribution comparison

dim() to get dimensions of table, show column names
How many splice event types are present in each table


