---
title: Add clinical metadata to TARGET compendium v1 sample sheet
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable 
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

This notebook adds clinical metadata to the TARGET accessions sheet. Clinical metadata is obtained from the UCSC Xena browser: https://xenabrowser.net/datapages/?hub=https://gdc.xenahubs.net:443

We plan to include the following metadata fields in the accessions sheet:

- age at dx

- race/ethnicity

- sex

- survival

- Anything with a single number or value per column so it can be computed on easily

## Setup

### Read in directories and files

```{r directories_and_files}

## Define paths to directories ##

# find the project directory
base_dir <- here::here()

# define directory paths
# accession metadata dir
metadata_dir <- file.path(base_dir, "metadata")
# clinical TARGET metadata directory
clinical_target_dir <- file.path(metadata_dir, "clinical", "target")
# filtered accessions directory
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")
# sample sheet dir of samples that were downloaded and processed
config_dir <- file.path(base_dir, "config")

## Define paths to input files ##

# full TARGET dataset SRA accessions and metadata
target_sra_path <- file.path(gtex_target_metadata_dir, "SraRunTable-TARGET.csv")
# final list of accessions that were downloaded and processed, including pilot accessions
compendium_samples_path <- file.path(config_dir, "sample_sheet.tsv")

# TARGET clinical metadata files
# Define list of TARGET clinical metadata files
target_clinical_files <- c(
  all_p1_clinical = "TARGET-ALL-P1.clinical.tsv",
  all_p2_clinical = "TARGET-ALL-P2.clinical.tsv",
  aml_clinical = "TARGET-AML.clinical.tsv",
  nbl_clinical = "TARGET-NBL.clinical.tsv",
  rt_clinical = "TARGET-RT.clinical.tsv",
  wt_clinical = "TARGET-WT.clinical.tsv"
)

target_survival_files <- c(
  all_p1_survival = "TARGET-ALL-P1.survival.tsv",
  all_p2_survival = "TARGET-ALL-P2.survival.tsv",
  aml_survival = "TARGET-AML.survival.tsv",
  nbl_survival = "TARGET-NBL.survival.tsv",
  rt_survival = "TARGET-RT.survival.tsv",
  wt_survival = "TARGET-WT.survival.tsv"
)

# construct TARGET clinical metadata paths
target_clinical_metadata_paths <-file.path(clinical_target_dir, target_clinical_files)
names(target_clinical_metadata_paths) <- names(target_clinical_files)
target_survival_metadata_paths <- file.path(clinical_target_dir, target_survival_files)
names(target_survival_files) <- names(target_survival_files)

## Output ##
# define path to output file
target_accessions_with_clinical_metadata <- file.path(gtex_target_metadata_dir, "target_accessions_clinical.tsv")
```

Read in files

```{r read_files}

## Sample sheets ##

# SRA accessions tables - target_sra "Run" column is to be associated with compendium_samples "sample" column
# target SRA table
target_sra <- readr::read_csv(target_sra_path, col_types = readr::cols(.default = "c"))
# sample sheet of what was downloaded and processed
compendium_samples <- readr::read_tsv(compendium_samples_path, col_types = readr::cols( .default = "c"))

## Clinical metadata ##

# read in metadata paths
target_clinical_metadata <- target_clinical_metadata_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  }) |> 
  purrr::list_rbind()

target_survival_metadata <- target_survival_metadata_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default="c"))
  }) |>
  purrr::list_rbind()

```

The "sample" column in the survival and clinical metadata tables have entries that look like: TARGET-10-PAMDRM-09A

The "sample" column in the metadata table correspond to the "biospecimen_repository_sample_id" column in the target accessions table, which looks like: TARGET-10-PAMDRM-09A-01D

"biospecimen_repository_sample_id" needs to be cleaned so the last substring separated by "-" is removed

Combine metadata with accessions sheet and filter target SRA metadata for those corresponding to samples that were downloaded and processed

```{r combine_metadata}

# combine target clinical and survival metadata
target_metadata <- dplyr::full_join(
  target_clinical_metadata,
  target_survival_metadata,
  by = "sample"
)

# make a list of target accessions that were downloaded and processed
target_downloaded <- compendium_samples |>
  dplyr::filter(group == "target") |>
  dplyr::pull(sample)

# filter target accessions for accessions that are in the sample sheet
target_sra_metadata <- target_sra |>
  dplyr::filter(Run %in% target_downloaded) |>
  # select for relevant columns
  dplyr::select(
    Run, biospecimen_repository, `DATASTORE filetype`, body_site, Bytes,
    study_name, `DATASTORE provider`, analyte_type, `Assay Type`, `SRA Study`,
    BioProject, BioSample, LibraryLayout, LibrarySelection, `Center Name`,
    biospecimen_repository_sample_id
  ) |>
  # derive target sample ID from sample ID 
  dplyr::mutate(
    # remove the hyphen followed by characters that are not hyphens that are at the end of the string
    target_sample_id = sub("-[^-]+$", "", biospecimen_repository_sample_id)
  )
  
# Check target metadata contents
colnames(target_metadata)

```

There are more fields in the metadata than we planned to include, so some discussion on what may or may not be useful as a group will be helpful. For now, I won't exclude any metadata and will just focus on merging the metadata with the accessions sheet

Combine target metadata with SRA sheet of samples that were downloaded

```{r combine_metadata_with_sra}

# merge target accessions with clinical metadata 
target_clinical_accessions <- target_sra_metadata |>
  dplyr::left_join(target_metadata, by = dplyr::join_by(target_sample_id == sample))

# print head to eyeball contents
target_clinical_accessions |>
  dplyr::slice_sample( n = 5) |>
  head(n = 5)

```

Not all samples have all clinical/survival metadata fields.

## Write output

```{r write_output}
# holding off on this until the code has gone through review
```
