---
title: Add clinical metadata to TARGET compendium v1 sample sheet
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable 
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

This notebook adds clinical metadata to 

## Setup

### Read in directories and files

Clinical metadata columns are:
sample	
id	
disease_type	
case_id	
submitter_id	
primary_site	
cause_of_death.demographic	
race.demographic	
gender.demographic	
ethnicity.demographic	
vital_status.demographic	
age_at_index.demographic	
days_to_birth.demographic	
age_is_obfuscated.demographic	
days_to_death.demographic	
primary_site.project	
project_id.project	
disease_type.project	
name.project	
name.program.project	entity_submitter_id.annotations	notes.annotations	submitter_id.annotations	classification.annotations	entity_id.annotations	created_datetime.annotations	annotation_id.annotations	entity_type.annotations	updated_datetime.annotations	case_id.annotations	state.annotations	category.annotations	status.annotations	case_submitter_id.annotations	tissue_or_organ_of_origin.diagnoses	age_at_diagnosis.diagnoses	morphology.diagnoses	classification_of_tumor.diagnoses	icd_10_code.diagnoses	days_to_diagnosis.diagnoses	primary_diagnosis.diagnoses	year_of_diagnosis.diagnoses	diagnosis_is_primary_disease.diagnoses	site_of_resection_or_biopsy.diagnoses	age_at_earliest_diagnosis.diagnoses.xena_derived	age_at_earliest_diagnosis_in_years.diagnoses.xena_derived	protocol_identifier.treatments.diagnoses	updated_datetime.treatments.diagnoses	treatment_id.treatments.diagnoses	submitter_id.treatments.diagnoses	state.treatments.diagnoses	treatment_or_therapy.treatments.diagnoses	created_datetime.treatments.diagnoses	sample_type_id.samples	tumor_descriptor.samples	sample_id.samples	sample_type.samples	tumor_code.samples	preservation_method.samples	freezing_method.samples	tumor_code_id.samples	oct_embedded.samples	specimen_type.samples	tissue_type.samples

sample field: TARGET-10-PAMDRM-09A	

survival metadata columns: sample	OS.time	OS	_PATIENT
sample field: TARGET-10-PANSFD-09A

```{r directories_and_files}

## Define paths to directories ##

# find the project directory
base_dir <- here::here()
# define directory paths
# accession metadata dir
metadata_dir <- file.path(base_dir, "metadata")
# clinical TARGET metadata directory
clinical_target_dir <- file.path(metadata_dir, "clinical", "target")
# filtered accessions directory
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")
# sample sheet dir of samples that were downloaded and processed
config_dir <- file.path(base_dir, "config")

## Define paths to input files ##

# TARGET accessions with dbGaP metadata
target_accessions_path <-  file.path(gtex_target_metadata_dir, "target_accessions.tsv")
# final list of accessions that were downloaded and processed
compendium_samples_path <- file.path(config_dir, "sample_sheet.tsv")

# TARGET clinical metadata files

# Define list of TARGET clinical metadata files
target_clinical_files <- c(
  all_p1_clin = "TARGET-ALL-P1.clinical.tsv",
  all_p1_surv = "TARGET-ALL-P1.survival.tsv",
  all_p2_clin = "TARGET-ALL-P2.clinical.tsv",
  all_p2_surv = "TARGET-ALL-P2.survival.tsv",
  aml_clin = "TARGET-AML.clinical.tsv",
  aml_surv = "TARGET-AML.survival.tsv",
  nbl_clin = "TARGET-NBL.clinical.tsv",
  nbl_surv = "TARGET-NBL.survival.tsv",
  rt_clin = "TARGET-RT.clinical.tsv",
  rt_surv = "TARGET-RT.survival.tsv",
  wt_clin = "TARGET-WT.clinical.tsv",
  wt_surv = "TARGET-WT.survival.tsv"
)

# construct TARGET clinical metadata paths
target_clinical_metadata_paths <-file.path(clinical_target_dir, target_clinical_files)
names(target_clinical_metadata_paths) <- names(target_clinical_files)
```

Read in files 

```{r read_files}

## Sample sheets ##

# filtered target accessions to download
target_accessions <- readr::read_tsv(target_accessions_path, col_types = readr::cols( .default = "c"))
# sample sheet of what was downloaded and processed
compendium_samples <- readr::read_tsv(compendium_samples_path, col_types = readr::cols( .default = "c"))

## Clinical metadata ##

# read in metadata paths
target_metadata <- target_clinical_metadata_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  })

# combine clinical and survival metadata 
combined_target_metadata <- purrr::list_rbind(target_metadata, names_to = "metadata_type")

```

