---
title: Add clinical metadata to TARGET compendium v1 sample sheet
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable 
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

This notebook adds clinical metadata to the TARGET accessions sheet. Clinical metadata is obtained from the UCSC Xena browser: https://xenabrowser.net/datapages/?hub=https://gdc.xenahubs.net:443

We plan to include the following metadata fields in the accessions sheet:

- age at dx

- race/ethnicity

- sex

- survival

- Anything with a single number or value per column so it can be computed on easily

## Setup

### Read in directories and files

```{r directories_and_files}

## Define paths to directories ##

# find the project directory
base_dir <- here::here()

# define directory paths
# accession metadata dir
metadata_dir <- file.path(base_dir, "metadata")
# clinical TARGET metadata directory
clinical_target_dir <- file.path(metadata_dir, "clinical", "target")
# filtered accessions directory
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")
# sample sheet dir of samples that were downloaded and processed
config_dir <- file.path(base_dir, "config")

## Define paths to input files ##

# full TARGET dataset SRA accessions and metadata
target_sra_path <- file.path(gtex_target_metadata_dir, "SraRunTable-TARGET.csv")
# final list of accessions that were downloaded and processed, including pilot accessions
compendium_samples_path <- file.path(config_dir, "sample_sheet.tsv")

# TARGET clinical metadata files
# Define list of TARGET clinical metadata files
target_clinical_files <- c(
  all_p1_clinical = "TARGET-ALL-P1.clinical.tsv",
  all_p2_clinical = "TARGET-ALL-P2.clinical.tsv",
  aml_clinical = "TARGET-AML.clinical.tsv",
  nbl_clinical = "TARGET-NBL.clinical.tsv",
  rt_clinical = "TARGET-RT.clinical.tsv",
  wt_clinical = "TARGET-WT.clinical.tsv"
)

target_survival_files <- c(
  all_p1_survival = "TARGET-ALL-P1.survival.tsv",
  all_p2_survival = "TARGET-ALL-P2.survival.tsv",
  aml_survival = "TARGET-AML.survival.tsv",
  nbl_survival = "TARGET-NBL.survival.tsv",
  rt_survival = "TARGET-RT.survival.tsv",
  wt_survival = "TARGET-WT.survival.tsv"
)

# construct TARGET clinical metadata paths
target_clinical_metadata_paths <-file.path(clinical_target_dir, target_clinical_files)
names(target_clinical_metadata_paths) <- names(target_clinical_files)
target_survival_metadata_paths <- file.path(clinical_target_dir, target_survival_files)
names(target_survival_files) <- names(target_survival_files)

## Output ##
# define path to output file
target_accessions_with_clinical_metadata <- file.path(gtex_target_metadata_dir, "target_accessions_clinical.tsv")
```

Read in files

```{r read_files}

## Sample sheets ##

# SRA accessions tables - target_sra "Run" column is to be associated with compendium_samples "sample" column
# target SRA table
target_sra <- readr::read_csv(target_sra_path, col_types = readr::cols(.default = "c"))
# sample sheet of what was downloaded and processed
compendium_samples <- readr::read_tsv(compendium_samples_path, col_types = readr::cols( .default = "c"))

## Clinical metadata ##

# read in metadata paths
target_clinical_metadata <- target_clinical_metadata_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  }) |> 
  purrr::list_rbind()

target_survival_metadata <- target_survival_metadata_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default="c"))
  }) |>
  purrr::list_rbind()

```

The "sample" column in the survival and clinical metadata tables have entries that look like: TARGET-10-PAMDRM-09A

The "sample" column in the metadata table correspond to the "biospecimen_repository_sample_id" column in the target accessions table, which looks like: TARGET-10-PAMDRM-09A-01D

"biospecimen_repository_sample_id" needs to be cleaned so the last substring separated by "-" is removed

Combine metadata with accessions sheet and filter target SRA metadata for those corresponding to samples that were downloaded and processed

```{r combine_metadata}

# combine target clinical and survival metadata
target_metadata <- dplyr::full_join(
  target_clinical_metadata,
  target_survival_metadata,
  by = "sample"
)

# make a list of target accessions that were downloaded and processed
target_downloaded <- compendium_samples |>
  dplyr::filter(group == "target") |>
  dplyr::pull(sample)

# filter target accessions for accessions that are in the sample sheet
target_sra_metadata <- target_sra |>
  dplyr::filter(Run %in% target_downloaded) |>
  # select for relevant columns
  dplyr::select(
    Run, biospecimen_repository, `DATASTORE filetype`, body_site,
    study_name, analyte_type, `Assay Type`, `SRA Study`,
    BioProject, BioSample, LibraryLayout, LibrarySelection, `Center Name`,
    biospecimen_repository_sample_id
  ) |>
  # derive target sample ID from sample ID 
  dplyr::mutate(
    # remove the hyphen followed by characters that are not hyphens that are at the end of the string
    target_sample_id = sub("-[^-]+$", "", biospecimen_repository_sample_id)
  )
  
# Check target metadata contents
colnames(target_metadata)

```

There are more fields in the metadata than we planned to include, so some discussion on what may or may not be useful as a group will be helpful. For now, I won't exclude any metadata and will just focus on merging the metadata with the accessions sheet

Combine target metadata with SRA sheet of samples that were downloaded

Columns excluded from the final dataframe due to having information that is not useful:

vital_status (values: Dead, Alive), name.program.project (value = TARGET),

Columns excluded due to having multiple values per row/ID: primary_site.project, entity.submitter_id.annotations, notes.annotations, submitter_id.annotations, classification.annotaitons, entity_id.annotations, annotation_id.annotations, entity_type.annotations,

```{r combine_metadata_with_sra}

# merge target accessions with clinical metadata 
target_clinical_accessions <- target_sra_metadata |>
  dplyr::left_join(target_metadata, by = dplyr::join_by(target_sample_id == sample)) |>
  # select for columns that are relevant to clinical metadata
  dplyr::select(
    # these columns have the same values in other columns
    # I select for the most clearly named column of these, or rename the column with the most values for clarity
    submitter_id,
    disease_type,
    age_at_diagnosis_days = age_at_diagnosis.diagnoses,
    case_id,
    age_at_earliest_diagnosis_in_years.diagnoses.xena_derived, # selected over age_at_index.demographic due to having unrounded numbers
    # the remaining columns are not repeated in the data frame and are selected for clinical relevance
    disease_type,
    submitter_id,
    primary_site,
    cause_of_death.demographic,
    race.demographic,
    gender.demographic,
    ethnicity.demographic,
    project_id.project,
    name.project,
    tissue_or_organ_of_origin.diagnoses,
    morphology.diagnoses,
    classification_of_tumor.diagnoses,
    icd_10_code.diagnoses,
    primary_diagnosis.diagnoses,
    sample_type_id.samples,
    tumor_descriptor.samples,
    sample_type.samples,
    tumor_code_id.samples,
    sample_type.samples,
    tissue_type.samples,
    inss_stage.diagnoses,
    cog_neuroblastoma_risk_group.diagnoses,
    mitosis_karyorrhexis_index.diagnoses,
    inpc_grade.diagnoses,
    pathology_detail_id.pathology_details.diagnoses,
    necrosis_percent.pathology_details.diagnoses,
    percent_tumor_nuclei.pathology_details.diagnoses,
    pediatric_kidney_staging.diagnoses,
    wilms_tumor_histologic_subtype.diagnoses,
    OS.time,
    OS,
    `_PATIENT`
  )

# print head to eyeball contents
target_clinical_accessions |>
  dplyr::slice_sample( n = 5) |>
  head(n = 5)

```

Not all samples have all clinical/survival metadata fields.

```{r}
target_clinical_accessions |>
  dplyr::select(age_at_earliest_diagnosis.diagnoses.xena_derived, age_at_diagnosis.diagnoses, age_at_index.demographic)
```

## Write output

```{r write_output}
# holding off on this until the code has gone through review
```
