---
title: Add clinical metadata to TARGET compendium v1 sample sheet
author: "Cindy Liang (celiang@ucsc.edu)"
format:
  gfm:
    df-print: kable 
editor: visual
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

## Introduction

This notebook adds clinical metadata to the TARGET accessions sheet. Clinical metadata is obtained from the [UCSC Xena browser metadata hub](https://xenabrowser.net/datapages/?hub=https://gdc.xenahubs.net:443). Metadata related to clinical attributes of the sample are split into two files for each TARGET cancer type in the linked page.

- "Survival data" - Contains the survival information of each TARGET sample within the specified cancer type (e.g. overall survival, overall survival time)

- "Survival data" - Contains the survival information of each TARGET sample within the specified cancer type (e.g. overall survival, overall survival time)

We plan to include the following metadata fields in the accessions sheet:

- Age at dx

- Race/ethnicity

- Sex

- Survival

- Sequencing center

Prior to combining survival and phenotype metadata to existing TARGET accession metadata, the survival and phenotype information is filtered to meet the above criteria.

## Setup

### Read in directories and files

```{r directories_and_files}

## Define paths to directories ##

# find the project directory
base_dir <- here::here()

# define directory paths
# accession metadata dir
metadata_dir <- file.path(base_dir, "metadata")
# clinical TARGET metadata directory
clinical_target_dir <- file.path(metadata_dir, "clinical", "target")
# filtered accessions directory
gtex_target_metadata_dir <- file.path(metadata_dir, "filter_target_gtex")
# sample sheet dir of samples that were downloaded and processed
config_dir <- file.path(base_dir, "config")

## Define paths to input files ##

# full TARGET dataset SRA accessions and metadata
target_sra_path <- file.path(gtex_target_metadata_dir, "SraRunTable-TARGET.csv")
# final list of accessions that were downloaded and processed, including pilot accessions
compendium_samples_path <- file.path(config_dir, "sample_sheet.tsv")

# TARGET clinical metadata files
# Define list of TARGET clinical metadata files
target_clinical_files <- c(
  all_p1_clinical = "TARGET-ALL-P1.clinical.tsv",
  all_p2_clinical = "TARGET-ALL-P2.clinical.tsv",
  all_p3_clinical = "TARGET-ALL-P3.clinical.tsv",
  aml_clinical = "TARGET-AML.clinical.tsv",
  nbl_clinical = "TARGET-NBL.clinical.tsv",
  rt_clinical = "TARGET-RT.clinical.tsv",
  wt_clinical = "TARGET-WT.clinical.tsv",
  ccsk_clinical = "TARGET-CCSK.clinical.tsv"
)

target_survival_files <- c(
  all_p1_survival = "TARGET-ALL-P1.survival.tsv",
  all_p2_survival = "TARGET-ALL-P2.survival.tsv",
  all_p3_survival = "TARGET-ALL-P3.survival.tsv",
  aml_survival = "TARGET-AML.survival.tsv",
  nbl_survival = "TARGET-NBL.survival.tsv",
  rt_survival = "TARGET-RT.survival.tsv",
  wt_survival = "TARGET-WT.survival.tsv",
  ccsk_clinical = "TARGET-CCSK.survival.tsv"
)

# construct TARGET clinical metadata paths
target_clinical_metadata_paths <-file.path(clinical_target_dir, target_clinical_files)
names(target_clinical_metadata_paths) <- names(target_clinical_files)
target_survival_metadata_paths <- file.path(clinical_target_dir, target_survival_files)
names(target_survival_files) <- names(target_survival_files)

## Output ##
# define path to output file
target_clinical_metadata_output <- file.path(gtex_target_metadata_dir, "target_accessions_clinical.tsv")
```

Read in files

```{r read_files}

## Sample sheets ##

# SRA accessions tables - target_sra "Run" column is to be associated with compendium_samples "sample" column
# target SRA table
target_sra <- readr::read_csv(target_sra_path, col_types = readr::cols(.default = "c"))
# sample sheet of what was downloaded and processed
compendium_samples <- readr::read_tsv(compendium_samples_path, col_types = readr::cols( .default = "c"))

## Clinical metadata ##

# read in metadata paths
target_clinical_metadata <- target_clinical_metadata_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default = "c"))
  }) |> 
  purrr::list_rbind()

target_survival_metadata <- target_survival_metadata_paths |>
  purrr::map(\(path){
    readr::read_tsv(path, col_types=readr::cols(.default="c"))
  }) |>
  purrr::list_rbind()

```

The "sample" column in the survival and clinical metadata tables have entries that look like: TARGET-10-PAMDRM-09A

The "sample" column in the metadata table correspond to the "biospecimen_repository_sample_id" column in the target accessions table, which looks like: TARGET-10-PAMDRM-09A-01D

The sample IDs are harmonized in these metadata files by removing the last substring of "biospecimen_repository_sample_id" by "-".

Then, the TARGET accessions metadata is combined with the accessions sheet and the TARGET SRA metadata is filtered for accessions corresponding to samples that were downloaded and processed.

## Combine TARGET clinical and survival metadata

Combine each TARGET cancer type's clinical and survival metadata and filter for accessions that are present in the compendium

```{r combine_metadata}

# combine target clinical and survival metadata
target_metadata <- dplyr::full_join(
  target_clinical_metadata,
  target_survival_metadata,
  by = "sample"
)

# make a list of target accessions that were downloaded and processed for the compendium
target_downloaded <- compendium_samples |>
  dplyr::filter(group == "target") |>
  dplyr::pull(sample)

# filter target accessions for accessions that are in the compendium sample sheet
target_sra_metadata <- target_sra |>
  dplyr::filter(Run %in% target_downloaded) |>
  # derive target sample ID from sample ID 
  dplyr::mutate(
    # remove the hyphen followed by characters that are not hyphens that are at the end of the string
    target_sample_id = sub("-[^-]+$", "", biospecimen_repository_sample_id)
  )
```

There are some cases where more than one sample in the compendium that corresponds to one entry in the clinical metadata. In these cases, the sample comes from the same patient. These cases are handled by duplicating the clinical metadata for the samples that come from the same patient (don't assume one-to-one relationship in merge).

## Merge TARGET accessions with clinical metadata

```{r merge_accessions_with_metadata}
# merge target accessions with clinical metadata 
target_clinical_accessions <- dplyr::left_join(
  target_sra_metadata,
  target_metadata, 
  by = dplyr::join_by(
    target_sample_id == sample),
  relationship = "many-to-many"
    )
```

## Combine target metadata with SRA metadata for downloaded samples and select columns to include

```{r combine_metadata_with_sra}
# merge target accessions with clinical metadata 
target_clinical_accessions <- target_clinical_accessions |>
  # select for columns that are relevant to clinical metadata
  dplyr::select(
    target_sample_id, # from dbgap metadata, shared with clinical metadata
    Run, # from dbgap metadata
    age_at_diagnosis_days = age_at_diagnosis.diagnoses, 
    # selected over age_at_index.demographic due to having unrounded numbers
    age_at_earliest_diagnosis_in_years.diagnoses.xena_derived, 
    race.demographic,
    gender.demographic,
    OS,
    OS.time,
    disease_type,
    race.demographic,
    gender.demographic,
    ethnicity.demographic,
    project_id.project,
    name.project,
    classification_of_tumor.diagnoses,
    primary_diagnosis.diagnoses,
    sample_type.samples,
    tissue_type.samples,
    `_PATIENT`,
    # these columns are from the dbGaP SRA metadata
    BioProject,
    BioSample,
    biospecimen_repository,
    biospecimen_repository_sample_id,
    `Center Name`,
    Experiment,
    `dbGaP accession`,
    Instrument,
    `Sample Name`,
    `SRA Study`,
    study_name,
    submitted_subject_id,
    histological_type,
    body_site
  ) 
```

Check for samples without entries in any of the selected clinical metadata fields

```{r}
target_clinical_accessions |>
  dplyr::filter(
    if_all(c(
    age_at_diagnosis_days,
    age_at_earliest_diagnosis_in_years.diagnoses.xena_derived,
    disease_type,
    race.demographic,
    gender.demographic,
    ethnicity.demographic,
    project_id.project,
    name.project,
    classification_of_tumor.diagnoses,
    primary_diagnosis.diagnoses,
    sample_type.samples,
    tissue_type.samples,
    OS.time,
    OS,
    `_PATIENT`
  ), 
  is.na)) |>
  
  # summarize the studies samples with missing information come fro
  dplyr::summarise(
    .by = study_name,
    n = dplyr::n()
  )

```

All samples in the compendium have clinical data accounted for. Some ALL expansion phase 2 samples appear to be in the Xena browsers [ALL Phase 3](https://xenabrowser.net/datapages/?host=https%3A%2F%2Fgdc.xenahubs.net&dataset=TARGET-ALL-P3.clinical.tsv&allSamples=true&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443) metadata, instead of the phase 2 metadata as expected.

## Write output

```{r write_output}
readr::write_tsv(target_clinical_accessions, target_clinical_metadata_output, col_names = TRUE)
```
